### 2023-04-03 basics.Rmd
Basic file I use to generate all necessary tables

---
title: "UMAP_basics"
output: html_document
date: "2023-01-18"
---

Put here the location of your data, and where you want the created files to be stored, I recommend to make a new folder for this
```{r}
data_source <- "L:/DMS/AV_501_Pooja_Mandaviya/AV501v3_Pooja_Mandaviya_20220825.sav"     
output_location <- "P:/FSE_MACSBIO/Smit, Femke/clustering project/test"
#e.g. "C:/Users/Fariba/Documents"
```

```{r results="hide"}
set.seed(123)
library(haven)
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(ggplot2)
library(ggdensity)
library(patchwork)
library(Matrix, warn.conflicts = FALSE)
library(uwot)
library(survival)
library(igraph)
library(broom)
library(ggh4x)
library(mvtnorm)
library(stringr)
```

Reading the data:
```{r results = "hide"}
setwd(output_location)
data <- read_sav(data_source)
nrow(data)
# Subsetting only the columns that you need! Adapt this to the column names of your variables
dat <- data %>% select(
  RandomID,
  Age,
  SEX, 
  smoking_packyears,    ## Smoking should be a binary variable with 1 for yes (current smoker) and 0 for no (not current smoker, you can adapt this later if that is not how it is coded in your datafile)
  smoking_3cat,
  smoking_4cat,
  bmi,
  N_WHR,
  OSBP,
  ODBP,
  LDL,
  HDL,
  Triglyc, 
  CRP_2016F,   #creatinine
  Glucose_t0_FP,
  ALAT_serum,
  med_LP,
  med_HT,
  GFR_CKDEPI_cr)

# Coding for Females and Males in the right way
dat$SEX1 <- ifelse(dat$SEX == 2, "Female", "Male")  #In this script, male was coded as 1, female as 2
dat$SEX <- NULL
```

Looking into the smoking variables
```{r}
# Smoking should be a binary variable with 1 for yes (current smoker) and 0 for no (not current smoker)
# Here I create a new category, smoking_2cat, where this gets recoded (0 and 1 (never and former smoker) both get turned into 0, 2 (current smoker) gets turned into 1)
summary(dat[,c("smoking_packyears", "smoking_3cat", "smoking_4cat")])
dat$smoking_2cat <- replace(dat$smoking_3cat, dat$smoking_3cat==1, 0)
dat$smoking_2cat <- replace(dat$smoking_2cat, dat$smoking_2cat==2, 1)

dat$smoking_3cat <- NULL
dat$smoking_4cat <- NULL
dat$smoking_packyears <- NULL
```

```{r}
# Remove outliers and complete cases
remove_outliers <- function(x){
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  upperbound <- m + (5*s)
  lowerbound <- m - (5*s)
  ifelse((x > lowerbound) & (x < upperbound), x, NaN)
}

dat <- dat %>% mutate(across(-c(RandomID, SEX1, Age), remove_outliers)) %>% drop_na #drop_na to take only complete cases

#renaming variables
#replace the variables after the = with your variable names
dat <- rename(dat, eid = RandomID)
dat <- rename(dat, age = Age)
dat <- rename(dat, sex = SEX1)
dat <- rename(dat, whr = N_WHR)
dat <- rename(dat, hdl = HDL)
dat <- rename(dat, tg = Triglyc)
dat <- rename(dat, crp = CRP_2016F)
dat <- rename(dat, fg = Glucose_t0_FP)
dat <- rename(dat, alt = ALAT_serum)
dat <- rename(dat, ldl = LDL)
dat <- rename(dat, sbp = OSBP)
dat <- rename(dat, dbp = ODBP)
dat <- rename(dat, smoking = smoking_2cat)
dat <- rename(dat, scr = GFR_CKDEPI_cr)

#reorder columns
dat <- dat[,c("eid","age","sex","bmi","whr","sbp","dbp","alt","scr","crp","hdl","tg","ldl","fg","smoking")]
#dat_smallsmoking <- dat
#dat_smallsmoking[,"smoking"] != 0 
recoded_dat <- dat
```
```{r}
#write.table(strat_dat, file = "P:/FSE_MACSBIO/Smit, Femke/clustering project/data/strat_dat.tsv", sep = "\t", row.names = FALSE)
```

# Stratification by sex and transformation
Residualization by BMI, age and smoking. Then scaling residuals to mean 0 and SD 1.
```{r}
strat_dat <- recoded_dat %>%
  split(f = .$sex) %>%
  map(
    ~{
      .x %>%
        names %>%
        setdiff(c("eid", "sex", "age", "smoking", "bmi")) %>%
        map_dfc(
          function(feature){
            paste(feature, "~ age + smoking + bmi") %>%
              lm(data = .x) %>%
              resid %>%
              scale %>%
              data.frame %>%
              setNames(feature)
          }
        ) %>%
        mutate(eid = .x$eid, .before = 1)
    }
  )
map(strat_dat, head)
strat_dat %>%
  map(nrow)
```

# Non-linear projection using UMAP
Separately for each sex:
```{r}
umap_res <- strat_dat %>%
  map(
    ~{
      dat <- select(.x, -eid)
      n_total <- nrow(dat)
#      nn <- round(10 + 15 * (log10(n_total) - 3)) #was -4, made into -3 as it gave an error, said it had to be >=2
      nn <- 10    #as instructed by Daniel
      umap(
        # Data and number of dimensions to return
        dat, n_components = 2, 
        # Parameters to build the nearest-neighbor graph
        n_neighbors = nn, nn_method = "annoy", n_trees = 100, 
        n_sgd_threads = "auto", 
        # Parameters for the embedding
        init = "pca", n_epochs = 500, approx_pow = TRUE,
        binary_edge_weights = TRUE, dens_scale = 1, 
        # Additional elements to return
        ret_extra = c("model", "nn", "fgraph"),
        verbose = FALSE,
        ret_model = TRUE
      )
    }
  )
```

# Clustering
```{r}
umap_graphs <- umap_res %>%
  map(~igraph::graph_from_adjacency_matrix(.x$fgraph, mode = "undirected")) %>%
  map2(strat_dat, ~{ igraph::V(.x)$name <- .y$eid; .x })

umap_graphs %>%
  map(~.x)
```

# Finding communities within the graphs
Clustering using the Leiden algorithm, with initial seeds using the leading eigenvector:
```{r}
graph_clusters <- umap_graphs %>%
  map(~igraph::cluster_leading_eigen(.x)) %>%
  map2(umap_graphs, 
       ~igraph::cluster_leiden(.y, objective_function = "modularity", initial_membership = .x$membership, n_iterations = 500))
```

Extracting memberships:  
```{r}
memb_dat <- graph_clusters %>%  #said graph_lev first instead of graph_clusters
  map(igraph::membership) %>%
  map(~data.frame(eid = as.numeric(names(.x)), cluster = as.numeric(.x)))
map(memb_dat, head)
```

Extracting the corresponding subgraphs of each cluster:
```{r}
cluster_subgraphs <- memb_dat %>%
  map2(
    umap_graphs,
    ~{
      .x %>%
        split(f = .$cluster) %>%
        map(
          function(clus){
            igraph::induced_subgraph(.y, as.character(clus$eid))
          }
        )
    }
  )
```

To locate the 'centers' of these clusters, we can use the eigenvector centrality, which returns a value between 0 and 1. The higher the value, the higher the importance of an individual in the network:
```{r}
ev_dat <- cluster_subgraphs %>%
  map(
    ~{
      map_df(
        .x, 
        function(clus){ 
          evcent <- igraph::eigen_centrality(clus, directed = FALSE)$vector
          data.frame(
            eid = as.numeric(names(evcent)),
            value = as.numeric(evcent)
          )
        },
        .id = "cluster"
      ) %>%
        mutate(cluster = as.numeric(cluster))
    }
  )
map(ev_dat, head)
```

# Obtaining subdistributions from each cluster
For this we need to estimate mean/covariance matrix using eigen centrality as weights:

```{r}
cluster_pars <- ev_dat %>%
    map2(strat_dat, inner_join) %>%
    map(
        ~{
            .x %>%
                split(f = .$cluster) %>%
                map(
                    function(clusdat){
                        cov.wt(select(clusdat, whr, sbp, dbp, alt, scr, crp, hdl, tg, ldl, fg), wt = clusdat$value, cor = TRUE)
                    }
                )
        }
    )
```

# Adding 'central' subdistribution

This would correspond to individuals whose biomarker values are perfectly explained by BMI:
```{r}
center_subdis <- list(
    cov = diag(rep(1, ncol(strat_dat$Female) - 1)),
    center = rep(0, ncol(strat_dat$Female) - 1)
)
center_subdis

cluster_pars0 <- cluster_pars %>%
    map(~{.x$`0` <- center_subdis; .x})
map(cluster_pars0, names)
```

```{r}
gmm_res <- cluster_pars0 %>%
    map2(
        strat_dat,
        ~{
            d <- select(.y, -eid)
            mus <- map(.x, function(clus){ clus$center })
            covmats <- map(.x, function(clus){ clus$cov })
            pdfs <- map2(mus, covmats, function(mu, covmat){ mvtnorm::dmvnorm(d, mu, covmat) })
            k <- length(mus)
            ws <- rep(1 / k, k)
            L <- map2(pdfs, ws, function(pd, w){ pd * w })
            Lmat <- do.call(cbind, L)
            probs <- Lmat / rowSums(Lmat)
            Lclus <- colSums(probs)
            loglik <- sum(log(rowSums(Lmat)))
            Delta <- 1
            iter <- 1
            itermax <- 100
            while(Delta > 1e-10 && iter <= itermax){
                ws <- Lclus / sum(Lclus)
                L <- map2(pdfs, ws, function(pd, w){ pd * w })
                Lmat <- do.call(cbind, L)
                probs <- Lmat / rowSums(Lmat)
                Lclus <- colSums(probs)
                loglik_current <- sum(log(rowSums(Lmat)))
                Delta <- loglik_current - loglik
                loglik <- loglik_current
                iter <- iter + 1
            }
            print(paste("Convergence reached in", iter, "iterations"))
            list(probs = probs, weights = ws)
        }
    )

```

```{r}
cluster_params <- cluster_pars0 %>%
    map2(gmm_res, 
         ~map2(
             .x, .y$weights,
             function(clus, clusw){
                 clus$weight <- clusw
                 clus[c("center", "cov", "weight")]
                 }
             )
         )

varnames <- names(cluster_params$Female[[1]]$center)

map(cluster_params, map, names) %>%
    unlist %>%
    unique
```

# Calculating probabilities to belong to each subdistribution for every individual

Function to get probabilities to be assigned to each subdistribution:

```{r}
# Function that returns probabilities for a given mixture of Gaussian distributions
# X = Data
# center = List of centers of each Gaussian distribution
# covmats = List of covariance matrices for each Gaussian distribution
# weights = List o weights for each Gaussian distribution
getclusprob <- function(X, centers, covmats, weights){
    # Calculating probability density functions
    pdfs <- map2(centers, covmats, function(mu, covmat){ mvtnorm::dmvnorm(X, mu, covmat) })
    # Calculating likelihoods
    L <- map2(pdfs, weights, function(pd, w){ pd * w })
    # Joining in a matrix
    Lmat <- do.call(cbind, L)
    # Scaling by row to obtain probabilities
    probs <- Lmat / rowSums(Lmat)
    probs
}
```

```{r}

library(Hmisc)

getclusproball <- function(dataset, cluspars){
  d <- select(dataset, -eid)
  centers <- map(cluspars, pluck, "center")
  covmats <- map(cluspars, pluck, "cov")
  weights <- map(cluspars, pluck, "weight")
  getclusprob(d, centers, covmats, weights)
}

cluster_probs <- map2(strat_dat, cluster_params, getclusproball)

```

### 2023-04-04 get_cluster_descriptives.R
Update get_cluster_descriptives.R
Adapted weighted mean/median/quantiles calculation to use own calculations instead of Hmisc ones because those gave an error using other datasets. Should work for both continuous and binary data. Doesn't give prevalence per 1000.

@@ -1,20 +1,36 @@
-get_cluster_descriptives <- function(dataset, cluster_probs){
-  long_summary <- data.frame(cluster = integer(0), variable = character(0), n_allocated = integer(0), perc_allocated = numeric(0), wtd_mean = numeric(0), wtd_std = numeric(0), wtd_q0 = numeric(0), wtd_q25 = numeric(0), wtd_q50 = numeric(0), wtd_q75 = numeric(0), wtd_q100 = numeric(0))
-  for (col in 1:ncol(cluster_probs)){
-    clus <- colnames(cluster_probs)[col]
-    n_allocated <- sum(cluster_probs[,col] > 0.8)
+get_cluster_descriptives <- function(dataset, clus_probs){
+  dataset <- data.frame(dataset)
+  long_summary <- data.frame(cluster = integer(0), n_allocated = integer(0), perc_allocated = numeric(0), variable = character(0),  wtd_mean = numeric(0) , wtd_std = numeric(0), wtd_q0 = numeric(0), wtd_q25 = numeric(0), wtd_q50 = numeric(0), wtd_q75 = numeric(0), wtd_q100 = numeric(0))
+  for (col in 1:ncol(clus_probs)){
+    clus_name <- colnames(clus_probs)[col]
+    probs_col <- clus_probs[,col]
+    n_allocated <- sum(probs_col > 0.8)
     perc_allocated <- n_allocated/nrow(dataset)
-    for (var in 1:ncol(dataset)){
-      varia <- colnames(dataset)[var]
-      wtd_mean <- wtd.mean(dataset[,var], cluster_probs[,col])
-      wtd_std <- wtd.var(dataset[,var], cluster_probs[,col]) %>% sqrt()
-      wtd_iqr <- wtd.quantile(dataset[,var], cluster_probs[,col])
+    prob_sum <- sum(probs_col)
+    for (var in 2:ncol(dataset)){
+      var_name <- colnames(dataset)[var]
+      var_col <- dataset[,var]
+      wtd_mean <- sum(probs_col*var_col)/prob_sum
+      wtd_std <- sum(probs_col*(var_col - wtd_mean))/prob_sum %>% sqrt()
+      var_prob <- data.frame(var_col, probs_col)
+      var_prob <- var_prob[order(var_prob$var_col),]
+      var_prob$probs_cum <- cumsum(var_prob$probs_col)
+      q0 <- var_prob$var_col[1]
[…]

### 2023-04-05 get_descriptives_cohorts_nice.Rmd
Final cluster probabilities functions and how to use them

---
title: "functions and useage"
output: html_document
date: "2023-04-05"
---

```{r}
library(haven)
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
```

loading biomarker data 
```{r}
recoded_dat <- read_tsv("P:/FSE_MACSBIO/Smit, Femke/clustering project/data/recoded_dat.tsv", show_col_types = FALSE)
# [is.na(recoded_dis)] <- 0

strat_raw_dat <- recoded_dat %>% split( f = .$sex) %>% map(select, -c("sex"))
map(strat_raw_dat, head)

strat_dat <- recoded_dat %>%
  split(f = .$sex) %>%
  map(
    ~{
      .x %>%
        names %>%
        setdiff(c("eid", "sex", "age", "smoking", "bmi")) %>%
        map_dfc(
          function(feature){
            paste(feature, "~ age + smoking + bmi") %>%
              lm(data = .x) %>%
              resid %>%
              scale %>%
              data.frame %>%
              setNames(feature)
          }
        ) %>%
        mutate(eid = .x$eid, .before = 1)
    }
  )
map(strat_dat, head)
```

loading disease data
```{r}
recoded_dis <- read_tsv("P:/FSE_MACSBIO/Smit, Femke/clustering project/data/recoded_dis2.tsv", show_col_types = FALSE)  #any cancer is skin + other cancers
head(recoded_dis)

strat_dis <- merge(recoded_dat[,c("eid","sex")], recoded_dis, by = "eid") %>% split(f = .$sex) %>% map(select, -"sex")
map(strat_dis, head)
```

defining function cluster probabilities, cluster characteristics, cluster weighted disease prevalences
```{r}
getclusprob <- function(X, centers, covmats, weights){
    # Calculating probability density functions
    pdfs <- map2(centers, covmats, function(mu, covmat){ mvtnorm::dmvnorm(X, mu, covmat) })
    # Calculating likelihoods
    L <- map2(pdfs, weights, function(pd, w){ pd * w })
    # Joining in a matrix
    Lmat <- do.call(cbind, L)
    # Scaling by row to obtain probabilities
    probs <- Lmat / rowSums(Lmat)
    probs
}

getclusproball <- function(dataset, parameters){
  d <- select(dataset, -eid)
  cluspars <- parameters
  centers <- map(cluspars, pluck, "center")
  covmats <- map(cluspars, pluck, "cov")
  weights <- map(cluspars, pluck, "weight")
  getclusprob(d, centers, covmats, weights)
}

get_cluster_descriptives <- function(dataset, clus_probs){
  dataset <- data.frame(dataset)
  long_summary <- data.frame(cluster = integer(0), n_allocated = integer(0), perc_allocated = numeric(0), variable = character(0),  wtd_mean = numeric(0) , wtd_std = numeric(0), wtd_q0 = numeric(0), wtd_q25 = numeric(0), wtd_q50 = numeric(0), wtd_q75 = numeric(0), wtd_q100 = numeric(0))
  for (col in 1:ncol(clus_probs)){
    clus_name <- colnames(clus_probs)[col]
    probs_col <- clus_probs[,col]
    n_allocated <- sum(probs_col > 0.8)
    perc_allocated <- n_allocated/nrow(dataset)
    prob_sum <- sum(probs_col)
    for (var in 2:ncol(dataset)){
      var_name <- colnames(dataset)[var]
      var_col <- dataset[,var]
      wtd_mean <- sum(probs_col*var_col)/prob_sum
      wtd_std <- sum(probs_col*(var_col - wtd_mean))/prob_sum %>% sqrt()
      var_prob <- data.frame(var_col, probs_col)
      var_prob <- var_prob[order(var_prob$var_col),]
      var_prob$probs_cum <- cumsum(var_prob$probs_col)
      q0 <- var_prob$var_col[1]
      i25 <- which.min(abs(var_prob$probs_cum - 0.25*prob_sum))
      q25 <- var_prob$var_col[i25]
      i50 <- which.min(abs(var_prob$probs_cum - 0.5*prob_sum))
      q50 <- var_prob$var_col[i50]
      i75 <- which.min(abs(var_prob$probs_cum - 0.75*prob_sum))
      q75 <- var_prob$var_col[i75]
      q100 <- max(var_col)
      row <- nrow(long_summary) + 1
      long_summary[row,] <- c(clus_name, n_allocated, perc_allocated, var_name, wtd_mean, wtd_std, q0, q25, q50, q75, q100)
    }
  }
  long_summary
}

get_disease_prev <- function(dataset, clus_probs){
  #first column dataset needs to contain ids  
  dataset <- data.frame(dataset)
  long_summary <- data.frame(cluster = integer(0), n_allocated = integer(0), perc_allocated = numeric(0), variable = character(0),  wtd_prev = numeric(0))
  for (col in 1:ncol(clus_probs)){
    clus_name <- colnames(clus_probs)[col]
    probs_col <- clus_probs[,col]
    n_allocated <- sum(probs_col > 0.8)
    perc_allocated <- n_allocated/nrow(dataset)
    prob_sum <- sum(probs_col)
    for (var in 2:ncol(dataset)){
      var_name <- colnames(dataset)[var]
      var_col <- dataset[,var]
      wtd_prev <- sum(probs_col*var_col)/prob_sum*1000
      row <- nrow(long_summary) + 1
      long_summary[row,] <- c(clus_name, n_allocated, perc_allocated, var_name, wtd_prev)
    }
  }
  long_summary
}
```

getting cluster probabilities
```{r}
load("C:/Users/P70074150/Downloads/cluster_params_Maastricht.RData")
cluster_probs <- map2(strat_dat, cluster_params, getclusproball) 
map(cluster_probs, head)
```

getting cluster characteristics
```{r}
cluster_descriptives_resids <- map2(strat_dat, cluster_probs, get_cluster_descriptives)
map(cluster_descriptives_resids, head)

cluster_descriptives_raw <- map2(strat_raw_dat, cluster_probs, get_cluster_descriptives)
map(cluster_descriptives_raw, head)
```

getting disease prevalence
We can consider using this function also for smoking (and medicine use, etc, just all binary variables) as only the prevalence makes sense for these, there's no variance or uncertainty estimation relevant here.
```{r}
cluster_disease_prev <- map2(strat_dis, cluster_probs, get_disease_prev)
map(cluster_disease_prev, head)

```

### 2023-05-02 cross_sectional_FX.R
Update cross_sectional_FX.R

@@ -1,225 +1,875 @@
-# Functions
+## Functions to run clustering analysis to detect subgroups with different relationships between BMI and risk
 
-# ==============================================================================
-#
-# DESCRIPTIVE OVERVIEW OF NUMERIC AND FACTOR VARIABLES
-# WITH OVERALL TESTING OF DIFFERENCES BETWEEN BMI CATEGORIES
-#
-# Author: Carl Delfin (CRDF)
-# Version: 3
-# Date: 2022-10-28
-#
-# Instructions:
-#
-# (1) You must have the openxlsx package in order to save output as Excel sheets:
-#     install.packages("openxlsx", dependencies = TRUE)
-#
-# (2) You must have the data loaded in a data frame named 'df' before running
-#     the script, and the data must be processed according to the variable 
-#     format table provided.
-#
-# (3) You must set a path where the output will be saved, edit the 'path'
-#     variable below so that it matches your system.
-#
-# (4) If you want to have output by groups, you can add as many grouping
-#     variables as you want in the 'group_vars' vector below. You will get
-#     one output for each grouping variable.
-#
-# (5) OPTIONAL: If you want to filter out cases where the smallest group or
-#     count is less than 5, set 'filter_5' to TRUE below.
-#
-# (6) When the above is done, just run the script, e.g. by pressing the
-#     'Source' button in RStudio.
-#
-# ==============================================================================
-df <- recoded_dat
-df$sex <- as.factor(df$sex)
-df$smoking <- as.factor(df$smoking)
-# change as necessary:
-path <- ""
+# Declaring variables that are expected to be found in the data to run analysis
+IDCOL <- "eid"
+SEXVAR <- "sex"
+SEXGROUPS <- c("Female", "Male")
+BODYSIZEINDEX <- "bmi"
+COVARIATES <- c("age", "smoking")
+BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
+PERSEXVARS <- c(BODYSIZEINDEX, COVARIATES, BIOMARKERS)
+ALLVARS <- c(IDCOL, SEXVAR, PERSEXVARS)
 
-# change as necessary
-filter_5 <- FALSE 
 
-# WOMEN
-
-df <- filter(df, sex == "Female")
[…]

### 2023-05-02 cross_sectional_FX.R
Update cross_sectional_FX.R

@@ -785,19 +785,6 @@ wtd_roc <- function(label, preds, weights = rep(1, length(label))){
     data.frame(thresholds, TN, TP, FP, FN)
 }
 
-# ROC AUC 
-aucfx <- function(roctab){
-    roctab$FPR <- roctab$FP / (roctab$FP + roctab$TN)
-    roctab$TPR <- roctab$TP / (roctab$TP + roctab$FN)
-    right <- roctab[-nrow(roctab),]
-    left <- roctab[-1,]
-    width <- right$FPR - left$FPR
-    rect.area <- left$TPR * width
-    triangle.h <- right$TPR - left$TPR
-    triangle.area <- triangle.h * width / 2
-    sum(rect.area, triangle.area)
-}
-
 # Weighted ROC AUC variance parameters
 wtd_rocaucvar <- function(labels, preds, weights = rep(1, length(labels))) {
     cases <- preds[labels == 1]
@@ -822,7 +809,7 @@ wtd_rocaucvar <- function(labels, preds, weights = rep(1, length(labels))) {
 
 # Global survival metrics at 5 years
 global_survmetrics <- function(survmod, sexstrat, incidencedf){
-    purrr::map2_dfr(
+    purrr::map2(
         survmod, sexstrat,
         function(mod, X){
             STATUS_Y5 <- dplyr::transmute(
@@ -834,11 +821,52 @@ global_survmetrics <- function(survmod, sexstrat, incidencedf){
             STATUS_Y5 <- dplyr::inner_join(STATUS_Y5, X, by = "eid")
             STATUS_Y5$pred_survprob <- survival:::predict.coxph(object = mod, newdata = STATUS_Y5, type = "survival")
             roctab <- wtd_roc(label = STATUS_Y5$outcome_value, preds = STATUS_Y5$pred_survprob)
-            rocauc <- aucfx(roctab)
             rocaucvar <- wtd_rocaucvar(label = STATUS_Y5$outcome_value, preds = STATUS_Y5$pred_survprob)
-            return(list(roctab = roctab, rocauc = rocauc, rocaucvar = rocaucvar))
+            return(list(roctab = roctab, rocaucvar = rocaucvar))
         }
     )
+}
+
+# Global survival metrics at 5 years
+cluster_survmetrics <- function(survmod, sexstrat, incidencedf, clusres){
+    purrr::pmap(
+        list(survmod, sexstrat, clusres),
+        function(mod, X, clusdat){
+            STATUS_Y5 <- dplyr::transmute(
[…]

### 2023-05-03 GuidelineValidation.Rmd
Add files via upload

---
title: An R Markdown document converted from "GuidelineValidation.ipynb"
output: html_document
---

# Heterogeneity in the relationship between BMI and risk biomarkers

---

***Authors:***

- Daniel E. Coral

- Femke Smit

- Elena Santos

- Ali Farzaneh

---

## Introduction

This is a guideline to be followed by analysts in SOPHIA who are participating in the cross-sectional clustering project in the general population in Working Group 1. The idea is to standardize every step of the analysis across cohorts.

Not everyone has to follow all the steps, as we require different things from every cohort. We have divided the participating cohorts into 4 groups:

|Cohort group|Cohorts|
|-:|-:|
|*Discovery*|UK Biobank|
|*Validation*|Maastricht|
||Rotterdam|
||GHS|
|*Mental health*|Girona|
||Maastricht|
|*Intervention*|SCALE|
||ABOS|

This guideline is designed to be applied in discovery and validation cohorts. We will then apply what we learn in these cohorts to the mental health and intervention cohorts.

---

As a background, generally the relationship between BMI and multiple diseases is assumed to follow a continuum -- the higher the BMI, the higher the risk. However, it has also been found that in certain groups of people this relationship is disproportionally stronger or weaker for any given BMI. Our objective is to test the hypothesis that clustering-based approaches can be used to better capture these subgroups. An overview of the steps of our pipeline to test this hypothesis is shown in Figure 1.

|![image.png](attachment:f8b0dea3-1987-4811-8e4a-0877b1297a75.png)|
|-|
|*Figure 1. Pipeline overview*|

---

## Software

All steps are intended to be followed in the R environment. To facilitate the analyses we have put together a list of functions that can be used to run every step of this guideline. They are located in the accompanying file `cross_sectional_FX.R`, which you can load like this:

```{r}
source("cross_sectional_FX.R")
```

These functions have dependencies on the following packages:

```{r}
suppressMessages({
    library(tibble)
    library(readr)
    library(dplyr)
    library(tidyr)
    library(purrr)
    library(uwot)
    library(igraph)
    library(mvtnorm)
    library(survival)
})
```

The environment where discovery analysis in UK Biobank was executed is then the following:

```{r}
sessionInfo()
```

---

## Initial input

We have selected 10 traits, based on biological systems that are commonly affected by obesity:

- Blood pressure: SBP and DBP.
- Lipids: HDL, LDL, TG.
- Fat distribution: WHR.
- Glycemia: Fasting glucose.
- Liver metabolism: ALT.
- Kidney function: Creatinine.
- Inflammation: CRP.

The covariates that will be needed are:

- Sex.
- Age.
- Current smoking status, coded as 1 if current smoker, 0 otherwise.

The initial input table should look like the following:

```{r}
recoded_dat <- read_tsv("../data/recoded_dat.tsv", show_col_types = FALSE)
head(recoded_dat)
```

The units of the contiuous variables we expect to run the analysis:

- Age in years.
- Sex as a string of either "Female" or "Male".
- BMI in kg/m2.
- WHR is unitless, calculated by dividing waist and hip circumferences measured in cm.
- SBP and DBP in mmHg.
- ALT in U/L.
- sCr in umol/L.
- CRP in mg/L.
- HDL in mmol/L.
- TG in mmol/L.
- LDL in mmol/L.
- FG in mmol/L.
- Smoking as a dummy variable: 1 if currently smoking, 0 otherwise.

<font color=red size=3.5>***Note:***</font>

For the functions included in the `cross_sectional_FX.R` to work, the input table should be ***exactly*** as shown above.

---

## Missing data

Since our clustering method ignores individuals with missing values for any biomarker, the input data should only contain individuals who have all biomarker values. From the previous analyses, we know that:

- Only including complete cases without losing too much data is possible in UK Biobank, Maastricht, GHS and ABOS.
- In Rotterdam the initial input table contains some values have been imputed using a random forest algorithm.
- In Girona only a small subset of individuals have CRP values.
- In SCALE there are values for waist but not for hip circumference, so it is not possible to calculate WHR.

Based on these observations we have made the following decisions on how to deal with missing values:

|How to handle missing values|Cohorts|
|-:|-:|
|Only include complete cases.|UK Biobank|
||Maastricht|
||GHS|
||ABOS|
|||
|Use data that has been imputed.|Rotterdam|
|||
|Retain all individuals.|Girona|
||SCALE|

***For the cohorts in the last group***, to be able to apply our method, we will assume that BMI explains the variability in the biomarkers that are missing. This assumption is based on what we have observed in the other cohorts. In practice, this means that the clustering method will focus on the biomarkers that are available to group individuals into clusters. The input table should still have the same columns so that the functions in `cross_sectional_FX.R` work properly.

---

## Remove possible errors in measurement

In discovery and validation cohorts we will exclude biomarker measurements that are 5 SD away from the mean, under the assumption that these are most likely measurement errors. This can be done using the `remove_outliers` function that we have provided, which replaces outliers with `NA` values. Then we again make sure to have only complete cases:

```{r}
recoded_dat <- mutate(recoded_dat, 
                      across(c(bmi, whr, sbp, dbp, alt, scr, crp, hdl, tg, ldl, fg), 
                             ~remove_outliers(.x, sdunits = 5)))
recoded_dat <- recoded_dat[complete.cases(recoded_dat),]
```

---

## Stratify by sex

All the pipeline is applied separately in each sex group. The functions we have designed work on a list containing two dataframes for each sex group, which we can obtain like this:

```{r}
strat_dat <- split(recoded_dat, ~sex)
```

To see the first lines of the two elements in the list:

```{r}
lapply(strat_dat, head)
```

---

## Summary of initial input

We need a table summarising the initial input, which can be generated like this:

```{r}
gendesc_tab <- get_general_descriptives(strat_dat)
gendesc_tab
```

For smoking:

- `Summary1` contains the categories.
- `Summary2` contains the proportion of each category.

For the rest (continuous) variables:

- `Summary1` contains the mean and standard deviation.
- `Summary2` contains the median and interquartile range.

---

## Estimates of BMI-biomarker associations

The first step of the pipeline is to generate sex-specific linear models of BMI for each variable, adjusting for age, and smoking. To do that we have the following function:

```{r}
mods <- get_bmimods(strat_dat)
```

The result is a table with a column that contains the models specific for each sex and biomarker:

```{r}
print(mods)
```

As an example, we can print the summary of the female model for CRP:

```{r}
summary(mods$mod[mods$sex == "Female" & mods$Biomarker == "crp"][[1]])
```

We then use this table to generate a table containing the estimates of the effect of BMI on every biomarker:

```{r}
bmicoefs_tab <- get_bmicoefs(mods)
bmicoefs_tab
```

---

## Generate table with residual values

The next step in the pipeline is to calculate for each individual how much they deviate from the expected given the BMI. To do that we have the following function:

```{r}
residtab <- get_residtabs(strat_dat, mods)
```

```{r}
lapply(residtab, head)
```

---

## Obtaining clusters and probabilities of allocation

We use this residuals to run UMAP, which we use not only to obtain a projection of residual data in 2 dimensions, but also to run a probabilistic network clustering algorithm. We have wrapped all the clustering steps in a single function that we apply to the `residtab` object:

```{r}
cluster_results <- get_cluster_results(residtab)
```

The result is a list containing the UMAP model, the parameters of the clusters found, the modularity score, which is a measure of the quality of the network partition, and the allocation probabilities of individuals to all clusters:

```{r}
lapply(cluster_results, names)
```

---

## Table summarising clusters

To understand what characterizes the clusters found, we will need a table with descriptives of the distribution of residuals per cluster, which can be generated like this:

```{r}
clustersummary <- get_clustersummary(cluster_results)
```

```{r}
head(clustersummary)
```

We also need the distribution of the biomarkers in the clusters in their natural scale. To calculate this for a specific cluster, we use all individuals, and weigh each individual by their corresponding cluster allocation probability. The function for this is the following:

```{r}
cluster_descriptives <- get_cluster_descriptives(cluster_results, strat_dat)
```

```{r}
head(cluster_descriptives)
```

---

## Relationship between BMI and biomarkers in each cluster

Given that the relationship between BMI and biomarkers is expected to be different in these clusters compared to the overall relationship found above, we will quantify how much it changed by comparing models weighted by probabilities of each cluster. The function is the following:

```{r}
clusbmicoefs <- get_bmicoefs_clusters(cluster_results, strat_dat)
```

```{r}
head(clusbmicoefs)
```

---

## Disease prevalence

The next step is to check whether the prevalence of certain diseases differ in the clusters found, which would give us an initial idea of the clinical relevance of these subgroups. To do that we expect a table where each column represents a disease, coded as 1 or 0 depending on the presence or absence of the disease for each individual ***included in the clustering analysis***:

```{r}
disease_dat <- inner_join(read_tsv("../data/selected_dx.tsv", show_col_types = FALSE),
                          read_tsv("../data/anycadat.tsv", show_col_types = FALSE),
                          by = "eid") %>%
    ## Only people who were included in the clustering analysis
    filter(eid %in% recoded_dat$eid)
head(disease_dat)
```

Diseases in this table are commonly associated with obesity; the list includes:

```{r}
colnames(disease_dat)[-1]
```

***Note:*** We don't expect to have all diagnoses available in all validation cohorts, but the more the better!

The first thing we need is a table summarising the sex-specific prevalence of each disease:

```{r}
overall_diseasesum <- get_diseasesummary(disease_dat, strat_dat)
```

```{r}
head(overall_diseasesum)
```

Then, we will need a table summarising the prevalence within each cluster, which is again calculated using the cluster probabilites as weights:

```{r}
cluster_diseasesum <- get_diseasesummary_clusters(disease_dat, cluster_results)
```

```{r}
head(cluster_diseasesum)
```

Some of the prevalences may be affected by subtle residual differences in the covariates we adjusted for before clustering (i.e., age, smoking and most importantly, BMI), and also certain medications. To control for that:

- We will treat 'cluster 0', the cluster characterized by all biomarkers being aligned with BMI, as the reference cluster.
- For each disease, we will calculate the increase in disease OR that corresponds to a unit increase in cluster allocation probability of every cluster relative to 'cluster 0', while all other covariates remain equal.

To do that we need an additional table containing medication data, which should look like this:

```{r}
meds_dat <- read_tsv("../data/selected_meds.tsv", show_col_types = FALSE)
head(meds_dat)
```

We defined people receiving medications using the following ATC codes:

|Medication group|ATC codes|
|:-:|:-:|
|Insulin|A10A|
|Other antidiabetic|A10B|
|Antihypertensives|C01|
||C02|
||C03|
||C07|
||C08|
||C09|
|Lipid lowering|C10|

The function to run an adjusted analysis is the following:

```{r}
cluster_adjdiseasesum <- get_adjdiseasesummary_clusters(cluster_results, disease_dat, meds_dat, strat_dat)
```

```{r}
head(cluster_adjdiseasesum)
```

These latter estimates represent the change in the logarithm of the OR for a disease per each unit change in the logarithm of cluster probability. The rationale for using this transformation is to back-transform the estimates so that they represent the change in the OR for a certain percent increase/decrease in probability.

As a complementary analysis, we would calculate these OR by allocating individuals to a certain cluster based on different thresholds. The function for that is the following:

```{r}
cluster_adjdiseasesum_thresh <- get_adjdiseasesummary_clustersthresh(cluster_results, disease_dat, meds_dat, strat_dat, thresh = c(.6, .7, .8))
```

```{r}
head(cluster_adjdiseasesum_thresh)
```

These estimates represent the change in the logarithm of the OR for change in cluster membership relative to cluster 0 - the concordant cluster.

---

## MACE incidence - UK Biobank, Rotterdam (GHS?)

Next, we will calculate the risk of future disease given the probability of cluster allocation. The table we expect to able to run this part looks like the following:

```{r}
survmacedat <- read_tsv("../data/survmacedat.tsv", show_col_types = FALSE)
head(survmacedat)
```

The first thing we need, as done before, is a sex-specific summary of the data available:

```{r}
macesum <- get_incidencesummary(survmacedat, strat_dat)
```

```{r}
macesum
```

And again, we would run similar cluster-specific analyses as done for prevalence:

- Crude incidence per cluster:

```{r}
cluster_macesum <- get_incidencesummary_clusters(survmacedat, cluster_results)
```

```{r}
cluster_macesum
```

- Unadjusted and adjusted survival models with cluster probabilities as predictors:

```{r}
cluster_adjmacesum <- get_adjincidencesummary_clusters(cluster_results, survmacedat, meds_dat, strat_dat)
```

```{r}
head(cluster_adjmacesum)
```

- Unadjusted and adjusted survival models with hard cluster allocation at different thresholds as predictor:

```{r}
cluster_adjmacesum_thresh <- get_adjincidencesummary_clustersthresh(cluster_results, survmacedat, meds_dat, strat_dat, thresh = c(.6, .7, .8))
```

```{r}
head(cluster_adjmacesum_thresh)
```

We want to know whether a simple survival model using all biomarkers is able to capture the risk we observe in the clusters. To do that we first construct this simple survival model:

```{r}
globalmod_mace <- global_survivalmodel(strat_dat, survmacedat)
```

The result is a sex specific survival model:

```{r}
lapply(globalmod_mace, class)
```

Each model includes all variables, e.g.:

```{r}
summary(globalmod_mace$Female)$coefficients
```

We then extract the probability assigned by these models to each individual to be free of MACE events after 5 years of follow-up, and then we assess both the overall and the cluster-specific predictive performance. For the first we have the following function:

```{r}
global_survmetrics_mace <- global_survmetrics(globalmod_mace, strat_dat, survmacedat)
```

This function returns for each sex a list containing 2 elements:
- A table containing multiple thresholds of survival probabilities, and their corresponding number of true/false positives and true/false negatives, which we can use to calculate ROC curves.
- A calculation of the variance parameters around the ROC AUC estimate, which we will use to formally compare 2 ROC AUCs using Delong's method.

For example, these is how this object looks like for females:

```{r}
str(global_survmetrics_mace$Female)
```

We will also calculate these same estimates for each cluster. We will do this by weighting individuals by their cluster probabilities:

```{r}
cluster_survmetrics_mace <- cluster_survmetrics(globalmod_mace, strat_dat, survmacedat, cluster_results)
```

The result is similar as before, for each cluster identified:

```{r}
str(cluster_survmetrics_mace$Female$cluster_1)
```

---

## Diabetes risk - UK Biobank, Rotterdam (GHS?)

Given the strong link between BMI and T2D, we will also test whether one of the clusters identified is associated with higher or lower risk of developing diabetes over time. The data we need looks just like the one we used above for MACE:

```{r}
survdmdat <- read_tsv("../data/survdmdat.tsv", show_col_types = FALSE)
head(survdmdat)
```

We will then apply the same functions as we did for MACE:

- Sex-specific summary of the data available:

```{r}
timetodmsum <- get_incidencesummary(survdmdat, strat_dat)
```

```{r}
timetodmsum
```

- Crude incidence per cluster:

```{r}
cluster_timetodmsum <- get_incidencesummary_clusters(survdmdat, cluster_results)
```

```{r}
head(cluster_timetodmsum)
```

- Unadjusted and adjusted survival models with cluster probabilities as predictors (here ignoring medication):

```{r}
cluster_adjtimetodmsum <- get_adjincidencesummary_clusters(cluster_results, survdmdat, meds_dat[,"eid"], strat_dat)
```

```{r}
head(cluster_adjtimetodmsum)
```

- Unadjusted and adjusted survival models with hard cluster allocation at different thresholds as predictor:

```{r}
cluster_adjtimetodmsum_thresh <- get_adjincidencesummary_clustersthresh(cluster_results, survdmdat, meds_dat[,"eid"], strat_dat, thresh = c(.6, .7, .8))
```

```{r}
head(cluster_adjtimetodmsum_thresh)
```

- Overall survival model:

```{r}
globalmod_timetodm <- global_survivalmodel(strat_dat, survdmdat)
```

```{r}
lapply(globalmod_timetodm, class)
```

- Overall predictive performance at 5 years:

```{r}
global_survmetrics_timetodm <- global_survmetrics(globalmod_timetodm, strat_dat, survdmdat)
```

- Cluster-specific predictive performance at 5 years, weighted by probabilities:

```{r}
cluster_survmetrics_timetodm <- cluster_survmetrics(globalmod_timetodm, strat_dat, survdmdat, cluster_results)
```

---

## Gathering and saving results

We'll save all results in a single file:

```{r}
result_file <- list(
    General_descriptives = gendesc_tab,
    BMI_coefficients = bmicoefs_tab,
    Cluster_results = cluster_results,
    Cluster_summary = clustersummary,
    Cluster_descriptives = cluster_descriptives,
    Cluster_BMI_coefficients = clusbmicoefs,
    Disease_summary = overall_diseasesum,
    Cluster_Disease_summary = cluster_diseasesum,
    Cluster_Disease_summary_Adjusted = cluster_adjdiseasesum,
    Cluster_Disease_summary_Threshold = cluster_adjdiseasesum_thresh,
    MACE_summary = macesum,
    Cluster_MACE_summary = cluster_macesum,
    Cluster_MACE_summary_Adjusted = cluster_adjmacesum,
    Cluster_MACE_summary_Threshold = cluster_adjmacesum_thresh,
    Global_SurvMetrics_MACE = global_survmetrics_mace,
    Cluster_SurvMetrics_MACE = cluster_survmetrics_mace,
    TimetoDM_summary = timetodmsum,
    Cluster_TimetoDM_summary = cluster_timetodmsum,
    Cluster_TimetoDM_Adjusted = cluster_adjtimetodmsum,
    Cluster_TimetoDM_Threshold = cluster_adjtimetodmsum_thresh,
    Global_SurvMetrics_TimetoDM = global_survmetrics_timetodm,
    Cluster_SurvMetrics_TimetoDM = cluster_survmetrics_timetodm
)
```

```{r}
save(result_file, file = "result_file.RData")
```

---

## Uploading results

The results shown do not contain any individual level data. We require that you upload this 'result_file.RData' file to the Teams folder in SOPHIA, which would be located here:

> CrossWP > Analyst working groups > WG1 > UMAP_project > *cohort_name* > data

```{r}
rmarkdown::convert_ipynb(, output = xfun::with_ext(input, "Rmd"))
```

### 2023-05-03 cross_sectional_FX.R
Update cross_sectional_FX.R

@@ -194,7 +194,7 @@ umapclus_fx <- function(RESIDMAT){
     eids <- RESIDMAT$eid
     Xmat <- RESIDMAT[,names(RESIDMAT) != "eid"]
     n_total <- nrow(Xmat)
-    nn <- round(10 + 15 * (log10(n_total) - 4))
+    nn <- max(10, round(10 + 15 * (log10(n_total) - 4)))
     message("\t2. Running UMAP...")
     umapres <- uwot::umap(
         Xmat, n_components = 2, 

### 2023-05-03 cross_sectional_FX.R
Update cross_sectional_FX.R

@@ -579,12 +579,12 @@ adjclusdiseasesumfx_thresh <- function(clusdat, diseasedf, medsdf, X, thresh){
     probs <- clusdat$probs
     joindf <- tidyr::pivot_longer(probs, -eid, names_to = "cluster", values_to = "cluster_prob")
     basecovardf <- X[,c("eid", BODYSIZEINDEX, COVARIATES)]
-    joindf <- dplyr::inner_join(joindf, basecovardf, by = "eid", relationship = "many-to-many")
+    joindf <- dplyr::inner_join(joindf, basecovardf, by = "eid")
     meds <- names(medsdf)
     meds <- meds[meds != "eid"]
-    joindf <- dplyr::inner_join(joindf, medsdf, by = "eid", relationship = "many-to-many")
+    joindf <- dplyr::inner_join(joindf, medsdf, by = "eid")
     dxdf <- tidyr::pivot_longer(diseasedf, -eid, names_to = "disease", values_to = "status", values_drop_na = TRUE)
-    joindf <- dplyr::inner_join(joindf, dxdf, by = "eid", relationship = "many-to-many")
+    joindf <- dplyr::inner_join(joindf, dxdf, by = "eid")
     joindf <- tidyr::drop_na(joindf)
     joindf$cluster <- relevel(factor(joindf$cluster), ref = "cluster_0")
     purrr::map_dfr(
@@ -717,11 +717,11 @@ adjclusincidencesumfx_thresh <- function(clusdat, incidencedf, medsdf, X, thresh
     probs <- clusdat$probs
     joindf <- tidyr::pivot_longer(probs, -eid, names_to = "cluster", values_to = "cluster_prob")
     basecovardf <- X[,c("eid", BODYSIZEINDEX, COVARIATES)]
-    joindf <- dplyr::inner_join(joindf, basecovardf, by = "eid", relationship = "many-to-many")
+    joindf <- dplyr::inner_join(joindf, basecovardf, by = "eid")
     meds <- names(medsdf)
     meds <- meds[meds != "eid"]
-    joindf <- dplyr::inner_join(joindf, medsdf, by = "eid", relationship = "many-to-many")
-    joindf <- dplyr::inner_join(joindf, incidencedf, by = "eid", relationship = "many-to-many")
+    joindf <- dplyr::inner_join(joindf, medsdf, by = "eid")
+    joindf <- dplyr::inner_join(joindf, incidencedf, by = "eid")
[…]

### 2023-05-03 get_cluster_descriptives.R
Delete get_cluster_descriptives.R

@@ -1,36 +1,0 @@
-get_cluster_descriptives <- function(dataset, clus_probs){
-  dataset <- data.frame(dataset)
-  long_summary <- data.frame(cluster = integer(0), n_allocated = integer(0), perc_allocated = numeric(0), variable = character(0),  wtd_mean = numeric(0) , wtd_std = numeric(0), wtd_q0 = numeric(0), wtd_q25 = numeric(0), wtd_q50 = numeric(0), wtd_q75 = numeric(0), wtd_q100 = numeric(0))
-  for (col in 1:ncol(clus_probs)){
-    clus_name <- colnames(clus_probs)[col]
-    probs_col <- clus_probs[,col]
-    n_allocated <- sum(probs_col > 0.8)
-    perc_allocated <- n_allocated/nrow(dataset)
-    prob_sum <- sum(probs_col)
-    for (var in 2:ncol(dataset)){
-      var_name <- colnames(dataset)[var]
-      var_col <- dataset[,var]
-      wtd_mean <- sum(probs_col*var_col)/prob_sum
-      wtd_std <- sum(probs_col*(var_col - wtd_mean))/prob_sum %>% sqrt()
-      var_prob <- data.frame(var_col, probs_col)
-      var_prob <- var_prob[order(var_prob$var_col),]
-      var_prob$probs_cum <- cumsum(var_prob$probs_col)
-      q0 <- var_prob$var_col[1]
-      i25 <- which.min(abs(var_prob$probs_cum - 0.25*prob_sum))
-      q25 <- var_prob$var_col[i25]
-      i50 <- which.min(abs(var_prob$probs_cum - 0.5*prob_sum))
-      q50 <- var_prob$var_col[i50]
-      i75 <- which.min(abs(var_prob$probs_cum - 0.75*prob_sum))
-      q75 <- var_prob$var_col[i75]
-      q100 <- max(var_col)
-      row <- nrow(long_summary) + 1
-      long_summary[row,] <- c(clus_name, n_allocated, perc_allocated, var_name, wtd_mean, wtd_std, q0, q25, q50, q75, q100)
-    }
-  }
-  long_summary
-}
-
-# use the function like this: cluster_descriptives <- map2(strat_dat, cluster_probs, get_cluster_descriptives)
-# works for both continuous data and binary data
-# doesn't give prevalence per 1000, should be easily adaptable
-# doesn't require Hmisc anymore

### 2023-05-10 cross_sectional_FX.R
Delete cross_sectional_FX.R

@@ -1,903 +1,0 @@
-## Functions to run clustering analysis to detect subgroups with different relationships between BMI and risk
-
-# Declaring variables that are expected to be found in the data to run analysis
-IDCOL <- "eid"
-SEXVAR <- "sex"
-SEXGROUPS <- c("Female", "Male")
-BODYSIZEINDEX <- "bmi"
-COVARIATES <- c("age", "smoking")
-BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
-PERSEXVARS <- c(BODYSIZEINDEX, COVARIATES, BIOMARKERS)
-ALLVARS <- c(IDCOL, SEXVAR, PERSEXVARS)
-
-
-# Outlier removal
-remove_outliers <- function(x, sdunits = 5){
-    m <- mean(x, na.rm = TRUE)
-    s <- sd(x, na.rm = TRUE)
-    upperbound <- m + (sdunits*s)
-    lowerbound <- m - (sdunits*s)
-    ifelse((x > lowerbound) & (x < upperbound), x, NaN)
-}
-
-# Checking data input
-checkinput <- function(X){
-    stopifnot(is.data.frame(X))
-    stopifnot(ALLVARS %in% names(X))
-    stopifnot(names(X) %in% ALLVARS)
-}
-
-# Variable types
-vartype <- function(x){
-    if(is.numeric(x)){
-        if(length(unique(x)) > 10){
-            return("Numeric")
-        } else { 
-            return("Categorical")
-        }
-    } else {
-        if(is.character(x) | is.factor(x)){
-            return("Categorical")
-        } else {
-            return(NA)
-        }
-    }
-}
-
-# Summary of categorical variables
-categorical_sumstats <- function(x){
-    N <- length(x)
-    N_miss <- sum(is.na(x)) 
-    props <- table(x)
-    props <- sort(props, decreasing = TRUE)
-    props <- data.frame(N = N, N_miss = N_miss,
-                        Summary1 = names(props), 
-                        Summary2 = as.vector(props))
-    props$Summary2 <- paste0(props$Summary2, " (", round(100 * props$Summary2 / sum(props$Summary2), 2), "%)")
-    props
-}
-
-# Summary of continuous variables
-continuous_sumstats <- function(x){
-    N <- length(x)
-    N_miss <- sum(is.na(x))
-    Mn <- round(mean(x, na.rm = TRUE), 2)
[…]

### 2023-05-10 cross_sectional_FX.R
Add files via upload

@@ -1,0 +1,908 @@
+## Functions to run clustering analysis to detect subgroups with different relationships between BMI and risk
+
+# Declaring variables that are expected to be found in the data to run analysis
+IDCOL <- "eid"
+SEXVAR <- "sex"
+SEXGROUPS <- c("Female", "Male")
+BODYSIZEINDEX <- "bmi"
+COVARIATES <- c("age", "smoking")
+BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
+PERSEXVARS <- c(BODYSIZEINDEX, COVARIATES, BIOMARKERS)
+ALLVARS <- c(IDCOL, SEXVAR, PERSEXVARS)
+
+
+# Outlier removal
+remove_outliers <- function(x, sdunits = 5){
+    m <- mean(x, na.rm = TRUE)
+    s <- sd(x, na.rm = TRUE)
+    upperbound <- m + (sdunits*s)
+    lowerbound <- m - (sdunits*s)
+    ifelse((x > lowerbound) & (x < upperbound), x, NaN)
+}
+
+# Checking data input
+checkinput <- function(X){
+    stopifnot(is.data.frame(X))
+    stopifnot(ALLVARS %in% names(X))
+    stopifnot(names(X) %in% ALLVARS)
+}
+
+# Variable types
+vartype <- function(x){
+    if(is.numeric(x)){
+        if(length(unique(x)) > 10){
+            return("Numeric")
+        } else { 
+            return("Categorical")
+        }
+    } else {
+        if(is.character(x) | is.factor(x)){
+            return("Categorical")
+        } else {
+            return(NA)
+        }
+    }
+}
+
+# Summary of categorical variables
+categorical_sumstats <- function(x){
+    N <- length(x)
+    N_miss <- sum(is.na(x)) 
+    props <- table(x)
+    props <- sort(props, decreasing = TRUE)
+    props <- data.frame(N = N, N_miss = N_miss,
+                        Summary1 = names(props), 
+                        Summary2 = as.vector(props))
+    props$Summary2 <- paste0(props$Summary2, " (", round(100 * props$Summary2 / sum(props$Summary2), 2), "%)")
+    props
+}
+
+# Summary of continuous variables
+continuous_sumstats <- function(x){
+    N <- length(x)
+    N_miss <- sum(is.na(x))
+    Mn <- round(mean(x, na.rm = TRUE), 2)
[…]

### 2023-05-10 GuidelineValidation.Rmd
Add files via upload

@@
 - `Summary1` contains the mean and standard deviation.
-- `Summary2` contains the median and interquartile range.
+- `Summary2` contains the median and percentiles 2.5, 25, 75 and 97.5.
 
@@
 
-We then use this table to generate a table containing the estimates of the effect of BMI on every biomarker:
+We then use this table to generate a table containing the estimates of the effect of BMI as well as the covariates on every biomarker:
 
@@
 ```{r}
-save(result_file, file = "result_file.RData")
+save(result_file, file = "../data/ukb/result_file.RData")
 ```
@@
 > CrossWP > Analyst working groups > WG1 > UMAP_project > *cohort_name* > data
-
-```{r}
-rmarkdown::convert_ipynb(, output = xfun::with_ext(input, "Rmd"))
-```

### 2023-07-29 cross_sectional_FX.R
Delete cross_sectional_FX.R

@@ -1,908 +1,0 @@
-## Functions to run clustering analysis to detect subgroups with different relationships between BMI and risk
-
-# Declaring variables that are expected to be found in the data to run analysis
-IDCOL <- "eid"
-SEXVAR <- "sex"
-SEXGROUPS <- c("Female", "Male")
-BODYSIZEINDEX <- "bmi"
-COVARIATES <- c("age", "smoking")
-BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
-PERSEXVARS <- c(BODYSIZEINDEX, COVARIATES, BIOMARKERS)
-ALLVARS <- c(IDCOL, SEXVAR, PERSEXVARS)
-
-
-# Outlier removal
-remove_outliers <- function(x, sdunits = 5){
-    m <- mean(x, na.rm = TRUE)
-    s <- sd(x, na.rm = TRUE)
-    upperbound <- m + (sdunits*s)
-    lowerbound <- m - (sdunits*s)
-    ifelse((x > lowerbound) & (x < upperbound), x, NaN)
-}
-
-# Checking data input
-checkinput <- function(X){
-    stopifnot(is.data.frame(X))
-    stopifnot(ALLVARS %in% names(X))
-    stopifnot(names(X) %in% ALLVARS)
-}
-
-# Variable types
-vartype <- function(x){
-    if(is.numeric(x)){
-        if(length(unique(x)) > 10){
-            return("Numeric")
-        } else { 
-            return("Categorical")
-        }
-    } else {
-        if(is.character(x) | is.factor(x)){
-            return("Categorical")
-        } else {
-            return(NA)
-        }
-    }
-}
-
-# Summary of categorical variables
-categorical_sumstats <- function(x){
-    N <- length(x)
-    N_miss <- sum(is.na(x)) 
-    props <- table(x)
-    props <- sort(props, decreasing = TRUE)
-    props <- data.frame(N = N, N_miss = N_miss,
-                        Summary1 = names(props), 
-                        Summary2 = as.vector(props))
-    props$Summary2 <- paste0(props$Summary2, " (", round(100 * props$Summary2 / sum(props$Summary2), 2), "%)")
-    props
-}
-
-# Summary of continuous variables
-continuous_sumstats <- function(x){
-    N <- length(x)
-    N_miss <- sum(is.na(x))
-    Mn <- round(mean(x, na.rm = TRUE), 2)
[…]

### 2023-07-29 cross_sectional_FX.R
Replacing files with new versions

@@ -1,0 +1,456 @@
+## Functions to run clustering analysis to detect subgroups with different relationships between BMI and risk
+
+# Declaring variables that are expected to be found in the data to run analysis
+IDCOL <- "eid"
+SEXVAR <- "sex"
+SEXGROUPS <- c("Female", "Male")
+BODYSIZEINDEX <- "bmi"
+COVARIATES <- c("age", "smoking")
+BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
+PERSEXVARS <- c(BODYSIZEINDEX, COVARIATES, BIOMARKERS)
+ALLVARS <- c(IDCOL, SEXVAR, PERSEXVARS)
+
+
+# Outlier removal
+remove_outliers <- function(x, sdunits = 5){
+    m <- mean(x, na.rm = TRUE)
+    s <- sd(x, na.rm = TRUE)
+    upperbound <- m + (sdunits*s)
+    lowerbound <- m - (sdunits*s)
+    ifelse((x > lowerbound) & (x < upperbound), x, NaN)
+}
+
+# Checking data input
+checkinput <- function(X){
+    stopifnot(is.data.frame(X))
+    stopifnot(ALLVARS %in% names(X))
+    stopifnot(names(X) %in% ALLVARS)
+}
+
+# Variable types
+vartype <- function(x){
+    if(is.numeric(x)){
+        if(length(unique(x)) > 10){
+            return("Numeric")
+        } else { 
+            return("Categorical")
+        }
+    } else {
+        if(is.character(x) | is.factor(x)){
+            return("Categorical")
+        } else {
+            return(NA)
+        }
+    }
+}
+
+# Summary of categorical variables
+categorical_sumstats <- function(x){
+    N <- length(x)
+    N_miss <- sum(is.na(x)) 
+    props <- table(x)
+    props <- sort(props, decreasing = TRUE)
+    props <- data.frame(N = N, N_miss = N_miss,
+                        Summary1 = names(props), 
+                        Summary2 = as.vector(props))
+    props$Summary2 <- paste0(props$Summary2, " (", round(100 * props$Summary2 / sum(props$Summary2), 2), "%)")
+    props
+}
+
+# Summary of continuous variables
+continuous_sumstats <- function(x){
+    N <- length(x)
+    N_miss <- sum(is.na(x))
+    Mn <- round(mean(x, na.rm = TRUE), 2)
[…]

### 2023-07-29 cross_sectional_FX2.R

@@ -1,0 +1,1104 @@
+## Functions to assess predictive ability of clusters
+
+## Columns needed
+BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
+BODYSIZEINDEX <- "bmi"
+COVARIATES <- c("age", "smoking")
+DISEASES <- c("HT", "CHD", "Stroke", "PAD", "CKD", "LiverFailure", "RA", "T2D", "T1D")
+MEDICATION <- c("Insulin", "AntiDM", "AntiHT", "LipidLower")
+
+# Calculating cluster probabilities
+clusterprobcalc <- function(X){
+    RES <- dplyr::mutate(
+        X,
+        # Calculate standardized residuals
+        resids = purrr::map2(
+            data, residmod,
+            function(DATA, RESIDMOD){
+                sapply(
+                    BIOMARKERS,
+                    function(biomarker){
+                        PREDICTDAT <- as.matrix(cbind(Intercept = 1, DATA[,c(BODYSIZEINDEX, COVARIATES)]))
+                        MODELDAT <- dplyr::filter(RESIDMOD, Biomarker == biomarker)
+                        RESIDSD <- MODELDAT$SDRes
+                        COEFDAT <- MODELDAT[,c("Intercept", BODYSIZEINDEX, COVARIATES)]
+                        COEFDAT <- t(COEFDAT)
+                        PREDVAL <- as.vector(PREDICTDAT %*% COEFDAT)
+                        RESIDVAL <- DATA[[biomarker]] - PREDVAL
+                        RESIDVAL / RESIDSD
+                    }
+                )
+            }
+        ),
+        # Calculate probabilities to valid clusters
+        probs = purrr::map2(
+            resids, clusmod,
+            function(RESIDDF, CLUSMOD){
+                mus <- CLUSMOD$validclus_centers
+                mus <- purrr::map(mus, function(mu){ mu[BIOMARKERS] })
+                covmats <- CLUSMOD$validclus_covmats
+                covmats <- purrr::map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
+                wts <- CLUSMOD$validclus_weights
+                pdfs <- mapply(function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) }, mus, covmats)
+                L <- pdfs %*% diag(wts)
[…]

### 2023-07-29 01_Clustering.Rmd
Add files via upload

--- 2023-05-10 GuidelineValidation.Rmd
@@
 ---
-title: An R Markdown document converted from "GuidelineValidation.ipynb"
+title: An R Markdown document converted from "01_Clustering.ipynb"
 output: html_document
@@
     library(mvtnorm)
-    library(survival)
 })
@@
 
-## Disease prevalence
-
-The next step is to check whether the prevalence of certain diseases differ in the clusters found, which would give us an initial idea of the clinical relevance of these subgroups.
-To do that we expect a table where each column represents a disease, coded as 1 or 0 depending on the presence or absence of the disease for each individual ***included in the clustering analysis***:
-
-```{r}
-disease_dat <- inner_join(read_tsv("../data/selected_dx.tsv", show_col_types = FALSE),
-                          read_tsv("../data/anycadat.tsv", show_col_types = FALSE),
-                          by = "eid") %>%
-    ## Only people who were included in the clustering analysis
-    filter(eid %in% recoded_dat$eid)
-head(disease_dat)
-```
-
-Diseases in this table are commonly associated with obesity; the list includes:
-
-```{r}
-colnames(disease_dat)[-1]
-```
-
-***Note:*** We don't expect to have all diagnoses available in all validation cohorts, but the more the better!
-
-The first thing we need is a table summarising the sex-specific prevalence of each disease:
-
-```{r}
-overall_diseasesum <- get_diseasesummary(disease_dat, strat_dat)
-```
-
-```{r}
-head(overall_diseasesum)
-```
-
-Then, we will need a table summarising the prevalence within each cluster, which is again calculated using the cluster probabilites as weights:
-
-```{r}
-cluster_diseasesum <- get_diseasesummary_clusters(disease_dat, cluster_results)
-```
-
-```{r}
-head(cluster_diseasesum)
-```
-
-Some of the prevalences may be affected by subtle residual differences in the covariates we adjusted for before clustering (i.e., age, smoking and most importantly, BMI), and also certain medications.
-To control for that:
-
-- We will treat 'cluster 0', the cluster characterized by all biomarkers being aligned with BMI, as the reference cluster.
-- For each disease, we will calculate the increase in disease OR that corresponds to a unit increase in cluster allocation probability of every cluster relative to 'cluster 0', while all other covariates remain equal.
-
-To do that we need an additional table containing medication data, which should look like this:
-
-```{r}
-meds_dat <- read_tsv("../data/selected_meds.tsv", show_col_types = FALSE)
-head(meds_dat)
-```
-
-We defined people receiving medications using the following ATC codes:
-
-|Medication group|ATC codes|
-|:-:|:-:|
-|Insulin|A10A|
-|Other antidiabetic|A10B|
-|Antihypertensives|C01|
-||C02|
-||C03|
-||C07|
-||C08|
-||C09|
-|Lipid lowering|C10|
-
-The function to run an adjusted analysis is the following:
-
-```{r}
-cluster_adjdiseasesum <- get_adjdiseasesummary_clusters(cluster_results, disease_dat, meds_dat, strat_dat)
-```
-
-```{r}
-head(cluster_adjdiseasesum)
-```
-
-These latter estimates represent the change in the logarithm of the OR for a disease per each unit change in the logarithm of cluster probability.
-The rationale for using this transformation is to back-transform the estimates so that they represent the change in the OR for a certain percent increase/decrease in probability.
-
-As a complementary analysis, we would calculate these OR by allocating individuals to a certain cluster based on different thresholds. The function for that is the following:
-
-```{r}
-cluster_adjdiseasesum_thresh <- get_adjdiseasesummary_clustersthresh(cluster_results, disease_dat, meds_dat, strat_dat, thresh = c(.6, .7, .8))
-```
-
-```{r}
-head(cluster_adjdiseasesum_thresh)
-```
-
-These estimates represent the change in the logarithm of the OR for change in cluster membership relative to cluster 0 - the concordant cluster.
-
----
-
-## MACE incidence - UK Biobank, Rotterdam (GHS?)
-
-Next, we will calculate the risk of future disease given the probability of cluster allocation. The table we expect to able to run this part looks like the following:
-
-```{r}
-survmacedat <- read_tsv("../data/survmacedat.tsv", show_col_types = FALSE)
-head(survmacedat)
-```
-
-The first thing we need, as done before, is a sex-specific summary of the data available:
-
-```{r}
-macesum <- get_incidencesummary(survmacedat, strat_dat)
-```
-
-```{r}
-macesum
-```
-
-And again, we would run similar cluster-specific analyses as done for prevalence:
-
-- Crude incidence per cluster:
-
-```{r}
-cluster_macesum <- get_incidencesummary_clusters(survmacedat, cluster_results)
-```
-
-```{r}
-cluster_macesum
-```
-
-- Unadjusted and adjusted survival models with cluster probabilities as predictors:
-
-```{r}
-cluster_adjmacesum <- get_adjincidencesummary_clusters(cluster_results, survmacedat, meds_dat, strat_dat)
-```
-
-```{r}
-head(cluster_adjmacesum)
-```
-
-- Unadjusted and adjusted survival models with hard cluster allocation at different thresholds as predictor:
-
-```{r}
-cluster_adjmacesum_thresh <- get_adjincidencesummary_clustersthresh(cluster_results, survmacedat, meds_dat, strat_dat, thresh = c(.6, .7, .8))
-```
-
-```{r}
-head(cluster_adjmacesum_thresh)
-```
-
-We want to know whether a simple survival model using all biomarkers is able to capture the risk we observe in the clusters. To do that we first construct this simple survival model:
-
-```{r}
-globalmod_mace <- global_survivalmodel(strat_dat, survmacedat)
-```
-
-The result is a sex specific survival model:
-
-```{r}
-lapply(globalmod_mace, class)
-```
-
-Each model includes all variables, e.g.:
-
-```{r}
-summary(globalmod_mace$Female)$coefficients
-```
-
-We then extract the probability assigned by these models to each individual to be free of MACE events after 5 years of follow-up, and then we assess both the overall and the cluster-specific predictive performance.
-For the first we have the following function:
-
-```{r}
-global_survmetrics_mace <- global_survmetrics(globalmod_mace, strat_dat, survmacedat)
-```
-
-This function returns for each sex a list containing 2 elements:
-- A table containing multiple thresholds of survival probabilities, and their corresponding number of true/false positives and true/false negatives, which we can use to calculate ROC curves.
-- A calculation of the variance parameters around the ROC AUC estimate, which we will use to formally compare 2 ROC AUCs using Delong's method.
-
-For example, these is how this object looks like for females:
-
-```{r}
-str(global_survmetrics_mace$Female)
-```
-
-We will also calculate these same estimates for each cluster. We will do this by weighting individuals by their cluster probabilities:
-
-```{r}
-cluster_survmetrics_mace <- cluster_survmetrics(globalmod_mace, strat_dat, survmacedat, cluster_results)
-```
-
-The result is similar as before, for each cluster identified:
-
-```{r}
-str(cluster_survmetrics_mace$Female$cluster_1)
-```
-
----
-
-## Diabetes risk - UK Biobank, Rotterdam (GHS?)
-
-Given the strong link between BMI and T2D, we will also test whether one of the clusters identified is associated with higher or lower risk of developing diabetes over time.
-The data we need looks just like the one we used above for MACE:
-
-```{r}
-survdmdat <- read_tsv("../data/survdmdat.tsv", show_col_types = FALSE)
-head(survdmdat)
-```
-
-We will then apply the same functions as we did for MACE:
-
-- Sex-specific summary of the data available:
-
-```{r}
-timetodmsum <- get_incidencesummary(survdmdat, strat_dat)
-```
-
-```{r}
-timetodmsum
-```
-
-- Crude incidence per cluster:
-
-```{r}
-cluster_timetodmsum <- get_incidencesummary_clusters(survdmdat, cluster_results)
-```
-
-```{r}
-head(cluster_timetodmsum)
-```
-
-- Unadjusted and adjusted survival models with cluster probabilities as predictors (here ignoring medication):
-
-```{r}
-cluster_adjtimetodmsum <- get_adjincidencesummary_clusters(cluster_results, survdmdat, meds_dat[,"eid"], strat_dat)
-```
-
-```{r}
-head(cluster_adjtimetodmsum)
-```
-
-- Unadjusted and adjusted survival models with hard cluster allocation at different thresholds as predictor:
-
-```{r}
-cluster_adjtimetodmsum_thresh <- get_adjincidencesummary_clustersthresh(cluster_results, survdmdat, meds_dat[,"eid"], strat_dat, thresh = c(.6, .7, .8))
-```
-
-```{r}
-head(cluster_adjtimetodmsum_thresh)
-```
-
-- Overall survival model:
-
-```{r}
-globalmod_timetodm <- global_survivalmodel(strat_dat, survdmdat)
-```
-
-```{r}
-lapply(globalmod_timetodm, class)
-```
-
-- Overall predictive performance at 5 years:
-
-```{r}
-global_survmetrics_timetodm <- global_survmetrics(globalmod_timetodm, strat_dat, survdmdat)
-```
-
-- Cluster-specific predictive performance at 5 years, weighted by probabilities:
-
-```{r}
-cluster_survmetrics_timetodm <- cluster_survmetrics(globalmod_timetodm, strat_dat, survdmdat, cluster_results)
-```
-
----
-
 ## Gathering and saving results
@@
     Cluster_descriptives = cluster_descriptives,
-    Cluster_BMI_coefficients = clusbmicoefs,
-    Disease_summary = overall_diseasesum,
-    Cluster_Disease_summary = cluster_diseasesum,
-    Cluster_Disease_summary_Adjusted = cluster_adjdiseasesum,
-    Cluster_Disease_summary_Threshold = cluster_adjdiseasesum_thresh,
-    MACE_summary = macesum,
-    Cluster_MACE_summary = cluster_macesum,
-    Cluster_MACE_summary_Adjusted = cluster_adjmacesum,
-    Cluster_MACE_summary_Threshold = cluster_adjmacesum_thresh,
-    Global_SurvMetrics_MACE = global_survmetrics_mace,
-    Cluster_SurvMetrics_MACE = cluster_survmetrics_mace,
-    TimetoDM_summary = timetodmsum,
-    Cluster_TimetoDM_summary = cluster_timetodmsum,
-    Cluster_TimetoDM_Adjusted = cluster_adjtimetodmsum,
-    Cluster_TimetoDM_Threshold = cluster_adjtimetodmsum_thresh,
-    Global_SurvMetrics_TimetoDM = global_survmetrics_timetodm,
-    Cluster_SurvMetrics_TimetoDM = cluster_survmetrics_timetodm
+    Cluster_BMI_coefficients = clusbmicoefs
 )
@@
 
-The results shown do not contain any individual level data. We require that you upload this 'result_file.RData' file to the Teams folder in SOPHIA, which would be located here:
+We require that you upload the 'result_file.RData' file, which does not contain any individual level data, to the Teams folder in SOPHIA, which would be located here:
 
 > CrossWP > Analyst working groups > WG1 > UMAP_project > *cohort_name* > data
+
+---
+
+## Saving discovery cohort used
+
+```{r}
+save(strat_dat, file = "../data/ukb/strat_dat.RData")
+```

### 2023-07-29 03_ValClusOutcomes.Rmd
Add files via upload

---
title: An R Markdown document converted from "03_ValClusOutcomes.ipynb"
output: html_document
---

# Cluster probabilities and outcomes

---

***Authors:***

- Daniel E. Coral

- Femke Smit

- Elena Santos

- Ali Farzaneh

---

In this second part of the analysis, we will examine how the clusters we have validated across cohorts are associated with prevalent diseases at the time of clustering, and also assess whether they add significant information for prediction of MACE events and diabetes progression on top  of commonly used risk stratification tools.

## Libraries and functions

The libraries needed to run this analysis:

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(survival)
```

And the functions we have prepared to facilitate some steps:

```{r}
source("cross_sectional_FX2.R")
```

---

## Loading data needed

### Initial input table of biomarkers and basic covariates

The input table is the same table of 10 traits we had prior to run UMAP. Here is a description of this table:

|System targeted |Biomarker               |Units                        |Column name|
|----------------|------------------------|-----------------------------|-----------|
|                |                        |                             |           |
|Individual ID   |-                       |-                            |eid        |
|                |                        |                             |           |
|Blood pressure  |Systolic blood pressure |millimeters of mercury (mmHg)|sbp        |
|                |Diastolic blood pressure|millimeters of mercury (mmHg)|dbp        |
|                |                        |                             |           |
|Lipid fractions |High density lipoprotein|mmol/L                       |hdl        |
|                |Low density lipoprotein |mmol/L                       |ldl        |
|                |Tryglicerides           |mmol/L                       |tg         |
|                |                        |                             |           |
|Glycemia        |Fasting glucose         |mmol/L                       |fg         |
|                |                        |                             |           |
|Liver metabolism|Alanine transaminase    |U/L                          |alt        |
|                |                        |                             |           |
|Fat distribution|Waist-to-hip ratio      |cm/cm                        |whr        |
|                |                        |                             |           |
|Kidney function |Serum creatinine        |umol/L                       |scr        |
|                |                        |                             |           |
|Inflammation    |C reactive protein      |mg/L                         |crp        |
|                |                        |                             |           |
|Basic covariates|Current smoking status  |1 if yes, 0 if not           |smoking    |
|                |Sex                     |String ("Female" or "Male")  |sex        |
|                |Age                     |Years                        |age        |

***Important note:*** All columns should be there in the units required, and the names should match, so that the functions we have prepared for the analyses work properly. This is true for this and all the following tables we require for our analysis.

This input table has been preprocessed by:

1. Filtering out values that are possible errors in measurement (>5 SD away from the mean in continuous variables).
2. Only including complete cases.
3. Stratifying by sex.

Here is how the input table should look like - a list of two data frames, one for each sex:

```{r}
load("../data/ukb/strat_dat.RData")
```

```{r}
map(strat_dat, head)
```

### Table of validated clusters

The second thing needed is the clusters we have validated. We have put this in an R file called `validclusmod`:

```{r}
load("../data/validclusmod.RData")
print(validclusmod)
```

This object contais, for each sex:
- `residmod`: The model to obtain residuals for each variable, i.e., the variability beyond what is explained by BMI, adjusting for age and smoking.
- `clusmod`: The clustering model to apply to the residuals.

### Table of pre-existing conditions and medications

The third thing we need is a table of pre-existing conditions and medications participants are currently taking:

```{r}
covar_dat <- read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

All the columns in this table are coded 1 or 0 representing current diagnosis of a disease or whether the person is taking the medications specified. The exception is `T2Dage`, which is the age of onset of T2D. This is what each column represent:

|Group       |Column name |Meaning
|------------|------------|--------
|Diagnoses   |HT          |Hypertension
|            |CHD         |Coronary heart disease
|            |Stroke      |Stroke
|            |PAD         |Peripheral artery disease
|            |CKD         |Chronic kidney disease
|            |LiverFailure|Liver failure
|            |RA          |Rheumatoid arthritis
|            |T2D         |Type 2 diabetes
|            |T1D         |Type 1 diabetes
|Age at onset|T2Dage      |Age at onset of T2D - It is 0 if `T2D` is 0. Needed in SCORE2.
|Medication  |Insulin     |Taking insulin
|            |AntiDM      |Taking medication for diabetes other than insulin
|            |AntiHT      |Taking medication for hypertension
|            |LipidLower  |Taking lipid-lowering medication

If any of the columns in this table are missing in your data, one option is to assume that none in your population had the disease.

### Survival data

Lastly, we need survival data for MACE and diabetes progression. They should look like this:

```{r}
survmacedat <- read_tsv("../data/survmacedat.tsv", show_col_types = FALSE)
head(survmacedat)
```

```{r}
survdmdat <- read_tsv("../data/survdmdat.tsv", show_col_types = FALSE)
head(survdmdat)
```

These two tables include individuals followed ***up to 10 years***. This means that any outcome after 10 years is censored. `outcome_value` is 1 if the person experienced the event during the follow-up time and 0 if not. `outcome_timeyrs` is the time of follow-up in years, up to the first event or up to 10 years.

It is important that these tables ***do not include*** individuals who already experience the events we will study. In any case, we will make sure of this in the next step, when we combine all the data. For example, any individual in the `survmacedat` table with a value of 1 in the columns `CHD`, `Stroke` or `PAD` of the `covar_tab` table, will be excluded from the analysis.

In case your cohort does not have survival data, then follow this guideline until the section below entitled "Prevalent diseases and medication".

---

## Calculation of cluster probabilities

With the data needed in place, we can start by calculating cluster allocation probabilities given the biomarker data. For that we will first add a new column called `data` to the `validclusmod` table where we will put the biomarker data for each sex:

```{r}
alldat <- mutate(
    validclusmod,
    data = map(sex, ~strat_dat[[.x]])
)
print(alldat)
```

Once we have this table, we can run the function to calculate cluster probabilities:

```{r}
alldat <- clusterprobcalc(alldat)
print(alldat)
```

Checking that the probabilities were calculated for each sex:

```{r}
head(alldat$data[[1]])
```

---

## Centered log-ratio transformation

Since using cluster allocation probabilities in regression is problematic because all probabilities for each individual add up to 1, we will use the centered log-ratio transformation (CLR), a technique commonly used when dealing with compositional data.

The clr transformation of the probability of an individual $i$ to belong to a cluster $c$ is the following:

$$
\text{CLR}(cluster_{c})_{i} = \log\left(\frac{P(cluster_{c})_{i}}{\exp\left(\frac{1}{K}\sum_{k=1}^{K}\log(P(cluster_{k})_{i})\right)}\right)
$$

The denominator in this formula is the geometric mean of the cluster allocation probabilities of individual $i$.

The reason why we chose this transformation is not only because it enables us to use cluster allocation in regression analysis, but also because it helps in the interpretability of the coefficients derived from the regression. In each regression we leave out the CLR of the reference cluster, which in our case is the cluster where biomarkers change concordantly with BMI. We have label this cluster "BC". Removing this CLR and only using the CLRs of the remaining clusters means that regression coefficients can be interpreted as the expected change in the dependent variable by increasing $P(cluster_{c})$ specifically at the expense of reducing $P(cluster_{BC})$ <cite id="abe09"><a href="#zotero|10831815/UYLBIXH3">(1)</a></cite>, or in other words, the effect of deviating from the concordant cluster towards a specific discordant phenotype.

To calculate the transformation:

```{r}
alldat <- clrcalc(alldat)
print(alldat)
```

Here is how the data looks like now:

```{r}
head(alldat$data[[1]])
```

---

## Descriptive statistics

At this point we will recheck some of the characteristics of the clusters as we did in our previous script, weighting calculations by cluster probabilities.

The distribution of biomarkers per cluster:

```{r}
markerdistribdf <- markerdistribfx(alldat)
```

```{r}
head(markerdistribdf)
```

The effect of BMI on biomarkers specifically within each cluster, adjusted for age and smoking:

```{r}
bmieffmarkerdf <- bmieffmarkerfx(alldat)
```

```{r}
head(bmieffmarkerdf)
```

---

## Prevalent diseases and medication

To add covariate data to the `alldat` table we will do the following:

```{r}
alldat <- mutate(
    alldat,
    data = map(data, inner_join, covar_dat, by = "eid")
)
print(alldat)
```

Checking again if the columns were added as expected:

```{r}
head(alldat$data[[1]])
```

We will first count the number of individuals with disease in each cluster. Here we will also count the number of individuals taking each class of medications in each cluster.

```{r}
countcovarsdf <- countcovarsfx(alldat)
```

```{r}
head(countcovarsdf)
```

We will use this table to calculate prevalences and compare prevalences across clusters.

We are also interesting in looking at the following proportions in each cluster:

- Proportion of individuals with hypertension receiving antihypertensives.
- Proportion of individuals with CHD receiving lipid-lowering medication.
- Proportion of individuals with T2D taking insulin.
- Proportion of individuals with T2D taking insulin or any other anti-diabetic medication.

```{r}
countspectxdf <- countspectxfx(alldat)
```

```{r}
head(countspectxdf)
```

We will also formally test the association between cluster allocation and diseases using logistic regressions where the outcome is each disease and the predictors are the cluster allocations. We will have two models for each disease, one with only clusters, and a second one adjusting for medication.

```{r}
assocdxdf <- assocdxfx(alldat)
```

```{r}
head(assocdxdf)
```

---

## Adding survival data

As explained before, since we want to be careful when adding survival data for analysis, we have prepared a function separately for both outcomes, and making sure we exclude individuals who already experience the events under study:

```{r}
alldat <- alldat %>%
    mutate(
        macedf = purrr::map(data, addsurvmacedat, SURVDATA = survmacedat),
        dmdf = purrr::map(data, addsurvdmdat, SURVDATA = survdmdat)
    )
print(alldat)
```

`macedf` and `dmdf` now contain the data necessary to run survival analysis.

---

## Overall and cluster-specific Kaplan-Meier estimates

The first thing to do is to obtain overall and cluster-specific cumulative incidence rates using the Kaplan-Meier method:

```{r}
kmestdf <- kmestfx(alldat)
```

```{r}
head(kmestdf)
```

---

## Cox models

### MACE

To quantify the association of clusters to MACE, as well as its potential contribution for prediction, we will compare two models. The reference model will include all predictors that are part of SCORE2, the risk stratification tool for CVD recommended by the European Society of Cardiology <cite id="pzdxs"><a href="#zotero|10831815/ZY2CL5NC">(2)</a></cite>. We will use a version of this score that has been validated in diabetic populations, and includes some additional clinically useful predictors <cite id="itp38"><a href="#zotero|10831815/FPUFQKFI">(3)</a></cite>. Additionally, for the sake of completeness, some important pre-existing conditions and pharmacological treatments, such as hypertension/antihypertensives, as well as any predictor that we had in our initial input table that are not part of SCORE2, will also be included. We will compare this reference model to one that includes also the cluster probabilities and then we'll compare the ability of these two models to predict MACE.

#### Fitting models

```{r}
coxmodmacedf <- coxmodelsmace(alldat)
```

```{r}
print(coxmodmacedf)
```

Here `score2` contains the predictors used in SCORE2-Diabetes, plus other predictors that we have in our input table and the CLR-transformed cluster allocations. The two models are contained in the last two columns. `mod_null` contains the null model, which we will use to calculate our metrics.

#### Coefficient estimates

```{r}
macesurvcoefs <- macesurvcoefx(coxmodmacedf)
```

```{r}
head(macesurvcoefs)
```

#### Comparison of predictive ability

We can now assess the predictive ability of each model and compare them. As the models are nested, we will use the gold-standard method, the likelihood ratio test. Given the wide use of the c-statistic, we will also use this metric. However, comparing two c-statistics is not as powerful as the likelihood ratio test.

```{r}
compmodmacedf <- comparemodsmace(coxmodmacedf)
compmodmacedf
```

Some details of these columns:

- `LRTp` is the p-value of the likelihood ratio test comparing models with or without cluster allocations.
- `AdeqInd` is the adequacy index comparing likelihood ratios of the two models. 1 minus this value represent the fraction of added information by cluster allocation.
- `cdiffp` is the p-value of the difference between c-statistics of the two models.

While p-values of both the likelihood ratio tests and the difference between C-statistics show evidence of added value, the adequacy index quantifies how much additional information is obtained when cluster allocations are added on top of common biomarkers contained the SCORE2.

#### Adequacy index by cluster

Next we quantify how much is the added value of the new model to each cluster. We do this by recalculating the log likelihood of each model but this time weighting individuals by their cluster probabilities.

```{r}
adeqindbyclusmace <- AdeqIndClusMACEFx(coxmodmacedf)
```

```{r}
adeqindbyclusmace
```

#### Adequacy index by SCORE2 probabilities

We will also assess the adequacy index in individuals over certain thresholds of 10-year probability of MACE calculated by SCORE2. This is to have an idea of the utility of adding cluster allocation information across the scale of SCALE2.

```{r}
adeqindbypremace <- AdeqIndByPreMACEFx(coxmodmacedf)
```

```{r}
adeqindbypremace
```

#### Decision curve analysis

The last step in assessing clinical utility of clustering allocations is to perform a decision curve analysis. First we will assess the overall net benefit of both models:

```{r}
dcamace <- DCurvMACEFx(coxmodmacedf)
head(dcamace)
```

We will also calculate this by cluster:

```{r}
dcaclusmace <- DCurvMACEbyClFx(coxmodmacedf)
```

```{r}
head(dcaclusmace)
```

### Diabetes

For diabetes progression, the reference model will include fasting glucose, in addition to all the components of the metabolic syndrome that we already have included in our input table. And as before, the second model will include cluster probabilities.

#### Fitting models

```{r}
coxmoddmdf <- coxmodelsdm(alldat)
```

```{r}
print(coxmoddmdf)
```

#### Coefficient estimates

```{r}
dmsurvcoefs <- dmsurvcoefx(coxmoddmdf)
```

```{r}
head(dmsurvcoefs)
```

#### Comparison of predictive ability

```{r}
compmoddmdf <- comparemodsdm(coxmoddmdf)
compmoddmdf
```

#### Adequacy index by cluster

```{r}
adeqindbyclusdm <- AdeqIndClusDMFx(coxmoddmdf)
```

```{r}
adeqindbyclusdm
```

#### Adequacy index by probabilities of base model

```{r}
adeqindbypredm <- AdeqIndByPreDMFx(coxmoddmdf)
```

```{r}
adeqindbypredm
```

#### Decision curve analysis

The last step in assessing clinical utility of clustering allocations is to perform a decision curve analysis. First we will assess the overall net benefit of both models:

```{r}
dcadm <- DCurvDMFx(coxmoddmdf)
head(dcadm)
```

We will also calculate this by cluster:

```{r}
dcaclusdm <- DCurvDMbyClFx(coxmoddmdf)
```

```{r}
head(dcaclusdm)
```

---

## Saving data

As done before, we will ask you to save an R file that does not contain any individual data, only summary statistics, as follows:

```{r}
result_file2 <- list(
    MarkerDistrib = markerdistribdf,
    BMIeffOnMarker = bmieffmarkerdf,
    CountCovars = countcovarsdf,
    CountSpecDXMeds = countspectxdf,
    CrossSectAssoc = assocdxdf,
    KaplanMeierDF = kmestdf,
    MACESurvCoefs = macesurvcoefs,
    ComparisonMACE = compmodmacedf,
    AdeqIndByClusMACE = adeqindbyclusmace,
    AdeqIndByPreMACE = adeqindbypremace,
    DCAResMACE = dcamace,
    DCAREsClusMACE = dcaclusmace,
    DMSurvCoefs = dmsurvcoefs,
    ComparisonDM = compmoddmdf,
    AdeqIndByClusDM = adeqindbyclusdm,
    AdeqIndByPreDM = adeqindbypredm,
    DCAResDM = dcadm,
    DCAREsClusDM = dcaclusdm
)
```

```{r}
save(
    result_file2,
    file = "../data/ukb/result_file2.RData"
)
```

This file should be uploaded to the respective folder of the cohort in Teams:

> CrossWP > Analyst working groups > WG1 > UMAP_project > *cohort_name* > data

---

## References

<!-- BIBLIOGRAPHY START -->
<div class="csl-bib-body">
  <div class="csl-entry"><i id="zotero|10831815/UYLBIXH3"></i>
    <div class="csl-left-margin">1. </div><div class="csl-right-inline">Coenders G, Pawlowsky-Glahn V. On interpretations of tests and effect sizes in regression models with a compositional predictor. SORT-Statistics and Operations Research Transactions [Internet]. 2020 Jun 26 [cited 2023 Jul 28];44(1):201–20. Available from: https://raco.cat/index.php/SORT/article/view/371189</div>
  </div>
  <div class="csl-entry"><i id="zotero|10831815/ZY2CL5NC"></i>
    <div class="csl-left-margin">2. </div><div class="csl-right-inline">SCORE2 working group and ESC Cardiovascular risk collaboration. SCORE2 risk prediction algorithms: new models to estimate 10-year risk of cardiovascular disease in Europe. European Heart Journal [Internet]. 2021 Jul 1 [cited 2023 Jul 20];42(25):2439–54. Available from: https://doi.org/10.1093/eurheartj/ehab309</div>
  </div>
  <div class="csl-entry"><i id="zotero|10831815/FPUFQKFI"></i>
    <div class="csl-left-margin">3. </div><div class="csl-right-inline">SCORE2-Diabetes Working Group and the ESC Cardiovascular Risk Collaboration. SCORE2-Diabetes: 10-year cardiovascular risk estimation in type 2 diabetes in Europe. European Heart Journal [Internet]. 2023 May 29 [cited 2023 Jul 13];ehad260. Available from: https://doi.org/10.1093/eurheartj/ehad260</div>
  </div>
</div>
<!-- BIBLIOGRAPHY END -->

### 2023-07-31 03_ValClusOutcomes.Rmd
Add files via upload

@@
 
-If any of the columns in this table are missing in your data, one option is to assume that none in your population had the disease.
+If any of the columns in this table are missing in your data, one option is to assume that none in your population had the disease, i.e., you should have a column with all 0.
 
@@
 <div class="csl-bib-body">
-  <div class="csl-entry"><i id="zotero|10831815/UYLBIXH3"></i>
-    <div class="csl-left-margin">1.
-</div><div class="csl-right-inline">Coenders G, Pawlowsky-Glahn V.
-On interpretations of tests and effect sizes in regression models with a compositional predictor.
-SORT-Statistics and Operations Research Transactions [Internet].
-2020 Jun 26 [cited 2023 Jul 28];44(1):201–20.
-Available from: https://raco.cat/index.php/SORT/article/view/371189</div>
-  </div>
-  <div class="csl-entry"><i id="zotero|10831815/ZY2CL5NC"></i>
-    <div class="csl-left-margin">2.
-</div><div class="csl-right-inline">SCORE2 working group and ESC Cardiovascular risk collaboration.
-SCORE2 risk prediction algorithms: new models to estimate 10-year risk of cardiovascular disease in Europe.
-European Heart Journal [Internet].
-2021 Jul 1 [cited 2023 Jul 20];42(25):2439–54.
-Available from: https://doi.org/10.1093/eurheartj/ehab309</div>
-  </div>
-  <div class="csl-entry"><i id="zotero|10831815/FPUFQKFI"></i>
-    <div class="csl-left-margin">3.
-</div><div class="csl-right-inline">SCORE2-Diabetes Working Group and the ESC Cardiovascular Risk Collaboration.
-SCORE2-Diabetes: 10-year cardiovascular risk estimation in type 2 diabetes in Europe.
-European Heart Journal [Internet].
-2023 May 29 [cited 2023 Jul 13];ehad260.
-Available from: https://doi.org/10.1093/eurheartj/ehad260</div>
-  </div>
 </div>

### 2023-07-31 cross_sectional_FX2.R

@@ -353,32 +353,36 @@ assocdxfx <- function(X){
         RES1 = purrr::map(
             data,
             function(DAT){
-                RES2 <- tibble::tibble(DX = DISEASES)
                 clusnames <- names(DAT)[grepl("^clr", names(DAT))]
+                RES2 <- tibble::tibble(DX = DISEASES)
                 RES2 <- dplyr::mutate(
                     RES2,
                     modres = purrr::map(
                         DX,
                         function(dx, D, clusnm){
-                            mod1 <- glm(
-                                formula = reformulate(
-                                    termlabels = clusnm, 
-                                    response = dx
-                                ), 
-                                family = binomial, data = D
-                            )
-                            mod2 <- glm(
-                                formula = reformulate(
-                                    termlabels = c(clusnm, MEDICATION), 
-                                    response = dx
-                                ), 
-                                family = binomial, data = D
-                            )
-                            bind_rows(
-                                OnlyClusters = dplyr::select(broom::tidy(mod1), term, estimate, se = std.error),
-                                ClustersMed = dplyr::select(broom::tidy(mod2), term, estimate, se = std.error),
-                                .id = "model"
-                            )
+                            if(sum(D[[dx]]) > 5){
+                                mod1 <- glm(
+                                    formula = reformulate(
+                                        termlabels = clusnm, 
+                                        response = dx
+                                    ), 
+                                    family = binomial, data = D
+                                )
+                                mod2 <- glm(
+                                    formula = reformulate(
+                                        termlabels = c(clusnm, MEDICATION), 
+                                        response = dx
+                                    ), 
+                                    family = binomial, data = D
+                                )
+                                bind_rows(
+                                    OnlyClusters = dplyr::select(broom::tidy(mod1), term, estimate, se = std.error),
+                                    ClustersMed = dplyr::select(broom::tidy(mod2), term, estimate, se = std.error),
+                                    .id = "model"
+                                )
+                            } else {
+                                data.frame(term = NA, estimate = NaN)
+                            }
                         },
                         D = DAT, clusnm = clusnames
                     )

### 2023-08-04 03_ValClusOutcomes.Rmd
Add files via upload

@@
 
-## Centered log-ratio transformation
-
-Since using cluster allocation probabilities in regression is problematic because all probabilities for each individual add up to 1, we will use the centered log-ratio transformation (CLR), a technique commonly used when dealing with compositional data.
-
-The clr transformation of the probability of an individual $i$ to belong to a cluster $c$ is the following:
-
-$$
-\text{CLR}(cluster_{c})_{i} = \log\left(\frac{P(cluster_{c})_{i}}{\exp\left(\frac{1}{K}\sum_{k=1}^{K}\log(P(cluster_{k})_{i})\right)}\right)
-$$
-
-The denominator in this formula is the geometric mean of the cluster allocation probabilities of individual $i$.
-
-The reason why we chose this transformation is not only because it enables us to use cluster allocation in regression analysis, but also because it helps in the interpretability of the coefficients derived from the regression.
-In each regression we leave out the CLR of the reference cluster, which in our case is the cluster where biomarkers change concordantly with BMI.
-We have label this cluster "BC".
-Removing this CLR and only using the CLRs of the remaining clusters means that regression coefficients can be interpreted as the expected change in the dependent variable by increasing $P(cluster_{c})$ specifically at the expense of reducing $P(cluster_{BC})$ <cite id="abe09"><a href="#zotero|10831815/UYLBIXH3">(1)</a></cite>, or in other words, the effect of deviating from the concordant cluster towards a specific discordant phenotype.
-
-To calculate the transformation:
-
-```{r}
-alldat <- clrcalc(alldat)
-print(alldat)
-```
-
-Here is how the data looks like now:
-
-```{r}
-head(alldat$data[[1]])
-```
-
----
-
 ## Descriptive statistics
@@
 To quantify the association of clusters to MACE, as well as its potential contribution for prediction, we will compare two models.
-The reference model will include all predictors that are part of SCORE2, the risk stratification tool for CVD recommended by the European Society of Cardiology <cite id="pzdxs"><a href="#zotero|10831815/ZY2CL5NC">(2)</a></cite>.
-We will use a version of this score that has been validated in diabetic populations, and includes some additional clinically useful predictors <cite id="itp38"><a href="#zotero|10831815/FPUFQKFI">(3)</a></cite>.
+The reference model will include all predictors that are part of SCORE2, the risk stratification tool for CVD recommended by the European Society of Cardiology <cite id="pzdxs"><a href="#zotero|10831815/ZY2CL5NC">(SCORE2 working group and ESC Cardiovascular risk collaboration, 2021)</a></cite>.
+We will use a version of this score that has been validated in diabetic populations, and includes some additional clinically useful predictors <cite id="itp38"><a href="#zotero|10831815/FPUFQKFI">(SCORE2-Diabetes Working Group and the ESC Cardiovascular Risk Collaboration, 2023)</a></cite>.
 Additionally, for the sake of completeness, some important pre-existing conditions and pharmacological treatments, such as hypertension/antihypertensives, as well as any predictor that we had in our initial input table that are not part of SCORE2, will also be included.
@@
 
-Here `score2` contains the predictors used in SCORE2-Diabetes, plus other predictors that we have in our input table and the CLR-transformed cluster allocations.
+Here `score2` contains the predictors used in SCORE2-Diabetes, plus other predictors that we have in our input table and cluster allocations.
 The two models are contained in the last two columns.
@@
 <div class="csl-bib-body">
+  <div class="csl-entry"><i id="zotero|10831815/ZY2CL5NC"></i>SCORE2 working group and ESC Cardiovascular risk collaboration.
+(2021).
+SCORE2 risk prediction algorithms: new models to estimate 10-year risk of cardiovascular disease in Europe.
+<i>European Heart Journal</i>, <i>42</i>(25), 2439–2454.
+https://doi.org/10.1093/eurheartj/ehab309</div>
+  <div class="csl-entry"><i id="zotero|10831815/FPUFQKFI"></i>SCORE2-Diabetes Working Group and the ESC Cardiovascular Risk Collaboration.
+(2023).
+SCORE2-Diabetes: 10-year cardiovascular risk estimation in type 2 diabetes in Europe.
+<i>European Heart Journal</i>, ehad260.
+https://doi.org/10.1093/eurheartj/ehad260</div>
 </div>

### 2023-08-04 cross_sectional_FX2.R

@@ -50,40 +50,6 @@ clusterprobcalc <- function(X){
     dplyr::select(RES, -c(resids, probs))
 }
 
-# CLR transformation
-clrcalc <- function(X){
-    dplyr::mutate(
-        X,
-        data = purrr::map(
-            data,
-            ~{
-                lower_bound <- 1e-10
-                upper_bound <- 1 - lower_bound
-                RES <- dplyr::mutate(
-                    .x,
-                    dplyr::across(
-                        dplyr::starts_with("prob"),
-                        function(CLUS){
-                            CLUS <- ifelse(CLUS < lower_bound, lower_bound, CLUS)
-                            CLUS <- ifelse(CLUS > upper_bound, upper_bound, CLUS)
-                            return(CLUS)
-                        },
-                        .names = "clr{.col}"
-                    ),
-                    Gmean = exp(rowMeans(log(dplyr::pick(dplyr::starts_with("clr"))))),
-                    dplyr::across(dplyr::starts_with("clr"), ~log(.x/Gmean))
-                )
-                RES <- dplyr::rename_with(
-                    RES,
-                    function(COLN){ gsub("prob", "", COLN) },
-                    dplyr::starts_with("clrprob")
-                )
-                dplyr::select(RES, -clrBC)
-            }
-        )
-    )
-}
-
 # Variable types
 vartype <- function(x){
     if(is.numeric(x)){
@@ -189,7 +155,7 @@ markerdistribfx <- function(X){
             data,
             ~{
                 RES2 <- tidyr::expand_grid(
-                    Variable = c(BIOMARKERS, COVARIATES),
+                    Variable = c(BIOMARKERS, COVARIATES, BODYSIZEINDEX),
                     Cluster = names(.x)[grepl("^prob", names(.x))]
                 )
                 RES2 <- dplyr::mutate(
@@ -259,21 +225,12 @@ countcovarsfx <- function(X){
             function(DAT){
                 clusnams <- names(DAT)[grepl("^prob", names(DAT))]
                 DXandMed <- c(DISEASES, MEDICATION)
-                RES1 <- tibble::tibble(
-                    Cluster = "Overall",
+                RES1 <- tidyr::expand_grid(
+                    Cluster = clusnams,
                     Covariate = DXandMed
                 )
                 RES1 <- dplyr::mutate(
[…]

### 2023-08-20 03_ValClusOutcomes.Rmd
Add files via upload

@@
 
-If any of the columns in this table are missing in your data, one option is to assume that none in your population had the disease, i.e., you should have a column with all 0.
+If any of the columns in this table are missing in your data, one option is to assume that none in your population had the disease, i.e., you should have a column with 0 for all individuals.
 
@@
 ```{r}
-alldat <- mutate(
-    validclusmod,
-    data = map(sex, ~strat_dat[[.x]])
-)
-print(alldat)
+clusterdfs <- clusterprobcalc(ClusModDf = validclusmod, StratDat = strat_dat)
 ```
 
-Once we have this table, we can run the function to calculate cluster probabilities:
-
 ```{r}
-alldat <- clusterprobcalc(alldat)
-print(alldat)
+print(clusterdfs)
 ```
@@
 ```{r}
-head(alldat$data[[1]])
+head(clusterdfs$data[[1]])
 ```
@@
 ```{r}
-markerdistribdf <- markerdistribfx(alldat)
+markerdistribdf <- markerdistribfx(clusterdfs)
 ```
@@
 ```{r}
-bmieffmarkerdf <- bmieffmarkerfx(alldat)
+bmieffmarkerdf <- bmieffmarkerfx(clusterdfs)
 ```
@@
 ```{r}
-alldat <- mutate(
-    alldat,
-    data = map(data, inner_join, covar_dat, by = "eid")
-)
-print(alldat)
+clusterdfs <- addcovardat(X = clusterdfs, CovarDat = covar_dat)
+```
+
+```{r}
+print(clusterdfs)
 ```
@@
 ```{r}
-countcovarsdf <- countcovarsfx(alldat)
+countcovarsdf <- countcovarsfx(clusterdfs)
 ```
@@
 
-We are also interesting in looking at the following proportions in each cluster:
-
-- Proportion of individuals with hypertension receiving antihypertensives.
-- Proportion of individuals with CHD receiving lipid-lowering medication.
-- Proportion of individuals with T2D taking insulin.
-- Proportion of individuals with T2D taking insulin or any other anti-diabetic medication.
+We are also interesting in looking at the proportion of individuals receiving medications in each cluster, stratified by each condition. This is obtained with the following function:
 
 ```{r}
-countspectxdf <- countspectxfx(alldat)
+countdxmed <- countdxmedfx(clusterdfs)
 ```
@@
 ```{r}
-head(countspectxdf)
+head(countdxmed)
 ```
@@
 ```{r}
-assocdxdf <- assocdxfx(alldat)
+assocdxdf <- assocdxfx(clusterdfs)
 ```
@@
 ```{r}
-head(assocdxdf)
+print(assocdxdf)
 ```
@@
 ```{r}
-alldat <- alldat %>%
-    mutate(
-        macedf = purrr::map(data, addsurvmacedat, SURVDATA = survmacedat),
-        dmdf = purrr::map(data, addsurvdmdat, SURVDATA = survdmdat)
-    )
-print(alldat)
+clustersurvdfs <- addsurvdat(X = clusterdfs, SurvMACEDf = survmacedat, SurvDMDf = survdmdat)
 ```
 
-`macedf` and `dmdf` now contain the data necessary to run survival analysis.
+```{r}
+print(clustersurvdfs)
+```
+
+`data` now contain the data necessary to run survival analysis.
 
@@
 
-## Overall and cluster-specific Kaplan-Meier estimates
+## Rates of outcomes by cluster
 
-The first thing to do is to obtain overall and cluster-specific cumulative incidence rates using the Kaplan-Meier method:
+Similar to what was done in the cross sectiona setting, we will calculate the number of cases and the total follow-up in each cluster using the weighted approach:
 
 ```{r}
-kmestdf <- kmestfx(alldat)
+ratesbyclus <- ratesclusfx(clustersurvdfs)
+```
+
+```{r}
+head(ratesbyclus)
+```
+
+We will also do this by medication status:
+
+```{r}
+ratesbyclusmed <- ratesclusmedfx(clustersurvdfs)
+```
+
+```{r}
+head(ratesbyclusmed)
+```
+
+We will also calculate overall and cluster-specific cumulative incidence rates using the Kaplan-Meier method:
+
+```{r}
+kmestdf <- kmestfx(clustersurvdfs)
 ```
@@
 
-### MACE
-
 To quantify the association of clusters to MACE, as well as its potential contribution for prediction, we will compare two models.
@@
 
+Similarly, for diabetes we will also fit two models, one containing all biomarkers and another one containing the biomarkers plus the cluster probabilities.
+
+The way we will introduce the cluster probabilities into the Cox models will be employing the log-ratio transformation <cite id="0lvet"><a href="#zotero|10831815/UYLBIXH3">(Coenders &#38; Pawlowsky-Glahn, 2020)</a></cite>.
+
 #### Fitting models
@@
 ```{r}
-coxmodmacedf <- coxmodelsmace(alldat)
+coxmoddf <- coxmodels(clustersurvdfs)
 ```
@@
 ```{r}
-print(coxmodmacedf)
+print(coxmoddf)
 ```
 
-Here `score2` contains the predictors used in SCORE2-Diabetes, plus other predictors that we have in our input table and cluster allocations.
-The two models are contained in the last two columns.
-`mod_null` contains the null model, which we will use to calculate our metrics.
+Here `mod_base` contains the baseline model, while `mod_clus` contains the baseline plus clusters model. `mod_null` contains the null model, which we will use to calculate our metrics.
 
@@
 
+To properly calculate the effect of clusters, we need the coefficients estimated by the log-ratio models as well as their covariance, contained in the `estimates` and `varcovmat` columns.
+To properly calculate the expected risk for a given phenotype, we need first the means of all variables in the model, contained in the column `Means` and the parameters of the baseline hazard, contained in the `Afit` column.
+
 ```{r}
-macesurvcoefs <- macesurvcoefx(coxmodmacedf)
+survcoefs <- survcoefx(coxmoddf)
 ```
@@
 ```{r}
-head(macesurvcoefs)
+print(survcoefs)
 ```
@@
 
-We can now assess the predictive ability of each model and compare them.
-As the models are nested, we will use the gold-standard method, the likelihood ratio test.
+To assess the predictive ability of the two nested models, we will use the gold-standard method: the likelihood ratio test.
 Given the wide use of the c-statistic, we will also use this metric.
@@
 ```{r}
-compmodmacedf <- comparemodsmace(coxmodmacedf)
-compmodmacedf
+compmoddf <- comparemods(coxmoddf)
+```
+
+```{r}
+compmoddf
 ```
@@
 
-While p-values of both the likelihood ratio tests and the difference between C-statistics show evidence of added value, the adequacy index quantifies how much additional information is obtained when cluster allocations are added on top of common biomarkers contained the SCORE2.
+While p-values of both the likelihood ratio tests and the difference between C-statistics show evidence of added value, the adequacy index quantifies how much additional information is obtained when cluster allocations are added on top of the baseline model.
 
@@
 ```{r}
-adeqindbyclusmace <- AdeqIndClusMACEFx(coxmodmacedf)
+adeqindbyclus <- AdeqIndClusFx(coxmoddf)
 ```
@@
 ```{r}
-adeqindbyclusmace
+adeqindbyclus
 ```
 
-#### Adequacy index by SCORE2 probabilities
+#### Adequacy index by MACE probability given by baseline model
 
-We will also assess the adequacy index in individuals over certain thresholds of 10-year probability of MACE calculated by SCORE2.
-This is to have an idea of the utility of adding cluster allocation information across the scale of SCALE2.
+We are also interested in how the more complex model behaves along the scale of MACE probabilities given by the baseline model:
 
 ```{r}
-adeqindbypremace <- AdeqIndByPreMACEFx(coxmodmacedf)
+adeqindbypre <- AdeqIndByPreFx(coxmoddf)
 ```
@@
 ```{r}
-adeqindbypremace
+head(adeqindbypre)
 ```
@@
 ```{r}
-dcamace <- DCurvMACEFx(coxmodmacedf)
-head(dcamace)
+dcares <- DCurvFx(coxmoddf)
+```
+
+```{r}
+head(dcares)
 ```
@@
 ```{r}
-dcaclusmace <- DCurvMACEbyClFx(coxmodmacedf)
+dcaclusres <- DCurvbyClFx(coxmoddf)
 ```
@@
 ```{r}
-head(dcaclusmace)
+head(dcaclusres)
 ```
 
-### Diabetes
+## Interaction between clusters and medications
 
-For diabetes progression, the reference model will include fasting glucose, in addition to all the components of the metabolic syndrome that we already have included in our input table.
-And as before, the second model will include cluster probabilities.
-
-#### Fitting models
+Finally, we will assess the interaction between certain medications and clusters:
 
 ```{r}
-coxmoddmdf <- coxmodelsdm(alldat)
+interactmods <- interactmodfx(coxmoddf)
 ```
@@
 ```{r}
-print(coxmoddmdf)
-```
-
-#### Coefficient estimates
-
-```{r}
-dmsurvcoefs <- dmsurvcoefx(coxmoddmdf)
-```
-
-```{r}
-head(dmsurvcoefs)
-```
-
-#### Comparison of predictive ability
-
-```{r}
-compmoddmdf <- comparemodsdm(coxmoddmdf)
-compmoddmdf
-```
-
-#### Adequacy index by cluster
-
-```{r}
-adeqindbyclusdm <- AdeqIndClusDMFx(coxmoddmdf)
-```
-
-```{r}
-adeqindbyclusdm
-```
-
-#### Adequacy index by probabilities of base model
-
-```{r}
-adeqindbypredm <- AdeqIndByPreDMFx(coxmoddmdf)
-```
-
-```{r}
-adeqindbypredm
-```
-
-#### Decision curve analysis
-
-The last step in assessing clinical utility of clustering allocations is to perform a decision curve analysis. First we will assess the overall net benefit of both models:
-
-```{r}
-dcadm <- DCurvDMFx(coxmoddmdf)
-head(dcadm)
-```
-
-We will also calculate this by cluster:
-
-```{r}
-dcaclusdm <- DCurvDMbyClFx(coxmoddmdf)
-```
-
-```{r}
-head(dcaclusdm)
+print(interactmods)
 ```
@@
     CountCovars = countcovarsdf,
-    CountSpecDXMeds = countspectxdf,
+    CountDXMeds = countdxmed,
     CrossSectAssoc = assocdxdf,
+    RatesByClus = ratesbyclus,
+    RatesByClusMeds = ratesbyclusmed,
     KaplanMeierDF = kmestdf,
-    MACESurvCoefs = macesurvcoefs,
-    ComparisonMACE = compmodmacedf,
-    AdeqIndByClusMACE = adeqindbyclusmace,
-    AdeqIndByPreMACE = adeqindbypremace,
-    DCAResMACE = dcamace,
-    DCAREsClusMACE = dcaclusmace,
-    DMSurvCoefs = dmsurvcoefs,
-    ComparisonDM = compmoddmdf,
-    AdeqIndByClusDM = adeqindbyclusdm,
-    AdeqIndByPreDM = adeqindbypredm,
-    DCAResDM = dcadm,
-    DCAREsClusDM = dcaclusdm
+    SurvCoefs = survcoefs,
+    Comparison = compmoddf,
+    AdeqIndByClus = adeqindbyclus,
+    AdeqIndByPre = adeqindbypre,
+    DCARes = dcares,
+    DCAClusREs = dcaclusres,
+    InteractMods = interactmods
 )
@@
 <div class="csl-bib-body">
+  <div class="csl-entry"><i id="zotero|10831815/UYLBIXH3"></i>Coenders, G., &#38; Pawlowsky-Glahn, V.
+(2020).
+On interpretations of tests and effect sizes in regression models with a compositional predictor.
+<i>SORT-Statistics and Operations Research Transactions</i>, <i>44</i>(1), 201–220.
+https://doi.org/10.2436/20.8080.02.100</div>
   <div class="csl-entry"><i id="zotero|10831815/ZY2CL5NC"></i>SCORE2 working group and ESC Cardiovascular risk collaboration.

### 2023-08-20 cross_sectional_FX2.R

@@ -8,9 +8,11 @@ DISEASES <- c("HT", "CHD", "Stroke", "PAD", "CKD", "LiverFailure", "RA", "T2D", 
 MEDICATION <- c("Insulin", "AntiDM", "AntiHT", "LipidLower")
 
 # Calculating cluster probabilities
-clusterprobcalc <- function(X){
-    RES <- dplyr::mutate(
-        X,
+clusterprobcalc <- function(ClusModDf, StratDat){
+    alldat <- dplyr::mutate(ClusModDf, data = purrr::map(sex, ~StratDat[[.x]]))
+    dplyr::transmute(
+        alldat,
+        sex,
         # Calculate standardized residuals
         resids = purrr::map2(
             data, residmod,
@@ -45,9 +47,9 @@ clusterprobcalc <- function(X){
                 data.frame(L / rowSums(L))
             }
         ),
-        data = purrr::map2(data, probs, bind_cols)
+        data = purrr::map2(data, probs, bind_cols),
+        resids = NULL, probs = NULL
     )
-    dplyr::select(RES, -c(resids, probs))
 }
 
 # Variable types
@@ -215,6 +217,14 @@ bmieffmarkerfx <- function(X){
     dplyr::mutate(RES, Cluster = gsub("prob", "", Cluster))
 }
 
+# Adding covariate data
+addcovardat <- function(X, CovarDat){
+    dplyr::mutate(
+        X,
+        data = purrr::map(data, inner_join, CovarDat, by = "eid")
+    )
+}
+
 # Counts of individuals with diseases/medication per cluster
 countcovarsfx <- function(X){
     RES <- dplyr::transmute(
@@ -229,18 +239,18 @@ countcovarsfx <- function(X){
                     Cluster = clusnams,
                     Covariate = DXandMed
                 )
-                RES1 <- dplyr::mutate(
+                dplyr::mutate(
                     RES1,
                     Nclus = purrr::map_dbl(Cluster, ~sum(DAT[[.x]])),
-                    NclusDX = purrr::map2_dbl(
+                    Ncases = purrr::map2_dbl(
                         Cluster, Covariate,
                         function(CL, CO, D){
                             sum(D[[CO]] * D[[CL]])
                         },
                         D = DAT
-                    )
+                    ),
+                    Nnoncases = Nclus - Ncases
                 )
-                dplyr::mutate(RES1, Cluster = gsub("prob", "", Cluster))
             }
         )
     )
[…]

### 2023-09-19 cross_sectional_FX2.R
Update cross_sectional_FX2.R

@@ -431,7 +431,7 @@ kmestfx <- function(X){
                     survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ 1, 
                     data = DAT
                 )
-                RES1 <- survival:::summary.survfit(RES1, times = 10)
+                RES1 <- survival:::summary.survfit(RES1, times = 10, extend = TRUE)
                 RES1 <- RES1[c("surv", "upper", "lower")]
                 RES1 <- 1 - data.frame(RES1)
                 colnames(RES1) <- c("risk", "lower", "upper")
@@ -723,6 +723,8 @@ AdeqIndClusFx <- function(X){
                     sapply(
                         DAT[CLUS],
                         function(PROBCL){
+                            PROBCL <- ifelse(PROBCL < 1e-15, 1e-15, PROBCL)
+                            PROBCL <- ifelse(PROBCL > 1 - 1e-15, 1 - 1e-15, PROBCL)
                             MODDAT$PROB <- PROBCL
                             MODNULLCL <- survival::coxph(
                                 survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ 1, 

### 2023-09-19 cross_sectional_FX2.R
Update cross_sectional_FX2.R

@@ -416,7 +416,7 @@ ratesclusmedfx <- function(X){
 }
 
 # Kaplan-Meier estimates
-kmestfx <- function(X){
+kmestfx <- function(X, times = 10){
     RES <- dplyr::mutate(
         X,
         data = purrr::map(
@@ -431,7 +431,7 @@ kmestfx <- function(X){
                     survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ 1, 
                     data = DAT
                 )
-                RES1 <- survival:::summary.survfit(RES1, times = 10, extend = TRUE)
+                RES1 <- survival:::summary.survfit(RES1, times = times)
                 RES1 <- RES1[c("surv", "upper", "lower")]
                 RES1 <- 1 - data.frame(RES1)
                 colnames(RES1) <- c("risk", "lower", "upper")
@@ -895,7 +895,7 @@ dca <- function(formula, data, thresholds, time, weights = rep(1, nrow(data)), r
 }
 
 # DCA
-DCurvFx <- function(X){
+DCurvFx <- function(X, time = 10){
     RES <- dplyr::transmute(
         X,
         sex, outcome,
@@ -910,7 +910,7 @@ DCurvFx <- function(X){
                     base + clus,
                     data = DAT,
                     thresholds = seq(0, .25, .01),
-                    time = 10
+                    time = time
                 )
             }
         )
@@ -919,7 +919,7 @@ DCurvFx <- function(X){
 }
 
 # DCA by cluster
-DCurvbyClFx <- function(X){
+DCurvbyClFx <- function(X, time = 10){
     RES <- dplyr::transmute(
         X,
         sex, outcome,
@@ -942,7 +942,7 @@ DCurvbyClFx <- function(X){
                                 base + clus,
                                 data = DATMOD,
                                 thresholds = seq(0, .25, .01),
-                                time = 10,
+                                time = time,
                                 weights = DAT[[CL]]
                             )
                         }

### 2023-09-20 03_ValClusOutcomes.Rmd
Add files via upload

@@
 
-These two tables include individuals followed ***up to 10 years***.
-This means that any outcome after 10 years is censored.
+These two tables should include individuals followed ***up to 10 years***.
+Any outcome after 10 years should be censored.
 `outcome_value` is 1 if the person experienced the event during the follow-up time and 0 if not.
@@
 
+## Creating follow-up subsets
+
+In each subset the follow-up will be censored at a specific point in time:
+
+```{r}
+clustersurvdfs <- futsubsetsfx(clustersurvdfs)
+print(clustersurvdfs)
+```
+
+## Summary of survival data
+
+We need information on the data available for survival analysis. Here is the function:
+
+```{r}
+survsum <- survsumfx(clustersurvdfs)
+survsum
+```
+
+---
+
 ## Rates of outcomes by cluster
@@
 head(ratesbyclusmed)
-```
-
-We will also calculate overall and cluster-specific cumulative incidence rates using the Kaplan-Meier method:
-
-```{r}
-kmestdf <- kmestfx(clustersurvdfs)
-```
-
-```{r}
-head(kmestdf)
 ```
@@
     CrossSectAssoc = assocdxdf,
+    SurvSum = survsum,
     RatesByClus = ratesbyclus,
     RatesByClusMeds = ratesbyclusmed,
-    KaplanMeierDF = kmestdf,
     SurvCoefs = survcoefs,

### 2023-09-20 cross_sectional_FX2.R

@@ -347,11 +347,58 @@ addsurvdat <- function(X, SurvMACEDf, SurvDMDf){
         data = purrr::map(data, dplyr::inner_join, SurvDMDf, by = "eid"),
         data = purrr::map(data, filter, T2D == 0, T1D == 0, Insulin == 0, AntiDM == 0)
     )
-    RES <- bind_rows(X1, X2)
+    RES <- dplyr::bind_rows(X1, X2)
     dplyr::mutate(
         RES,
         data = purrr::map(data, dplyr::select, -dplyr::where(\(x) all(x == 0)))
     )
+}
+
+# Follow-up subsets
+futsubsetsfx <- function(X){
+    RES <- tidyr::expand_grid(
+        X,
+        fut = c(5, 10)
+    )
+    RES <- dplyr::mutate(
+        RES,
+        data = purrr::map2(
+            data, fut,
+            function(DAT, FUT){
+                dplyr::mutate(
+                    DAT,
+                    outcome_value = outcome_value * (outcome_timeyrs <= FUT),
+                    outcome_timeyrs = pmin(outcome_timeyrs, FUT)
+                )
+            }
+        ),
+        maxfut = map_dbl(data, ~max(.x$outcome_timeyrs)),
+        valid = maxfut > (fut - 1)
+    )
+    RES <- dplyr::filter(RES, valid)
+    dplyr::select(RES, -c(valid, maxfut))
+}
+
+# Summary function
+survsumfx <- function(X){
+   RES <- dplyr::transmute(
+       X,
+       sex, outcome, fut,
+       data = purrr::map(
+           data,
+           function(DAT){
+               data.frame(
+                   N = nrow(DAT),
+                   Ncases = sum(DAT$outcome_value, na.rm = TRUE),
+                   TPT = sum(DAT$outcome_timeyrs, na.rm = TRUE),
+                   t(setNames(quantile(DAT$outcome_timeyrs, 
+                                       probs = c(.025, .25, .5, .75, .975)),
+                              paste0("timeq", c(2.5, 25, 50, 75, 97.5))))
+                )
+            }
+        )
+    )
+    unnest(RES, data)
 }
 
 # Rates of outcome by cluster
@@ -409,56 +456,6 @@ ratesclusmedfx <- function(X){
                     TPT = sum(Med_value * outcome_timeyrs * prob),
                     .groups = "drop"
                 )
-            }
-        )
-    )
-    tidyr::unnest(RES, data)
-}
-
[…]

### 2024-02-19 cross_sectional_FX.R
Removing old files

@@ -1,456 +1,0 @@
-## Functions to run clustering analysis to detect subgroups with different relationships between BMI and risk
-
-# Declaring variables that are expected to be found in the data to run analysis
-IDCOL <- "eid"
-SEXVAR <- "sex"
-SEXGROUPS <- c("Female", "Male")
-BODYSIZEINDEX <- "bmi"
-COVARIATES <- c("age", "smoking")
-BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
-PERSEXVARS <- c(BODYSIZEINDEX, COVARIATES, BIOMARKERS)
-ALLVARS <- c(IDCOL, SEXVAR, PERSEXVARS)
-
-
-# Outlier removal
-remove_outliers <- function(x, sdunits = 5){
-    m <- mean(x, na.rm = TRUE)
-    s <- sd(x, na.rm = TRUE)
-    upperbound <- m + (sdunits*s)
-    lowerbound <- m - (sdunits*s)
-    ifelse((x > lowerbound) & (x < upperbound), x, NaN)
-}
-
-# Checking data input
-checkinput <- function(X){
-    stopifnot(is.data.frame(X))
-    stopifnot(ALLVARS %in% names(X))
-    stopifnot(names(X) %in% ALLVARS)
-}
-
-# Variable types
-vartype <- function(x){
-    if(is.numeric(x)){
-        if(length(unique(x)) > 10){
-            return("Numeric")
-        } else { 
-            return("Categorical")
-        }
-    } else {
-        if(is.character(x) | is.factor(x)){
-            return("Categorical")
-        } else {
-            return(NA)
-        }
-    }
-}
-
-# Summary of categorical variables
-categorical_sumstats <- function(x){
-    N <- length(x)
-    N_miss <- sum(is.na(x)) 
-    props <- table(x)
-    props <- sort(props, decreasing = TRUE)
-    props <- data.frame(N = N, N_miss = N_miss,
-                        Summary1 = names(props), 
-                        Summary2 = as.vector(props))
-    props$Summary2 <- paste0(props$Summary2, " (", round(100 * props$Summary2 / sum(props$Summary2), 2), "%)")
-    props
-}
-
-# Summary of continuous variables
-continuous_sumstats <- function(x){
-    N <- length(x)
-    N_miss <- sum(is.na(x))
-    Mn <- round(mean(x, na.rm = TRUE), 2)
[…]

### 2024-02-19 cross_sectional_FX2.R

@@ -1,1034 +1,0 @@
-## Functions to assess predictive ability of clusters
-
-## Columns needed
-BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
-BODYSIZEINDEX <- "bmi"
-COVARIATES <- c("age", "smoking")
-DISEASES <- c("HT", "CHD", "Stroke", "PAD", "CKD", "LiverFailure", "RA", "T2D", "T1D")
-MEDICATION <- c("Insulin", "AntiDM", "AntiHT", "LipidLower")
-
-# Calculating cluster probabilities
-clusterprobcalc <- function(ClusModDf, StratDat){
-    alldat <- dplyr::mutate(ClusModDf, data = purrr::map(sex, ~StratDat[[.x]]))
-    dplyr::transmute(
-        alldat,
-        sex,
-        # Calculate standardized residuals
-        resids = purrr::map2(
-            data, residmod,
-            function(DATA, RESIDMOD){
-                sapply(
-                    BIOMARKERS,
-                    function(biomarker){
-                        PREDICTDAT <- as.matrix(cbind(Intercept = 1, DATA[,c(BODYSIZEINDEX, COVARIATES)]))
-                        MODELDAT <- dplyr::filter(RESIDMOD, Biomarker == biomarker)
-                        RESIDSD <- MODELDAT$SDRes
-                        COEFDAT <- MODELDAT[,c("Intercept", BODYSIZEINDEX, COVARIATES)]
-                        COEFDAT <- t(COEFDAT)
-                        PREDVAL <- as.vector(PREDICTDAT %*% COEFDAT)
-                        RESIDVAL <- DATA[[biomarker]] - PREDVAL
-                        RESIDVAL / RESIDSD
-                    }
-                )
-            }
-        ),
-        # Calculate probabilities to valid clusters
-        probs = purrr::map2(
-            resids, clusmod,
-            function(RESIDDF, CLUSMOD){
-                mus <- CLUSMOD$validclus_centers
-                mus <- purrr::map(mus, function(mu){ mu[BIOMARKERS] })
-                covmats <- CLUSMOD$validclus_covmats
-                covmats <- purrr::map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
-                wts <- CLUSMOD$validclus_weights
[…]

### 2024-02-19 02_ValClus.Rmd
Adding new files

---
title: An R Markdown document converted from "02_ValClus.ipynb"
output: html_document
---

# Validation results

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(ggplot2)
```

```{r}
options(repr.plot.res = 300)
```

## Importing results

```{r}
result_tab <- tibble(
    cohort = c("ukb", "maastricht", "rotterdam", "ghs")
)
result_tab
```

```{r}
result_tab <- result_tab %>%
    mutate(
        data = map(
            cohort,
            ~rio::import(paste0("../data/", .x, "/result_file.RData"))
        )
    )
print(result_tab)
```

## General descriptives

```{r}
gendesc <- result_tab %>%
    transmute(
        cohort,
        data = map(data, "General_descriptives")
    ) %>%
    unnest(data)
head(gendesc)
```

### Table summarising biomarkers from all cohorts

```{r}
numgendesc <- gendesc %>%
    filter(Type == "Numeric") %>%
    mutate(across(c(Summary1, Summary2), ~stringr::str_replace_all(.x, c(" \\(" = " - ", "\\)$" = "")))) %>%
    separate_wider_delim(cols = Summary1, delim = " - ", names = c("Mean", "SD")) %>%
    separate_wider_delim(cols = Summary2, delim = " - ", names = c("Median", paste0("p", c(2.5, 25, 75, 97.5)))) %>%
    mutate(across(c(Mean, SD, Median, p2.5, p25, p75, p97.5), as.numeric))
head(numgendesc)
```

```{r}
write_tsv(numgendesc, "../data/numgendesc.tsv")
```

```{r}
numgendesc %>%
    select(cohort, sex, N) %>%
    unique
```

```{r}
gendesc %>%
    filter(Type == "Categorical")
```

```{r}
percsmoke <- gendesc %>%
    filter(Type == "Categorical") %>%
    transmute(
        cohort, sex,
        currentsmoke = ifelse(Summary1 == 0, "No", "Yes"),
        Summary2 = stringr::str_replace_all(Summary2, c(" \\(" = " - ", "%\\)$" = ""))
    ) %>%
    separate_wider_delim(
        cols = Summary2, delim = " - ", 
        names = c("Ncat", "Perc")
    ) %>%
    mutate(across(c(Ncat, Perc), as.numeric))
head(percsmoke)
```

```{r}
options(repr.plot.width = 5, repr.plot.height = 3)
percsmoke %>%
    filter(currentsmoke == "Yes") %>%
    mutate(
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        )
    ) %>%
    ggplot(aes(cohort, Perc)) +
    geom_col(aes(fill = cohort), show.legend = FALSE) +
    scale_fill_viridis_d(option = "A", begin = .25, end = .75, direction = -1) +
    facet_wrap(
        ~sex
    ) +
    labs(x = NULL, y = "%") + 
    theme_bw()
```

### Boxplot summarising biomarkers from all cohorts

```{r}
options(repr.plot.width = 10, repr.plot.height = 5)
numgendesc %>%
    mutate(
        Variable = toupper(Variable),
        trait = Variable,
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        )
    ) %>%
    group_by(Variable) %>%
    nest %>%
    mutate(
        data = map(
            data,
            ~.x %>%
                ggplot(aes(cohort)) +
                geom_boxplot(aes(fill = cohort, ymin = p2.5, lower = p25, middle = Median, upper = p75, ymax = p97.5),
                             stat = "identity") +
                scale_fill_viridis_d(option = "A", begin = .25, end = .75, direction = -1) +
                ggh4x::facet_nested_wrap(~trait + sex, nrow = 1) +
                theme_bw() +
                theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank())
        )
    ) %>%
    pull(data) %>%
    patchwork::wrap_plots(nrow = 2, guides = "collect") &
    theme(legend.position = "top", legend.title = element_blank())
```

## Regression coefficients

```{r}
regcoefs <- result_tab %>%
    transmute(
        cohort,
        data = map(data, "BMI_coefficients")
    ) %>%
    unnest(data)
head(regcoefs)
```

```{r}
write_tsv(regcoefs, "../data/regcoefs.tsv")
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 5)
regcoefs %>%
    filter(term == "bmi") %>%
    mutate(
        Biomarker = toupper(Biomarker),
        trait = Biomarker,
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        )
    ) %>%
    nest(data = -Biomarker) %>%
    mutate(
        data = map(
            data,
            ~.x %>%
                ggplot(aes(cohort, Estimate)) +
                geom_hline(yintercept = 0, lty = "dashed") +
                geom_linerange(aes(ymin = lowerCI, ymax = upperCI)) +
                geom_point(aes(color = cohort)) +
                scale_color_viridis_d(option = "A", begin = .25, end = .75, direction = -1) +
                ggh4x::facet_nested_wrap(~trait + sex, nrow = 1) +
                theme_bw() +
                theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title = element_blank())
        )
    ) %>%
    pull(data) %>%
    patchwork::wrap_plots(nrow = 2, guides = "collect") &
    theme(legend.position = "top")
```

## UMAP projections

```{r}
umap_projs <- result_tab %>%
    transmute(
        cohort,
        data = map(data, "Cluster_results"),
        sex = map(data, names)
    ) %>%
    unnest(c(sex, data)) %>%
    mutate(data = map(data, pluck, "umap_model", "embedding"),
           data = map(data, data.frame)) %>%
    unnest(data)
print(umap_projs)
```

Discovery and validation cohorts:

```{r}
options(repr.plot.width = 10, repr.plot.height = 6)
umap_projs %>%
    mutate(
        cohort = recode_factor(
            factor(cohort), 
            "ukb" = "UKB",
            "rotterdam" = "RS",
            "maastricht" = "MS",
            "ghs" = "GHS"
        )
    ) %>%
    ggplot(aes(X1, X2)) +
    geom_point(aes(color = cohort), shape = ".", show.legend = FALSE) +
    scale_color_viridis_d(option = "A", begin = .25, end = .75, direction = 1) +
    ggh4x::facet_nested_wrap(~cohort + sex, scales = "free", nrow = 2) +
    theme_bw() +
    theme(
        axis.text = element_blank(),
        axis.ticks = element_blank()
    ) +
    labs(x = "UMAP1", y = "UMAP2")
```

## Clusters

```{r}
options(repr.plot.width = 12, repr.plot.height = 15)
result_tab %>%
    transmute(
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        ),
        data = map(data, "Cluster_results"),
        sex = map(data, names)
    ) %>%
    unnest(c(sex, data)) %>%
    mutate(
        data = map(data, "clusters"),
        cluster = map(data, names)
    ) %>%
    unnest(cluster) %>%
    mutate(
        data = map2(data, cluster, ~.x[[.y]]),
        center = map(data, "center"),
        trait = map(center, names)
    ) %>%
    select(-data) %>%
    unnest(c(center, trait)) %>%
    mutate(trait = toupper(trait)) %>%
    ggplot(aes(center, trait)) +
    geom_vline(xintercept = 0, lty = "dashed") +
    geom_segment(aes(xend = 0, yend = trait)) +
    geom_point() +
    ggh4x::facet_nested(sex + cohort ~ cluster) +
    labs(x = "Mean residual value (SD units)", y = NULL) +
    theme_bw()
```

## Individuals allocated to their closest cluster

```{r}
closest_cluster <- result_tab %>%
    transmute(
        cohort,
        data = map(data, "Cluster_results"),
        sex = map(data, names)
    ) %>%
    unnest(c(sex, data)) %>%
    mutate(
        embed = map(data, pluck, "umap_model", "embedding"),
        embed = map(embed, data.frame),
        clusterdat = map(data, pluck, "probs"),
        clusterdat = map2(clusterdat, embed, bind_cols),
        clusterdat = map(clusterdat, pivot_longer, -c(eid, X1, X2), names_to = "cluster", values_to = "prob"),
        clusterdat = map(clusterdat, group_by, eid),
        clusterdat = map(clusterdat, slice_max, prob, with_ties = FALSE)
    ) %>%
    select(cohort, sex, clusterdat) %>%
    unnest(clusterdat)
head(closest_cluster)
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 2)
closest_cluster %>%
    mutate(
        cohort = recode_factor(
            factor(cohort), 
            "ukb" = "UKBiobank",
            "ghs" = "GHS",
            "rotterdam" = "Rotterdam",
            "maastricht" = "Maastricht"
        )
    ) %>%
    ggplot(aes(X1, X2)) +
    geom_point(aes(fill = cluster), show.legend = FALSE,
               size = .2, shape = 21, color = alpha("magenta", 0.2), stroke = .01, alpha = .8) +
    ggh4x::facet_nested(~cohort + sex, scales = "free", independent = "all") +
    theme(
        plot.background = element_rect(fill = "black"),
        panel.background = element_rect(fill = "black"),
        panel.grid = element_blank(),
        strip.background = element_rect(fill = "black", color = "white"),
        strip.text = element_text(color = "white"),
        axis.text = element_blank(),
        axis.ticks = element_blank()
    ) +
    labs(x = "UMAP1", y = "UMAP2")
```

## Discovery data

```{r}
load("../data/ukb/strat_dat.RData")
```

```{r}
map(strat_dat, head)
```

## Residualization of discovery data by each model

```{r}
BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
```

```{r}
COVARIATES <- c("bmi", "age", "smoking")
```

```{r}
predicttab <- regcoefs %>%
    nest(coefdat = -c(cohort, sex, Biomarker)) %>%
    mutate(
        coefmat = map(coefdat, ~matrix(.x$Estimate, ncol = 1, dimnames = list(.x$term, "Estimate"))),
        coefmat = map(coefmat, ~.x[c("(Intercept)", COVARIATES), "Estimate", drop = FALSE]),
        observedval = map2(sex, Biomarker, ~pull(strat_dat[[.x]], .y)),
        predictormat = map(sex, ~as.matrix(cbind(Intercept = 1, strat_dat[[.x]][,COVARIATES]))),
        predictedval = map2(predictormat, coefmat, ~as.vector(.x %*% .y)),
        residval = map2(observedval, predictedval, ~.x - .y),
        MeanRes = map_dbl(residval, mean),
        SDRes = map_dbl(residval, sd)
    )
print(predicttab)
```

```{r}
residtab <- predicttab %>%
    mutate(scaleresidval = pmap(list(residval, MeanRes, SDRes), ~(..1 - ..2) / ..3)) %>%
    select(cohort, sex, Biomarker, scaleresidval) %>%
    pivot_wider(names_from = Biomarker, values_from = scaleresidval) %>%
    mutate(eid = map(sex, ~strat_dat[[.x]]$eid)) %>%
    unnest(-c(cohort, sex)) %>%
    select(cohort, sex, eid, all_of(BIOMARKERS))
head(residtab)
```

## Cluster validation

```{r}
cluster_results <- result_tab %>%
    transmute(
        cohort,
        clusdat = map(data, "Cluster_results"),
        sex = map(clusdat, names)
    ) %>%
    unnest(c(sex, clusdat)) %>%
    mutate(clusdat = map(clusdat, "clusters"))
print(cluster_results)
```

```{r}
clusprobdat <- cluster_results %>%
    mutate(
        residdat = map2(cohort, sex, ~filter(residtab, cohort == .x, sex == .y)),
        probdat = map2(
            residdat, clusdat,
            function(DAT, CLUSDAT){
                X <- DAT[,BIOMARKERS]
                mus <- map(CLUSDAT, "center")
                mus <- map(mus, function(mu){ mu[BIOMARKERS] })
                covmats <- map(CLUSDAT, "cov")
                covmats <- map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
                wts <- map_dbl(CLUSDAT, "weight")
                pdfs <- mapply(function(mu, covmat){ mvtnorm::dmvnorm(X, mu, covmat) }, mus, covmats)
                L <- pdfs %*% diag(wts)
                colnames(L) <- colnames(pdfs)
                data.frame(eid = DAT$eid, L / rowSums(L))
            }
        )
    )
print(clusprobdat)
```

```{r}
validationtab <- clusprobdat %>%
    transmute(
        cohort, sex, 
        probdat = map(probdat, pivot_longer, -eid, names_to = "cluster", values_to = "prob")
    ) %>%
    {
        ## Discovery vs. validation cohorts
        refdat <- filter(., cohort == "ukb")
        testdat <- filter(., cohort != "ukb")
        inner_join(refdat, testdat, by = "sex", suffix = c("_ref", "_test"))
    } %>%
    mutate(
        ## Individuals with high certainty of cluster allocation in discovery model
        probdat_ref = map(probdat_ref, filter, prob >= .8),
        probdat = map2(probdat_ref, probdat_test, inner_join, by = "eid", suffix = c("_ref", "_test"))
    ) %>%
    select(sex, cohort_ref, cohort_test, probdat) %>%
    unnest(probdat) %>%
    filter(cluster_ref != "cluster_0")
head(validationtab)
```

```{r}
medianprobtab <- validationtab %>%
    group_by(sex, cohort_ref, cohort_test, cluster_ref, cluster_test) %>%
    summarise(MedProb = median(prob_test), .groups = "drop")
head(medianprobtab)
```

```{r}
validatedtab <- medianprobtab %>%
    filter(MedProb >= .8)
head(validatedtab)
```

```{r}
options(repr.plot.width = 7, repr.plot.height = 4)
validatedtab %>%
    transmute(
        sex,
        from = gsub("cluster_", "U:", cluster_ref),
        to = case_match(
            cohort_test,
            "maastricht" ~ gsub("cluster_", "M:", cluster_test),
            "rotterdam" ~ gsub("cluster_", "R:", cluster_test),
            "ghs" ~ gsub("cluster_", "G:", cluster_test)
        )
    ) %>%
    group_by(sex) %>%
    nest %>%
    mutate(
        net = map(data, ~igraph::graph_from_data_frame(.x, directed = FALSE)),
        netplot = map2(
            net, sex,
            ~.x %>%
                ggraph::ggraph(layout = "igraph", algorithm = "nicely") +
                ggraph::geom_edge_link() +
                ggraph::geom_node_point(aes(fill = gsub("^([A-Z]):.+$", "\\1", name)), 
                                        shape = 21, size = 7) +
                scale_fill_viridis_d(
                    option = "A", begin = .25, end = .75, direction = -1,
                    labels = c(U = "UKB", M = "MS", R = "RS", G = "GHS"),
                    guide = guide_legend(override.aes = list(size = 2))
                ) +
                ggraph::geom_node_text(aes(label = name), color = "white", fontface = "bold", size = 2) +
                labs(title = .y, fill = "Cohort")
        )
    ) %>%
    pull(netplot) %>%
    patchwork::wrap_plots(nrow = 1, guides = "collect") &
    theme(legend.position = "bottom")
```

### Cluster mapper

```{r}
cluster_mapper <- validatedtab %>%
    select(sex, cluster_ref, cohort_test, cluster_test) %>%
    group_by(sex, cluster_ref) %>%
    filter(n() == 3) %>%
    mutate(
        validclus_id = paste0("VC", cur_group_id())
    ) %>%
    ungroup %>%
    arrange(sex, validclus_id)
head(cluster_mapper)
```

```{r}
cluster_mapper <- cluster_mapper %>%
    {
        refdat <- transmute(., sex, validclus_id, cohort = "ukb", cluster_id = cluster_ref)
        refdat <- unique(refdat)
        testdat <- transmute(., sex, validclus_id, cohort = cohort_test, cluster_id = cluster_test)
        bind_rows(refdat, testdat)
    } %>%
    arrange(sex, validclus_id)
head(cluster_mapper)
```

```{r}
options(repr.plot.width = 12, repr.plot.height = 3)
cluster_mapper %>%
    inner_join(cluster_results) %>%
    transmute(
        cohort, sex, validclus_id, cluster_id,
        centers = map2(clusdat, cluster_id, ~tibble(trait = BIOMARKERS, center = .x[[.y]][["center"]][BIOMARKERS]))
    ) %>%
    unnest(centers) %>%
    ggplot(aes(center, trait)) +
    geom_vline(xintercept = 0, lty = "dashed") +
    geom_linerange(aes(xmin = 0, xmax = center, group = cohort, color = cohort),
                 position = position_dodge(width = .5)) +
    geom_point(aes(group = cohort, color = cohort),
               position = position_dodge(width = .5)) +
    ggh4x::facet_nested(~ sex + validclus_id) +
    theme_bw() +
    theme(legend.position = "top")
```

## Naming clusters based on centers

```{r}
cluster_namer <- cluster_mapper %>%
    mutate(
        validclus_name = case_match(
            validclus_id,
            "VC1" ~ "DHT",
            "VC2" ~ "DHG",
            "VC3" ~ "DIS",
            "VC4" ~ "DAL",
            "VC5" ~ "DLT",
            "VC6" ~ "DAL",
            "VC7" ~ "DLT",
            "VC8" ~ "DHG",
            "VC9" ~ "DIS"
        )
    )
head(cluster_namer)
```

```{r}
write_tsv(cluster_namer, "../data/cluster_namer.tsv")
```

```{r}
options(repr.plot.width = 7, repr.plot.height = 5)
cluster_mapper %>%
    inner_join(cluster_results) %>%
    transmute(
        sex, cohort, validclus_id, cluster_id,
        centers = map2(clusdat, cluster_id, ~tibble(trait = BIOMARKERS, center = .x[[.y]][["center"]][BIOMARKERS]))
    ) %>%
    unnest(centers) %>%
    inner_join(cluster_namer) %>%
    mutate(
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        )
    ) %>%
    split(f = .$sex) %>%
    map(
        ~ggplot(.x, aes(center, trait)) +
            geom_vline(xintercept = 0, lty = "dashed") +
            geom_linerange(aes(xmin = 0, xmax = center, group = cohort, color = cohort),
                           position = position_dodge(width = .5)) +
            geom_point(aes(group = cohort, color = cohort),
                       position = position_dodge(width = .5)) +
            ggh4x::facet_nested(~ sex + validclus_name) +
            theme_bw()
    ) %>%
    patchwork::wrap_plots(ncol = 1, guides = "collect") &
    theme(legend.position = "top", legend.title = element_blank(), axis.title = element_blank())
```

## Saving validated clustering model

```{r}
valid_cluspars <- cluster_results %>%
    filter(cohort == "ukb") %>%
    inner_join(cluster_namer) %>%
    transmute(
        sex, validclus_name,
        validclus_centers = unname(map2(clusdat, cluster_id, ~.x[[.y]][["center"]])),
        validclus_covmats = unname(map2(clusdat, cluster_id, ~.x[[.y]][["cov"]])),
        validclus_weights = map2_dbl(clusdat, cluster_id, ~.x[[.y]][["weight"]])
    )
print(valid_cluspars)
```

Adding concordant cluster:

```{r}
concordantclus <- valid_cluspars %>%
    group_by(sex) %>%
    summarise(
        validclus_name = "BC",
        validclus_centers = list(setNames(rep(0, length(BIOMARKERS)), BIOMARKERS)),
        validclus_covmats = list(structure(diag(rep(1, length(BIOMARKERS))), 
                                           dimnames = list(BIOMARKERS, BIOMARKERS))),
        validclus_weights = 1 - sum(validclus_weights)
    )
print(concordantclus)
```

```{r}
valid_cluspars <- valid_cluspars %>%
    bind_rows(concordantclus) %>%
    arrange(sex, desc(validclus_weights))
print(valid_cluspars)
```

```{r}
valid_cluspars <- valid_cluspars %>%
    nest(clusmod = -sex)
print(valid_cluspars)
```

Regression model coefficients:

```{r}
RGC <- regcoefs %>% 
    filter(cohort == "ukb") %>% 
    transmute(sex, Biomarker, term = gsub("\\(|\\)", "", term), Estimate) %>%
    pivot_wider(names_from = term, values_from = Estimate)
head(RGC)
```

Standard deviation of residuals:

```{r}
RESIDSD <- predicttab %>% 
    filter(cohort == "ukb") %>% 
    select(sex, Biomarker, SDRes)
head(RESIDSD)
```

```{r}
LMMODELDF <- inner_join(RGC, RESIDSD)
head(LMMODELDF)
```

```{r}
LMMODELDF <- LMMODELDF %>%
    nest(residmod = -sex)
print(LMMODELDF)
```

Final per-sex model

```{r}
validclusmod <- inner_join(LMMODELDF, valid_cluspars)
print(validclusmod)
```

```{r}
save(validclusmod, file = "../data/validclusmod.RData")
```

### 2024-02-19 08_PrevDx.Rmd
Adding new files

---
title: An R Markdown document converted from "08_PrevDx.ipynb"
output: html_document
---

# Relation with prevalent diseases

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(ggplot2)
library(Matrix,  warn.conflicts = FALSE)
library(lme4, warn.conflicts = FALSE)
library(lmerTest, warn.conflicts = FALSE)
```

```{r}
options(repr.plot.res = 300)
```

## Importing results

```{r}
result_tab <- tibble(
    cohort = c("ukb", "maastricht", "rotterdam", "ghs")
)
result_tab
```

```{r}
result_tab <- result_tab %>%
    mutate(
        data = map(
            cohort,
            ~tryCatch(rio::import(paste0("../data/", .x, "/result_file2.RData")),
                      error = \(e) NULL)
        )
    )
print(result_tab)
```

```{r}
result_crosssect <- result_tab %>%
    expand_grid(
        element_name = c(
            "MarkerDistrib",
            "BMIeffOnMarker",
            "CountCovars",
            "CountDXMeds",
            "CrossSectAssoc"
        )
    ) %>%
    transmute(
        cohort, element_name,
        element_value = map2(
            data, element_name,
            ~.x[[.y]]
        )
    )
print(result_crosssect)
```

```{r}
result_crosssect <- result_crosssect %>%
    group_by(element_name) %>%
    summarise(
        element_value = list(
            bind_rows(
                map2(
                    element_value, cohort, 
                    ~mutate(.x, cohort = .y, .before = 1)
                )
            )
        )
    )
print(result_crosssect)
```

## Color map of clusters

```{r}
load("../data/validclusmod.RData")
```

```{r}
print(validclusmod)
```

```{r}
cluscolmap <- validclusmod %>%
    transmute(
        clusmod = map(clusmod, select, Cluster = validclus_name, W = validclus_weights)
    ) %>%
    unnest(clusmod) %>%
    group_by(Cluster) %>%
    summarise(W2 = exp(sum(log((W))))) %>%
    arrange(-W2) %>%
    mutate(
        Cluster = factor(Cluster, levels = Cluster),
        ClusCol = scales::hue_pal()(n())
    )
cluscolmap
```

## Proportion of individuals in each cluster

```{r}
propclusall <- result_crosssect %>%
    filter(element_name == "MarkerDistrib") %>%
    unnest(element_value) %>%
    select(cohort, sex, Cluster, N) %>%
    unique %>%
    nest(D = -sex) %>%
    mutate(
        D = map(D, pivot_wider, names_from = Cluster, values_from = N)
    ) %>%
    unnest(D)
propclusall
```

```{r}
write_tsv(propclusall, "../data/propclusall.tsv")
```

```{r}
options(repr.plot.width = 5, repr.plot.height = 2)
propclusall %>%
    pivot_longer(-c(sex, cohort), names_to = "Cluster", values_to = "N", values_drop_na = TRUE) %>%
    group_by(sex, cohort) %>%
    inner_join(cluscolmap, by = "Cluster") %>%
    mutate(
        Cluster = factor(Cluster, levels = cluscolmap$Cluster),
        Nsum = sum(N),
        prop = N/Nsum,
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        )
    ) %>%
    ggplot(aes(interaction(cohort, Cluster), prop)) +
    geom_col(aes(fill = ClusCol)) +
    scale_fill_identity() +
    scale_x_discrete(guide = ggh4x::guide_axis_nested()) +
    scale_y_continuous(labels = scales::percent) +
    ggh4x::facet_nested(
        ~sex, 
        scales = "free",
        strip = ggh4x::strip_nested( 
            size = "variable"
        )
    ) +
    labs(x = NULL, y = NULL) + 
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 4, angle = 90),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 6),
        strip.text = element_text(size = 6),
        legend.position = "top"
    )
```

## Distribution of biomarkers

```{r}
bm_meansd <- result_crosssect %>%
    filter(element_name == "MarkerDistrib") %>%
    unnest(element_value) %>%
    filter(Type == "Numeric") %>%
    mutate(
        Summary1 = gsub("\\(|\\)", "", Summary1),
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        )
    ) %>%
    separate_wider_delim(Summary1, delim = " ", names = c("Mean", "SD")) %>%
    transmute(cohort, sex, Variable, Cluster, across(c(N, Mean, SD), as.numeric))
head(bm_meansd)
```

```{r}
bm_meansd %>%
    select(-N) %>%
    nest(D = -sex) %>%
    mutate(
        D = map(D, pivot_wider, names_from = Cluster, values_from = c(Mean, SD), names_vary = "slowest"),
        D = map2(D, sex, ~mutate(.x, sex = .y, .before = 1)),
        D = map(D, arrange, Variable)
    ) %>%
    pull(D)
```

```{r}
pooledmean <- bm_meansd %>%
    nest(D = -c(sex, Variable, Cluster)) %>%
    mutate(
        D = map(
            D,
            ~meta::metamean(n = .x$N, mean = .x$Mean, sd = .x$SD)
        ),
        D = map(
            D,
            ~tibble(
                N = sum(.x$n),
                Mean = .x$TE.fixed, 
                SD = sqrt(sum(.x$n)) * .x$seTE.fixed
            )
        )
    ) %>%
    unnest(D)
head(pooledmean)
```

```{r}
allmeansd <- pooledmean %>%
    mutate(cohort = "Pooled", .before = 1) %>%
    bind_rows(bm_meansd)
head(allmeansd)
```

```{r}
write_tsv(allmeansd, "../data/allmeansd.tsv")
```

```{r}
pooledmean %>%
    select(-N) %>%
    nest(D = -sex) %>%
    mutate(
        D = map(
            D, 
            pivot_wider, 
            names_from = Cluster, 
            values_from = c(Mean, SD), 
            names_vary = "slowest"
        ),
        D = map(D, arrange, Variable)
    ) %>%
    pull(D)
```

```{r}
bm_iqr <- result_crosssect %>%
    filter(element_name == "MarkerDistrib") %>%
    unnest(element_value) %>%
    filter(Type == "Numeric") %>%
    mutate(
        Summary2 = stringr::str_replace_all(
            Summary2,
            c(
                "\\(" = "- ",
                "\\)" = ""
            )
        ),
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        )
    ) %>%
    separate_wider_delim(
        Summary2, 
        delim = " - ", 
        names = paste(
            "calc",
            c(
                "Median",
                "minval",
                "p25",
                "p75",
                "maxval"
            ),
            sep = "_"
        )
    ) %>%
    transmute(
        cohort, sex, Variable, Cluster, N,
        across(starts_with("calc"), as.numeric)
    ) %>%
    rename_with(\(x) gsub("calc_", "", x), starts_with("calc"))
head(bm_iqr)
```

```{r}
write_tsv(bm_iqr, "../data/bm_iqr.tsv")
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 5)
bm_iqr %>%
    inner_join(cluscolmap, by = "Cluster") %>%
    mutate(
        Variable = toupper(Variable),
        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster))
    ) %>%
    ggplot(aes(y = interaction(cohort, Cluster))) +
    geom_boxplot(
        aes(
            xmin = minval,
            xlower = p25,
            xmiddle = Median,
            xupper = p75,
            xmax = maxval,
            fill = ClusCol
        ),
        stat = "identity",
        size = .5
    ) +
    scale_fill_identity() +
    scale_y_discrete(guide = ggh4x::guide_axis_nested()) +
    ggh4x::facet_nested(
        sex ~ Variable, 
        scales = "free",
        space = "free_y",
        strip = ggh4x::strip_nested(
            text_y = ggh4x::elem_list_text(angle = 0), 
            size = "variable"
        )
    ) +
    labs(x = NULL, y = NULL) + 
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 4),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 6),
        strip.text = element_text(size = 6)
    )
```

Prevalence of smoking:

```{r}
options(repr.plot.width = 7, repr.plot.height = 2.5)
result_crosssect %>%
    filter(element_name == "MarkerDistrib") %>%
    unnest(element_value) %>%
    filter(Variable == "smoking", Summary1 == 1) %>%
    mutate(
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        ),
        cohort = factor(cohort, levels = c(sort(unique(cohort), decreasing = TRUE))),
        prop = as.numeric(Summary2)/N
    ) %>%
    inner_join(cluscolmap, by = "Cluster") %>%
    mutate(
        Cluster = factor(Cluster, levels = cluscolmap$Cluster)
    ) %>%
    ggplot(aes(Cluster, prop)) +
    geom_col(aes(fill = ClusCol)) +
    scale_fill_identity() +
    scale_y_continuous(labels = scales::percent) +
    ggh4x::facet_nested(
        ~sex + cohort, 
        scales = "free",
        strip = ggh4x::strip_nested( 
            size = "variable"
        )
    ) +
    labs(x = NULL, y = NULL) + 
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 4, angle = 45, hjust = 0.5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 6),
        strip.text = element_text(size = 6),
        legend.position = "top"
    )
```

## BMI effects by cluster

```{r}
bmieffdf <- result_crosssect %>%
    filter(element_name == "BMIeffOnMarker") %>%
    unnest(element_value) %>%
    filter(term == "bmi") %>%
    select(-c(element_name, term))
head(bmieffdf)
```

```{r}
pooledBMIEff <- bmieffdf %>%
    nest(D = -c(sex, Variable, Cluster)) %>%
    mutate(
        D = map(
            D,
            ~meta::metagen(TE = .x$estimate, seTE = .x$se)
        ),
        D = map(
            D,
            ~tibble(
                cohort = c("Pooled_FE", "Pooled_RE"),
                estimate = c(.x$TE.fixed, .x$TE.random), 
                se = c(.x$seTE.fixed, .x$seTE.random)
            )
        )
    ) %>%
    unnest(D)
head(pooledBMIEff)
```

```{r}
allbmieff <- pooledBMIEff %>%
    bind_rows(bmieffdf) %>%
    mutate(
        cohort = stringr::str_replace_all(
            cohort,
            c(
                "ukb" = "UKB",
                "maastricht" = "MS",
                "rotterdam" = "RS",
                "ghs" = "GHS"
            )
        ),
        pval = 2 * pnorm(-abs(estimate/se)),
        lwr = estimate - qnorm(.975)*se,
        upr = estimate + qnorm(.975)*se
    )
head(allbmieff)
```

```{r}
write_tsv(allbmieff, "../data/allbmieff.tsv")
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 8)
allbmieff %>%
    mutate(
        cohort = factor(
            cohort,
            levels = c("UKB", "MS", "RS", "GHS", "Pooled_FE", "Pooled_RE")
        ),
        Cluster = factor(Cluster, levels = cluscolmap$Cluster),
        Variable = toupper(Variable)
    ) %>%
    ggplot(aes(cohort, estimate)) +
    geom_hline(yintercept = 0, lty = "dashed") +
    geom_linerange(
        aes(
            ymin = lwr,
            ymax = upr
        )
    ) +
    geom_point(
        aes(shape = grepl("Pooled", cohort)),
        show.legend = FALSE
    ) +
    scale_shape_manual(
        values = c(20, 23)
    ) +
    ggh4x::facet_nested(
        Variable ~ sex + Cluster, 
        scales = "free",
        strip = ggh4x::strip_nested(
            text_y = ggh4x::elem_list_text(angle = 0), 
            size = "variable"
        )
    ) +
    labs(y = "Effect per BMI unit (95% CI)", x = NULL) + 
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 7, angle = 90),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 6),
        strip.text = element_text(size = 6)
    )
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 3)
allbmieff %>%
    mutate(
        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster)),
        Variable = toupper(Variable),
        Variable = case_match(
            Variable,
            "ALT" ~ paste(Variable, "(U/L)"),
            "CRP" ~ paste(Variable, "(mg/L)"),
            c("SBP", "DBP") ~ paste(Variable, "(mmHg)"),
            c("FG", "HDL", "LDL", "TG") ~ paste(Variable, "(mmol/L)"),
            "SCR" ~ paste(Variable, "(umol/L)"),
            "WHR" ~ paste(Variable, "(cm/cm)")
        )
    ) %>%
    filter(cohort == "Pooled_RE") %>%
    ggplot(aes(estimate, Cluster)) +
    geom_vline(xintercept = 0, lty = "dashed") +
    geom_rect(
        aes(
            xmin = ifelse(Cluster == "BC", lwr, NA), 
            xmax = ifelse(Cluster == "BC", upr, NA), 
            ymin = -Inf, ymax = Inf
        ), 
        fill = "pink", alpha = 0.8,
        na.rm = TRUE
    ) +
    geom_linerange(aes(xmin = lwr, xmax = upr)) +
    geom_point() +
    scale_x_continuous(n.breaks = 3, labels = function(x) ifelse(x == 0, "0", x)) +
    facet_grid(sex ~ Variable, scales = "free", space = "free_y") +
    labs(x = "Difference in biomarker per BMI unit increase", y = NULL) +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5),
        axis.title = element_text(size = 8)
    )
```

## ORs of diseases relative to concordant cluster

```{r}
percohortORs <- result_crosssect %>%
    filter(element_name == "CountCovars") %>%
    unnest(element_value) %>%
    filter(
        Covariate %in% c(
            "CHD", "Stroke", "T2D", "HT", "T1D", "LiverFailure", "RA"
        ),
        !(cohort %in% c("maastricht", "rotterdam") & 
          Covariate %in% c("LiverFailure", "RA", "T1D")),
        !(cohort == "ghs" & Covariate == "RA")
    ) %>%
    select(-c(element_name, Nclus)) %>%
    nest(D = -c(sex, cohort, Covariate)) %>%
    mutate(
        Covariate = factor(
            Covariate,
            levels = c(
                "CHD", "Stroke", "T2D", "HT", "T1D", "LiverFailure", "RA"
            )
        ),
        mod = map(
            D, 
            ~glm(
                cbind(round(Ncases), round(Nnoncases)) ~ Cluster, 
                data = .x, family = "binomial"
            )
        )
    ) %>%
    mutate(
        mod = map(mod, broom::tidy),
        mod = map(mod, mutate, term = gsub("Cluster", "", term)),
        D = map2(D, mod, left_join, by = c("Cluster" = "term")),
        mod = NULL
    ) %>%
    unnest(D) %>%
    arrange(sex, cohort, Covariate)
head(percohortORs)
```

```{r}
pooledORs <- result_crosssect %>%
    filter(element_name == "CountCovars") %>%
    unnest(element_value) %>%
    filter(
        Covariate %in% c(
            "CHD", "Stroke", "T2D", "HT", "T1D", "LiverFailure", "RA"
        ),
        !(cohort %in% c("maastricht", "rotterdam") & Covariate %in% c("LiverFailure", "RA")),
        !(cohort == "ghs" & Covariate == "RA")
    ) %>%
    select(-c(element_name, Nclus)) %>%
    nest(D = -c(sex, Covariate)) %>%
    mutate(
        Covariate = factor(
            Covariate,
            levels = c(
                "CHD", "Stroke", "T2D", "HT", "T1D", "LiverFailure", "RA"
            )
        ),
        Pooled_FE = map(
            D,
            ~if(length(unique(.x$cohort)) > 1){
                glm(
                    cbind(round(Ncases), round(Nnoncases)) ~ Cluster, 
                    data = .x, family = binomial
                )
            } else {
                NULL
            }                
        ),
        Pooled_RE = map(
            D,
            ~if(length(unique(.x$cohort)) > 1){
                glmer(
                    cbind(round(Ncases), round(Nnoncases)) ~ Cluster + (1|cohort), 
                    data = .x, family = binomial
                )
            } else {
                NULL
            }                
        )
    ) %>%
    pivot_longer(c(Pooled_FE, Pooled_RE), names_to = "cohort", values_to = "mod") %>%
    filter(!map_lgl(mod, is.null)) %>%
    mutate(
        mod = map2(cohort, mod, ~if(.x == "Pooled_FE"){ broom::tidy(.y) } else { broom.mixed::tidy(.y) }),
        mod = map(mod, mutate, term = gsub("Cluster", "", term)),
        D = map(D, group_by, Cluster),
        D = map(D, summarise, across(c(Ncases, Nnoncases), sum)),
        D = map2(D, mod, left_join, by = c("Cluster" = "term")),
        mod = NULL
    ) %>%
    unnest(D) %>%
    select(-c(effect, group)) %>%
    arrange(sex, Covariate)
head(pooledORs)
```

```{r}
allORs <- pooledORs %>%
    bind_rows(percohortORs) %>%
    transmute(
        sex, Covariate, 
        cohort = factor(
            case_match(
                cohort,
                "ukb" ~ "UKB",
                "maastricht" ~ "MS",
                "rotterdam" ~ "RS",
                "ghs" ~ "GHS",
                .default = cohort
            ),
            levels = c("UKB", "MS", "RS", "GHS", "Pooled_FE", "Pooled_RE")
        ),      
        Cluster = gsub("prob", "", Cluster), 
        across(everything()),
        OR = exp(estimate),
        LOWERCI = exp(estimate - qnorm(1 - .05/2)*std.error),
        UPPERCI = exp(estimate + qnorm(1 - .05/2)*std.error)
    ) %>%
    arrange(sex, Covariate, Cluster, cohort)
head(allORs)
nrow(allORs)
```

```{r}
write_tsv(allORs, "../data/allORs.tsv")
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 3)
allORs %>%
    drop_na %>%
    filter(cohort %in% c("Pooled_FE", "Pooled_RE") | Covariate == "RA") %>%
    mutate(
        cohort = gsub("Pooled_", "", cohort),
        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster))
    ) %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(
        aes(xmin = LOWERCI, xmax = UPPERCI, group = cohort),
        position = position_dodge(width = .5)
    ) +
    geom_point(
        aes(group = cohort, color = cohort),
        position = position_dodge(width = .5)
    ) +
    facet_grid(sex ~ Covariate, scales = "free", space = "free_y") +
    labs(x = "OR relative to BC", y = NULL, color = "Model") +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5)
    )
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 3)
allORs %>%
    drop_na %>%
    filter(cohort == "Pooled_RE" | Covariate == "RA") %>%
    mutate(
        bigeff = OR > 10, xint = ifelse(bigeff, NA, 1),
        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster))
    ) %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(aes(xintercept = xint), lty = "dashed", na.rm = TRUE) +
    geom_linerange(
        aes(xmin = LOWERCI, xmax = UPPERCI)
    ) +
    geom_point() +
    ggh4x::facet_nested(
        sex ~ Covariate + bigeff, scales = "free", space = "free_y",
        strip = ggh4x::strip_nested(
            size = "variable",
            background_x = list(element_rect(), element_rect(fill = NA, colour = NA)),
            text_x = list(element_text(), element_text(size = 0, margin = margin())),
            by_layer_x = TRUE
        )
    ) +
    ggh4x::force_panelsizes(cols = c(1, 1, 1, .5, 1, 1, .5, 1, 1)) +
    ggh4x::scale_x_facet(!bigeff, n.breaks = 4, limits = c(0, NA)) +
    ggh4x::scale_x_facet(bigeff, n.breaks = 4) +
    labs(x = "OR relative to BC", y = NULL, color = "Model") +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5)
    )
```

## ORs of medications relative to concordant cluster

```{r}
percohortORmeds <- result_crosssect %>%
    filter(element_name == "CountCovars") %>%
    unnest(element_value) %>%
    filter(
        Covariate %in% c(
            "Insulin", "AntiHT", "LipidLower"
        ),
        !(cohort %in% c("maastricht", "rotterdam") & Covariate == "Insulin")
    ) %>%
    select(-c(element_name, Nclus)) %>%
    nest(D = -c(sex, cohort, Covariate)) %>%
    mutate(
        Covariate = factor(
            Covariate,
            levels = c(
                "Insulin", "AntiHT", "LipidLower"
            )
        ),
        mod = map(
            D, 
            ~glm(
                cbind(round(Ncases), round(Nnoncases)) ~ Cluster, 
                data = .x, family = "binomial"
            )
        )
    ) %>%
    mutate(
        mod = map(mod, broom::tidy),
        mod = map(mod, mutate, term = gsub("Cluster", "", term)),
        D = map2(D, mod, left_join, by = c("Cluster" = "term")),
        mod = NULL
    ) %>%
    unnest(D) %>%
    arrange(sex, cohort, Covariate)
head(percohortORmeds)
```

```{r}
pooledORmeds <- result_crosssect %>%
    filter(element_name == "CountCovars") %>%
    unnest(element_value) %>%
    filter(
        Covariate %in% c(
            "Insulin", "AntiHT", "LipidLower"
        ),
        !(cohort %in% c("maastricht", "rotterdam") & Covariate == "Insulin")
    ) %>%
    select(-c(element_name, Nclus)) %>%
    nest(D = -c(sex, Covariate)) %>%
    mutate(
        Covariate = factor(
            Covariate,
            levels = c(
                "Insulin", "AntiHT", "LipidLower"
            )
        ),
        Pooled_FE = map(
            D,
            ~if(length(unique(.x$cohort)) > 1){
                glm(
                    cbind(round(Ncases), round(Nnoncases)) ~ Cluster, 
                    data = .x, family = binomial
                )
            } else {
                NULL
            }                
        ),
        Pooled_RE = map(
            D,
            ~if(length(unique(.x$cohort)) > 1){
                glmer(
                    cbind(round(Ncases), round(Nnoncases)) ~ Cluster + (1|cohort), 
                    data = .x, family = binomial
                )
            } else {
                NULL
            }                
        )
    ) %>%
    pivot_longer(c(Pooled_FE, Pooled_RE), names_to = "cohort", values_to = "mod") %>%
    filter(!map_lgl(mod, is.null)) %>%
    mutate(
        mod = map2(cohort, mod, ~if(.x == "Pooled_FE"){ broom::tidy(.y) } else { broom.mixed::tidy(.y) }),
        mod = map(mod, mutate, term = gsub("Cluster", "", term)),
        D = map(D, group_by, Cluster),
        D = map(D, summarise, across(c(Ncases, Nnoncases), sum)),
        D = map2(D, mod, left_join, by = c("Cluster" = "term")),
        mod = NULL
    ) %>%
    unnest(D) %>%
    select(-c(effect, group)) %>%
    arrange(sex, Covariate)
head(pooledORmeds)
```

```{r}
allORmeds <- pooledORmeds %>%
    bind_rows(percohortORmeds) %>%
    transmute(
        sex, Covariate, 
        cohort = factor(
            case_match(
                cohort,
                "ukb" ~ "UKB",
                "maastricht" ~ "MS",
                "rotterdam" ~ "RS",
                "ghs" ~ "GHS",
                .default = cohort
            ),
            levels = c("UKB", "MSS", "RS", "GHS", "Pooled_FE", "Pooled_RE")
        ),       
        Cluster = gsub("prob", "", Cluster), 
        across(everything()),
        OR = exp(estimate),
        LOWERCI = exp(estimate - qnorm(1 - .05/2)*std.error),
        UPPERCI = exp(estimate + qnorm(1 - .05/2)*std.error)
    ) %>%
    arrange(sex, Covariate, Cluster, cohort)
head(allORmeds)
nrow(allORmeds)
```

```{r}
write_tsv(allORmeds, "../data/allORmeds.tsv")
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 3)
allORmeds %>%
    drop_na %>%
    filter(cohort %in% c("Pooled_FE", "Pooled_RE")) %>%
    mutate(
        cohort = gsub("Pooled_", "", cohort),
        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster))
    ) %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(
        aes(xmin = LOWERCI, xmax = UPPERCI, group = cohort),
        position = position_dodge(width = .5)
    ) +
    geom_point(
        aes(group = cohort, color = cohort),
        position = position_dodge(width = .5)
    ) +
    facet_grid(sex ~ Covariate, scales = "free", space = "free_y") +
    labs(x = "OR relative to BC", y = NULL, color = "Model") +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5)
    )
```

```{r}
options(repr.plot.width = 5, repr.plot.height = 3)
allORmeds %>%
    drop_na %>%
    filter(cohort == "Pooled_RE") %>%
    mutate(
        bigeff = OR > 10, xint = ifelse(bigeff, NA, 1),
        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster))
    ) %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(aes(xintercept = xint), lty = "dashed", na.rm = TRUE) +
    geom_linerange(
        aes(xmin = LOWERCI, xmax = UPPERCI)
    ) +
    geom_point() +
    ggh4x::facet_nested(
        sex ~ Covariate + bigeff, scales = "free", space = "free_y",
        strip = ggh4x::strip_nested(
            size = "variable",
            background_x = list(element_rect(), element_rect(fill = NA, colour = NA)),
            text_x = list(element_text(), element_text(size = 0, margin = margin())),
            by_layer_x = TRUE
        )
    ) +
    ggh4x::force_panelsizes(cols = c(1, .6, 1, 1)) +
    ggh4x::scale_x_facet(!bigeff, limits = c(0, NA)) +
    ggh4x::scale_x_facet(bigeff, breaks = c(30, 40, 50)) +
    labs(x = "OR relative to BC", y = NULL, color = "Model") +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5)
    )
```

## Proportion of people with diseases that are medicated

```{r}
dxmeddf <- result_crosssect %>%
    filter(element_name %in% c("CountCovars", "CountDXMeds")) %>%
    pivot_wider(names_from = element_name, values_from = element_value) %>%
    mutate(
        CountCovars = map(
            CountCovars, filter, 
            Covariate %in% c("CHD", "Stroke", "T2D")
        ),
        CountCovars = map(
            CountCovars, 
            pivot_longer, c(Ncases, Nnoncases), 
            names_to = "DxStatus", values_to = "DxStatusN"
        ),
        CountDXMeds = map(
            CountDXMeds, filter, 
            Dx %in% c("CHD", "T2D"),
            Med %in% c("NoMed", "AntiHT", "LipidLower")
        ),
        CountDXMeds = map(CountDXMeds, pivot_longer, c(Ncases, Nnoncases), 
                          names_to = "DxStatus", values_to = "DxMedInN"),
        CountDXMeds = map(CountDXMeds, rename, Covariate = Dx),
        joindat = map2(CountCovars, CountDXMeds, inner_join)
    ) %>%
    select(joindat) %>%
    unnest(joindat) %>%
    mutate(
        DxMedOutN = DxStatusN - DxMedInN,
        DxStatus = ifelse(DxStatus == "Ncases", 1, 0),
        across(c(Nclus, DxStatusN), \(x){ NULL })
    )
head(dxmeddf)
```

```{r}
dxmedmods <- dxmeddf %>%
    nest(D = -c(cohort, sex, Med, Covariate)) %>%
    mutate(
        model = map(D, ~glm(cbind(round(DxMedInN), round(DxMedOutN)) ~ Cluster * DxStatus, data = .x, family = binomial)),
        model = map(model, broom::tidy),
        D = NULL
    ) %>%
    unnest(model) %>%
    filter(term != "(Intercept)") %>%
    mutate(
        OR = exp(estimate),
        LOWERCI = exp(estimate - qnorm(1 - .05/2)*std.error),
        UPPERCI = exp(estimate + qnorm(1 - .05/2)*std.error)
    )
head(dxmedmods)
nrow(dxmedmods)
```

```{r}
pooldxmedmods <- dxmeddf %>%
    nest(D = -c(sex, Med, Covariate)) %>%
    mutate(
        Pooled_FE = map(
            D, 
            ~glm(
                cbind(round(DxMedInN), round(DxMedOutN)) ~ Cluster * DxStatus, 
                data = .x, family = binomial
            )
        ),
        Pooled_RE = map(
            D, 
            ~glmer(
                cbind(round(DxMedInN), round(DxMedOutN)) ~ Cluster * DxStatus + (1|cohort), 
                data = .x, family = binomial
            )
        ),
        D = NULL
    ) %>%
    pivot_longer(c(Pooled_FE, Pooled_RE), names_to = "cohort", values_to = "model") %>%
    mutate(
        model = map2(cohort, model, ~if(.x == "Pooled_FE"){ broom::tidy(.y) } else { broom.mixed::tidy(.y) })
    ) %>%
    unnest(model) %>%
    filter(!grepl("(Intercept)", term)) %>%
    select(-c(effect, group)) %>%
    mutate(
        OR = exp(estimate),
        LOWERCI = exp(estimate - qnorm(1 - .05/2)*std.error),
        UPPERCI = exp(estimate + qnorm(1 - .05/2)*std.error),
    )
head(pooldxmedmods)
```

```{r}
alldxmedmods <- pooldxmedmods %>%
    mutate(cohort = "Pooled") %>%
    bind_rows(dxmedmods) %>%
    transmute(
        sex, Covariate, 
        cohort = factor(
            case_match(
                cohort,
                "ukb" ~ "UKB",
                "maastricht" ~ "MS",
                "rotterdam" ~ "RS",
                "ghs" ~ "GHS",
                .default = cohort
            ),
            levels = c("UKB", "MS", "RS", "GHS", "Pooled_FE", "Pooled_RE")
        ),      
        term = gsub("prob", "", term), 
        across(everything())
    ) %>%
    arrange(sex, Covariate, term, Med, cohort)
head(alldxmedmods)
nrow(alldxmedmods)
```

```{r}
write_tsv(alldxmedmods, "../data/alldxmedmods.tsv")
```

### 2024-02-19 11_DataEthnic.Rmd
Adding new files

---
title: An R Markdown document converted from "11_DataEthnic.ipynb"
output: html_document
---

## Data from other ethnicities

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
```

---

## Baseline data

```{r}
tab_name <- "~/projects/DVA/Data/UKBiobank/ukb_57232/data_files/41806.tab"
```

Locating columns needed:

```{r}
tab_vars <- tribble(
    ~colid, ~colnam, ~coltyp,
    ## Basic
    "f.eid", "eid", "n",
    "f.53.0.0", "date0", "D",
    "f.31.0.0", "sex", "n",
    "f.21003.0.0", "age", "n",
    ## Anthropometric
    "f.21001.0.0", "bmi", "n",
    "f.48.0.0", "waist_circumference", "n",
    "f.49.0.0", "hip_circumference", "n",
    "f.50.0.0", "height", "n",
    ## Blood pressure
    "f.4080.0.0", "sbp_01", "n",
    "f.4080.0.1", "sbp_02", "n",
    "f.4079.0.0", "dbp_01", "n",
    "f.4079.0.1", "dbp_02", "n",
    ## Liver function
    "f.30620.0.0", "alt", "n",
    ## Kidney function
    "f.30700.0.0", "creatinine", "n",
    ## Inflammatory markers
    "f.30710.0.0", "crp", "n",
    ## Lipid fractions
    "f.30760.0.0", "hdl", "n",
    "f.30780.0.0", "ldl", "n",
    "f.30870.0.0", "tg", "n",
    ## Glycemia
    "f.30740.0.0", "glucose", "n",
    ## Smoking
    "f.1239.0.0", "current_smoker", "n",
    ## Caucasian ancestry
    "f.22006.0.0", "caucasian", "n",
    ## Self-reported ethnicity
    "f.21000.0.0", "srethnicity", "n"
)
```

Collecting positions of these columns in the data:

```{r}
tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
    pipe %>%
    read_table(col_names = "colid", col_types = "c") %>%
    ## Position of columns in table
    mutate(colpos = row_number()) %>%
    ## Extracting position and name of variables of interest
    inner_join(tab_vars, by = "colid")
head(tab_vars)
```

Fasting time (preprocessed from a second application):

```{r}
fast_time <- read_tsv("~/projects/DVA/Data/UKBiobank/fasting_time_ukb.tsv", show_col_types = FALSE)
head(fast_time)
```

Preparing to join with rest of columns:

```{r}
fast_time <- transmute(fast_time, eid = eid_pascal, fasting_time)
head(fast_time)
```

Preparing the `bash` command to extract columns of interest:

```{r}
column_cmd <- paste("cut -f", paste(tab_vars$colpos, collapse = ","), tab_name)
cat(column_cmd)
```

Executing the command and passing to `R`:

```{r}
crude_dat <- read_tsv(
    file = pipe(column_cmd),
    ## Replacing column header with variable names
    skip = 1, col_names = tab_vars$colnam,
    ## All columns are numeric
    col_types = paste(tab_vars$coltyp, collapse = "")
) %>%
    ## Adding fasting time
    inner_join(fast_time, by = "eid")
head(crude_dat)
```

```{r}
nrow(crude_dat)
```

```{r}
recoded_ethn <- crude_dat %>%
    ## Only ancestries other than Caucasian
    filter(is.na(caucasian)) %>%
    select(-caucasian)
nrow(recoded_ethn)
```

```{r}
recoded_ethn$srethnicity %>% table
```

```{r}
ethnicitytab <- tribble(
    ~Ethnicity, ~srethnicity,
    "SouthAsian", c(3001, 3002, 3003),
    "African", c(4, 4001, 4002, 4003),
    "Chinese", c(5)
) %>%
    unnest(srethnicity)
ethnicitytab
```

```{r}
recoded_ethn <- recoded_ethn %>%
    inner_join(ethnicitytab, by = "srethnicity")
nrow(recoded_ethn)
```

```{r}
recoded_ethn <- recoded_ethn %>%
    transmute(eid, age, Ethnicity,
              sex = ifelse(sex == 0, "Female", "Male"),
              ## Anthropometric
              bmi, 
              whr = waist_circumference / hip_circumference,
              ## Blood pressure
              sbp = rowMeans(cbind(sbp_01, sbp_02), na.rm = TRUE),
              dbp = rowMeans(cbind(dbp_01, dbp_02), na.rm = TRUE),
              ## Liver function
              alt,
              ## Kidney function
              scr = creatinine,
              ## Inflammatory markers
              crp,
              ## Lipid fractions
              hdl, tg, ldl,
              ## Glycemia - At least fasting time of 4 hours
              fg = ifelse(fasting_time >= 4, glucose, NaN),
              ## Current smoking status
              smoking = case_when(current_smoker %in% c(1,2) ~ 1,
                                  current_smoker == 0 ~ 0,
                                  TRUE ~ NaN)) %>%
    ## Only complete cases
    drop_na
head(recoded_ethn)
nrow(recoded_ethn)
```

```{r}
table(recoded_ethn$Ethnicity)
```

```{r}
recoded_ethn <- recoded_ethn %>%
    group_by(Ethnicity) %>%
    filter(n() >= 1000) %>%
    ungroup
table(recoded_ethn$Ethnicity)
```

## Covariate data

```{r}
covar_dat <- read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

## MACE data from hospital and death records

### Ocurrences in hospital records

Hospital records - main table:

```{r}
hesin <- read_tsv(
    "~/projects/DVA/Data/UKBiobank/ukb_57232/hes_data/hesin.txt",
    guess_max = 50000, show_col_types = FALSE
)
print(data.frame(head(hesin)))
```

Hospital records - Diagnosis table:

```{r}
hesin_diag <- read_tsv(
    "~/projects/DVA/Data/UKBiobank/ukb_57232/hes_data/hesin_diag.txt", 
    skip = 1,
    col_types = "nnn-c-c-",
    col_names = c("eid", "ins_index", "arr_index", "diag_icd9", "diag_icd10")
)
head(hesin_diag)
```

Joining tables:

```{r}
hesin_join <- hesin %>%
    transmute(eid, ins_index, 
              event_date = coalesce(epistart, epiend, elecdate, admidate, epiend, disdate),
              event_date = lubridate::dmy(event_date)) %>%
    inner_join(hesin_diag) %>%
    select(-c(ins_index, arr_index)) %>%
    rename(ICD_9 = diag_icd9, ICD_10 = diag_icd10) %>%
    pivot_longer(c(ICD_9, ICD_10), names_to = "CODESYS", values_to = "CODE", values_drop_na = TRUE)
head(hesin_join)
```

### Occurrences from mortality data

Mortality data - date:

```{r}
mordate <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/mortality/death.txt", show_col_types = FALSE)
head(mordate)
```

Mortality data - cause:

```{r}
morcause <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/mortality/death_cause.txt", show_col_types = FALSE)
head(morcause)
```

Mortality - Join tables:

```{r}
mor <- inner_join(mordate, morcause) %>%
    transmute(eid, 
              event_date = lubridate::dmy(date_of_death),
              CODESYS = "ICD_10", CODE = cause_icd10)
head(mor)
```

### Joining occurrences

```{r}
hesmort_outcomes <- bind_rows(
    hesin_join, mor
)
head(hesmort_outcomes)
```

### Finding MACE events

MACE codes:

```{r}
mace_codes <- read_tsv("../data/MACE_CODES.tsv", show_col_types = FALSE)
head(mace_codes)
```

```{r}
outcome_lookup <- mace_codes %>%
    group_by(OUTCOME, CODESYS) %>%
    nest %>%
    mutate(
        data = map_chr(data, ~paste0("^", .x$CODE, collapse = "|"))
    ) %>%
    ungroup
outcome_lookup
```

```{r}
outcome_lookup_res <- outcome_lookup %>%
    transmute(
        outcome = OUTCOME,
        res = map2(
            CODESYS, data,
            ~hesmort_outcomes %>% 
                filter(CODESYS == .x, 
                       grepl(.y, CODE)) %>%
                select(eid, event_date)
        )
    ) %>%
    unnest(res)
head(outcome_lookup_res)
```

### Finding first occurrences of MACE events

```{r}
mace_firstoccur <- outcome_lookup_res %>%
    select(-outcome) %>%
    group_by(eid) %>%
    slice_min(event_date, with_ties = FALSE) %>%
    ungroup
head(mace_firstoccur)
```

### First and last date for all individuals - up to 10 years of follow-up

```{r}
censordate <- lubridate::ymd("2018-02-28")
```

```{r}
datedat <- mor %>%
    select(eid, lastdate = event_date) %>%
    group_by(eid) %>%
    slice_min(lastdate, with_ties = FALSE) %>%
    ungroup %>%
    right_join(crude_dat[,c("eid", "date0")]) %>%
    transmute(
        eid, date0, 
        lastdate = replace_na(lastdate, censordate),
        lastdate = if_else(lastdate > censordate, censordate, lastdate),
        lastdate = if_else(lastdate > date0 + lubridate::years(10), 
                           date0 + lubridate::years(10), lastdate)
    ) %>%
    drop_na
head(datedat)
nrow(datedat)
```

### Preparing data for survival analysis

```{r}
survmacedat <- mace_firstoccur %>%
    transmute(
        eid,
        outcome_value = 1, 
        outcome_date = event_date
    ) %>%
    right_join(datedat, by = "eid") %>%
    filter(eid %in% recoded_ethn$eid) %>%
    mutate(
        outcome_value = replace_na(outcome_value, 0),
        outcome_date = coalesce(outcome_date, censordate),
        event_after_censor = outcome_value == 1 & outcome_date > lastdate,
        outcome_value = if_else(event_after_censor, 0, outcome_value),
        outcome_date = if_else(event_after_censor, lastdate, outcome_date),
        outcome_timeyrs = as.numeric(outcome_date - date0) / 365.25
    ) %>%
    filter(outcome_date > date0) %>%
    select(-c(date0, outcome_date, lastdate, event_after_censor))
head(survmacedat)
nrow(survmacedat)
```

---

## Diabetes progression

```{r}
primcareinds <- read_tsv("../data/primcareinds.tsv", col_names = "eid", col_types = "n")
head(primcareinds)
```

```{r}
nrow(primcareinds)
```

We will only use data from individuals that we include in our primary analysis:

```{r}
primcareinds <- primcareinds %>%
    filter(eid %in% recoded_ethn$eid)
nrow(primcareinds)
```

We can add to this set of individuals their corresponding date of recruitment in UK Biobank and last date of follow-up, which we will need for the survival analysis:

```{r}
primcareinds <- primcareinds %>%
    inner_join(datedat, by = "eid")
head(primcareinds)
nrow(primcareinds)
```

We further need to exclude individuals who already reported having diabetes at recruitment:

```{r}
primcareinds <- primcareinds %>%
    inner_join(covar_dat[,c("eid", "T1D", "T2D", "Insulin", "AntiDM")], by = "eid")
head(primcareinds)
nrow(primcareinds)
```

```{r}
primcareinds <- primcareinds %>%
    filter(T1D == 0 & T2D == 0 & Insulin == 0 & AntiDM == 0) %>%
    select(-c(T1D, T2D, Insulin, AntiDM))
head(primcareinds)
nrow(primcareinds)
```

Next, we'll extract codes registered in primary care data corresponding to clinical entities equivalent to diabetes were selected from a curated list found [here](https://github.com/drkgyoung/UK_Biobank_codelists). They are organized in the following table:

Importing results:

```{r}
dmeventsprimcare <- read_tsv(
    "../data/DM_events_PrimCare.tsv", 
    col_names = c("eid", "data_provider", "event_dt", "read_2", "read_3", "value_1", "value_2", "value_3"),
    col_types = "nncccccc"
)
head(dmeventsprimcare)
```

Only individuals included in our analysis:

```{r}
dmevents <- dmeventsprimcare %>%
    transmute(eid, outcome_date = lubridate::dmy(event_dt)) %>%
    inner_join(primcareinds)
head(dmevents)
```

We also need to extract events from hospital records and mortality data. The codes to use are the following:

```{r}
dm_codes_hes <- tribble(
    ~CODESYS, ~CODE,
    "ICD_9", "250",
    "ICD_10", "E10",
    "ICD_10", "E11"
)
```

Similar to what we did for MACE:

```{r}
hesmort_dm <- hesmort_outcomes %>%
    inner_join(primcareinds)
head(hesmort_dm)
length(unique(hesmort_dm$eid))
```

Combining data from the two sources:

```{r}
dmevents <- dm_codes_hes %>%
    group_by(CODESYS) %>%
    nest %>%
    mutate(
        data = map_chr(data, ~paste0("^", .x$CODE, collapse = "|")),
        data = map2(
            CODESYS, data,
            ~hesmort_dm %>% 
                filter(CODESYS == .x, 
                       grepl(.y, CODE)) %>%
                select(eid, outcome_date = event_date, date0, lastdate)
        )
    ) %>%
    unnest(data) %>%
    ungroup %>%
    select(-CODESYS) %>%
    bind_rows(dmevents)
head(dmevents)
```

Now we need the first event recorded:

```{r}
dmevents <- dmevents %>%
    group_by(eid) %>%
    slice_min(outcome_date, with_ties = FALSE) %>%
    ungroup
head(dmevents)
nrow(dmevents)
```

Now formatting into the survival format:

```{r}
survdmdat <- dmevents %>%
    transmute(eid, outcome_value = 1, outcome_date) %>%
    right_join(primcareinds, by = "eid") %>%
    mutate(
        outcome_value = replace_na(outcome_value, 0),
        outcome_date = coalesce(outcome_date, censordate),
        event_after_censor = outcome_value == 1 & outcome_date > lastdate,
        outcome_value = if_else(event_after_censor, 0, outcome_value),
        outcome_date = if_else(event_after_censor, lastdate, outcome_date),
        outcome_timeyrs = as.numeric(outcome_date - date0) / 365.25
    ) %>%
    filter(outcome_date > date0) %>%
    select(-c(date0, outcome_date, lastdate, event_after_censor))
head(survdmdat)
nrow(survdmdat)
```

---

## Saving tables generated

Initial input:

```{r}
write_tsv(recoded_ethn, "../data/recoded_ethn.tsv")
```

Outcome data:

- MACE

```{r}
write_tsv(survmacedat, "../data/survmacedat_ethn.tsv")
```

- Time to diabetes diagnosis:

```{r}
write_tsv(survdmdat, "../data/survdmdat_ethn.tsv")
```

### 2024-02-19 14_GRSCalc.Rmd
Adding new files

---
title: An R Markdown document converted from "14_GRSCalc.ipynb"
output: html_document
---

# Polygenic score calculation in UKB

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(rbgen)
```

```{r}
for(file_name in list.files("../data/PRSFiles")){
    trait <- gsub("\\.gz", "", file_name)
    full_name <- paste("../data/PRSFiles", file_name, sep = "/")
    summdat <- read_tsv(full_name, comment = "#", show_col_types = FALSE)
    chromosomes <- unique(summdat$chr_name)
    grs_chrom <- list()
    for(chrom in chromosomes){
        chrom_dat <- filter(summdat, chr_name == chrom)
        bgen_filename <- paste0("/ludc/Raw_Data_Archive/UKBB/imp/ukb_imp_chr", chrom, "_v3.bgen")
        bgen_index <- paste0("/ludc/Home/daniel_c/projects/DVA/Data/UKBiobank/ukb_bgi/ukb_imp_chr", chrom, "_v3.bgen.bgi")
        geno_dat <- bgen.load(filename = bgen_filename, index.filename = bgen_index, rsids = chrom_dat$rsID)
        coeffs <- geno_dat %>%
            pluck("variants") %>%
            left_join(chrom_dat, by = c("rsid" = "rsID")) %>%
            mutate(aligned = case_when(allele1 == effect_allele & allele0 == other_allele ~ list(matrix(c(0,1,2), nrow = 1)),
                                       allele0 == effect_allele & allele1 == other_allele ~ list(matrix(c(2,1,0), nrow = 1)),
                                       TRUE ~ list(NaN)),
                   aligned = map2(effect_weight, aligned, ~.x * .y)) %>%
            pull(aligned) %>%
            reduce(rbind)
        coeffs <- coeffs %>%
            array(dim = c(nrow(coeffs), 1, 3)) %>%
            tensorA::to.tensor() %>%
            tensorA::`names<-.tensor`(c("rsid", "coeff", "probs"))
        geno_dat <- geno_dat %>%
            pluck("data") %>%
            tensorA::to.tensor() %>%
            tensorA::`names<-.tensor`(c("rsid", "samples", "probs")) %>%
            tensorA::mul.tensor("probs", coeffs, "probs", by = "rsid")
        geno_dat <- geno_dat[,1,]
        if(length(dim(geno_dat)) == 2){
            geno_dat <- rowSums(geno_dat, na.rm = TRUE)
        }
        grs_chrom[[as.character(chrom)]] <- geno_dat
    }
    grs_chrom <- reduce(grs_chrom, cbind)
    if(length(dim(grs_chrom)) == 2){
        grs_chrom <- rowSums(grs_chrom, na.rm = TRUE)
    }
    grs_chrom <- data.frame(G = grs_chrom)
    colnames(grs_chrom) <- trait
    write_tsv(grs_chrom, paste0("../data/PRScoresUKB/", trait, ".tsv"))
}       
```

***Note:*** The corresponding IDs for this matrix of genetic scores would be located in:

```{r}
"~/projects/DVA/Data/UKBiobank/ukb_57232/genetic_files/ukb57232_imp_chr1_v3_s487266.sample"
```

### 2024-02-19 15_GRSAssoc.Rmd
Adding new files

---
title: An R Markdown document converted from "15_GRSAssoc.ipynb"
output: html_document
---

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(ggplot2)
library(Matrix, warn.conflicts = FALSE)
library(uwot)
options(repr.plot.res = 300)
```

```{r}
sample_file <- "~/projects/DVA/Data/UKBiobank/ukb_57232/genetic_files/ukb57232_imp_chr1_v3_s487266.sample"
sample_file <- paste("tail -n+3", sample_file, "| cut -d' ' -f1") %>%
    pipe %>%
    read_tsv(col_types = "n", col_names = "eid")
head(sample_file)
nrow(sample_file)
```

```{r}
grs_tab <- map_dfc(list.files("../data/PRScoresUKB", full.names = TRUE),
                   read_tsv, show_col_types = FALSE) %>%
    mutate(eid = sample_file$eid, .before = 1)
head(grs_tab)
```

```{r}
grs_tab <- grs_tab %>%
    mutate(across(-eid, ~as.vector(scale(.x))))
head(grs_tab)
```

```{r}
genpcdat <- {
    tab_name <- "/ludc/Home/daniel_c/projects/DVA/Data/UKBiobank/ukb_57232/data_files/41806.tab"
    tab_vars <- tibble(
        colid = c("f.eid", paste0("f.22009.0.", 1:10)),
        colnam = c("eid", paste0("gpc", 1:10))
    )
    tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
        pipe %>%
        read_table(col_names = "colid", col_types = "c") %>%
        mutate(colpos = row_number()) %>%
        inner_join(tab_vars, by = "colid")
    column_cmd <- paste("cut -f", paste(tab_vars$colpos, collapse = ","), tab_name)
    read_tsv(
        pipe(column_cmd),
        skip = 1, col_names = tab_vars$colnam,
        col_types = cols(.default = "n")
    )
}
head(genpcdat)
```

```{r}
load("../data/validclusmod.RData")
print(validclusmod)
```

```{r}
load("../data/ukb/strat_dat.RData")
print(strat_dat)
```

```{r}
validclusmod <- validclusmod %>%
    mutate(
        bmtab = map(sex, ~strat_dat[[.x]]),
        umapmod = map(
            sex, 
            ~load_uwot(paste0("../data/ukb/umapmod_", .x))
        ),
        residtab = map(
            umapmod,
            ~map(
                1:.x$nn_index$ann$getNItems(),
                function(i){
                    .x$nn_index$ann$getItemsVector(i-1)
                }
            )
        ),
        residtab = map(residtab, do.call, what = rbind),
        probtab = map2(
            residtab, clusmod,
            function(RESIDDF, CLUSMOD){
                pdfs <- mapply(
                    function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) },
                    CLUSMOD$validclus_centers, 
                    CLUSMOD$validclus_covmats
                )
                L <- pdfs %*% diag(CLUSMOD$validclus_weights)
                colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
                tibble(data.frame(L / rowSums(L)))
            }
        ),
        prsprobtab = map(
            probtab,
            mutate,
            across(everything(), ~ifelse(.x < 1e-15, 1e-15, .x)),
            across(everything(), ~ifelse(.x > 1 - 1e-15, 1 - 1e-15, .x)),
            across(-probBC, ~log(.x/probBC)),
            probBC = NULL
        ),
        prsprobtab = map2(
            bmtab, prsprobtab,
            ~bind_cols(select(.x, eid), .y)
        ),
        prsprobtab = map(
            prsprobtab,
            ~inner_join(.x, grs_tab, by = "eid")
        ),
        prsprobtab = map(
            prsprobtab,
            ~inner_join(.x, genpcdat, by = "eid")
        ),
        prsprobtab = map(
            prsprobtab,
            pivot_longer,
            starts_with("prob"),
            names_to = "Cluster",
            values_to = "logratio"
        ),
        prsprobtab = map(
            prsprobtab,
            mutate,
            Cluster = gsub("prob", "", Cluster)
        ),
        prsprobtab = map(
            prsprobtab,
            pivot_longer,
            -c(eid, Cluster, logratio, starts_with("gpc")),
            names_to = "GRSName",
            values_to = "GRSVal"
        )
    )
print(validclusmod)
```

```{r}
validclusmod$prsprobtab[[1]] %>% head
```

```{r}
clus_prs <- validclusmod %>%
    transmute(sex, prsprobtab = map(prsprobtab, nest, Data = -c(Cluster, GRSName))) %>%
    unnest(prsprobtab)
print(clus_prs)
```

```{r}
head(clus_prs$Data[[1]])
```

```{r}
clus_prs <- clus_prs %>%
    mutate(Mod = map(Data, ~lm(logratio ~ GRSVal + gpc1 + gpc2 + gpc3 + gpc4 + gpc5 + gpc6 + gpc7 + gpc8 + gpc9 + gpc10, data = .x)))
print(clus_prs)
```

```{r}
clus_prs_est <- clus_prs %>%
    transmute(
        sex, Cluster, Trait = GRSName,
        ModRes = map(Mod, broom::tidy, conf.int = TRUE)
    ) %>%
    unnest(ModRes) %>%
    filter(term == "GRSVal")
head(clus_prs_est)
```

```{r}
options(repr.plot.height = 3.5, repr.plot.width = 12)
clus_prs_est %>%
    mutate(
        across(c(estimate, conf.low, conf.high), exp),
        padj = p.adjust(p.value, "fdr")
    ) %>%
    ggplot(aes(estimate, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(aes(xmin = conf.low, xmax = conf.high)) +
    geom_point(aes(color = padj < 0.05)) +
    facet_grid(sex ~ Trait, scales = "free", space = "free_y") +
    labs(
        x = "Odds ratio of having a discordant vs concordant profile\nper SD unit of PRS", 
        y = NULL,
        color = "p < 0.05 after 5% FDR correction"
    ) +
    theme_bw() +
    theme(
        legend.position = "top",
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 6),
        axis.text.x = element_text(size = 5.5),
        axis.title.x = element_text(size = 8)
    )
```

### 2024-02-20 scripts/00_DataPrep.Rmd
Organise in folders

--- 2024-02-19 11_DataEthnic.Rmd
@@
 ---
-title: An R Markdown document converted from "11_DataEthnic.ipynb"
+title: An R Markdown document converted from "00_DataPrep.ipynb"
 output: html_document
@@
 
-## Data from other ethnicities
+# Data preparation
+
+---
+
+## Libraries
 
@@
     ## Caucasian ancestry
-    "f.22006.0.0", "caucasian", "n",
-    ## Self-reported ethnicity
-    "f.21000.0.0", "srethnicity", "n"
+    "f.22006.0.0", "caucasian", "n"
 )
@@
 
-```{r}
-nrow(crude_dat)
-```
+Recoding data:
 
 ```{r}
-recoded_ethn <- crude_dat %>%
-    ## Only ancestries other than Caucasian
-    filter(is.na(caucasian)) %>%
-    select(-caucasian)
-nrow(recoded_ethn)
-```
-
-```{r}
-recoded_ethn$srethnicity %>% table
-```
-
-```{r}
-ethnicitytab <- tribble(
-    ~Ethnicity, ~srethnicity,
-    "SouthAsian", c(3001, 3002, 3003),
-    "African", c(4, 4001, 4002, 4003),
-    "Chinese", c(5)
-) %>%
-    unnest(srethnicity)
-ethnicitytab
-```
-
-```{r}
-recoded_ethn <- recoded_ethn %>%
-    inner_join(ethnicitytab, by = "srethnicity")
-nrow(recoded_ethn)
-```
-
-```{r}
-recoded_ethn <- recoded_ethn %>%
-    transmute(eid, age, Ethnicity,
+recoded_dat <- crude_dat %>%
+    ## Only European descent individuals
+    filter(caucasian == 1) %>%
+    select(-caucasian) %>%
+    transmute(eid, age, 
               sex = ifelse(sex == 0, "Female", "Male"),
@@
     drop_na
-head(recoded_ethn)
-nrow(recoded_ethn)
+head(recoded_dat)
+nrow(recoded_dat)
+```
+
+---
+
+## Pre-existing conditions
+
+```{r}
+tab_vars <- tibble(
+    colid = c("f.eid",  "f.31.0.0"),
+    colnam = c("eid", "sex"),
+) %>%
+    bind_rows(
+        data.frame(colid = paste0("f.20002.0.", 0:33),
+                   colnam = paste("noncancercode", 1:34, sep = "_")),
+        data.frame(colid = paste0("f.20009.0.", 0:33),
+                   colnam = paste("noncancerage", 1:34, sep = "_"))
+    )
+tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
+    pipe %>%
+    read_table(col_names = "colid", col_types = "c") %>%
+    ## Position of columns in table
+    mutate(colpos = row_number()) %>%
+    ## Extracting position and name of variables of interest
+    inner_join(tab_vars, by = "colid")
+head(tab_vars)
 ```
@@
 ```{r}
-table(recoded_ethn$Ethnicity)
+column_cmd <- paste("cut -f", paste(tab_vars$colpos, collapse = ","), tab_name)
+cat(column_cmd)
 ```
@@
 ```{r}
-recoded_ethn <- recoded_ethn %>%
-    group_by(Ethnicity) %>%
-    filter(n() >= 1000) %>%
-    ungroup
-table(recoded_ethn$Ethnicity)
+dx_dat <- read_tsv(
+    file = pipe(column_cmd),
+    ## Replacing column header with variable names
+    skip = 1, col_names = tab_vars$colnam,
+    ## All columns are numeric
+    col_types = cols(.default = "n")
+)
+head(dx_dat)
 ```
 
-## Covariate data
+```{r}
+nrow(dx_dat)
+```
+
+Long format
 
 ```{r}
-covar_dat <- read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
+dx_dat2 <- dx_dat %>%
+    pivot_longer(
+        -c(eid, sex),
+        names_sep = "_",
+        names_to = c(".value", NA),
+        values_drop_na = TRUE
+    )
+head(dx_dat2)
+```
+
+Joining with coding:
+
+```{r}
+dx_coding <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/coding6.tsv",
+                      show_col_types = FALSE)
+head(dx_coding)
+```
+
+```{r}
+dx_dat2 <- dx_dat2 %>%
+    inner_join(dx_coding[, c("coding", "meaning")],
+               by = c("noncancercode" = "coding"))
+head(dx_dat2)
+```
+
+### Selection of pre-existing conditions
+
+```{r}
+selected_dx <- dx_dat2 %>%
+    group_by(eid) %>%
+    summarise(
+        HT = any(meaning %in% c("hypertension", 
+                                "essential hypertension")),
+        CHD = any(meaning %in% c("heart/cardiac problem",
+                                 "angina", "heart attack/myocardial infarction")),
+        Stroke = any(meaning %in% c("stroke", "ischaemic stroke",
+                                    "transient ischaemic attack (tia)",
+                                    "subarachnoid haemorrhage")),
+        PAD = any(meaning %in% c("arterial embolism", "peripheral vascular disease",
+                                 "leg claudication/ intermittent claudication")),
+        CKD = any(meaning %in% c("renal/kidney failure",
+                                 "renal failure requiring dialysis",
+                                 "renal failure not requiring dialysis")),
+        LiverFailure = any(meaning %in% c("liver failure/cirrhosis",
+                                          "primary biliary cirrhosis",
+                                          "alcoholic liver disease / alcoholic cirrhosis")),
+        RA = any(meaning == "rheumatoid arthritis"),
+        T2D = any(meaning %in% c("diabetes", "type 2 diabetes", "diabetic eye disease", 
+                                 "diabetic neuropathy/ulcers")),
+        T1D = any(meaning %in% c("type 1 diabetes")),
+        T2Dage = noncancerage[
+            meaning %in% c("diabetes", "type 2 diabetes", "diabetic eye disease", 
+                           "diabetic neuropathy/ulcers")
+        ][1]
+    ) %>%
+    mutate(across(-c(eid, T2Dage), ~.x * 1)) %>%
+    right_join(dx_dat[,"eid"], by = "eid") %>%
+    mutate(
+        across(-eid, ~replace_na(.x, 0)),
+        T2Dage = ifelse(T2Dage < 0, NaN, T2Dage),
+        T1D = ifelse(T2D == 1 & T2Dage < 35, 1, T1D),
+        T2D = ifelse(T1D == 1, 0, T2D),
+        T2Dage = ifelse(T2D == 1 & T1D == 0, T2Dage, 0)
+    ) %>%
+    drop_na
+head(selected_dx)
+nrow(selected_dx)
+```
+
+---
+
+## Medication data
+
+```{r}
+tab_vars <- tibble(
+    colid = "f.eid",
+    colnam = "eid",
+) %>%
+    bind_rows(
+        data.frame(colid = paste0("f.20003.0.", 0:47),
+                   colnam = paste("medcode", 1:48, sep = "_"))
+    )
+tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
+    pipe %>%
+    read_table(col_names = "colid", col_types = "c") %>%
+    ## Position of columns in table
+    mutate(colpos = row_number()) %>%
+    ## Extracting position and name of variables of interest
+    inner_join(tab_vars, by = "colid")
+head(tab_vars)
+```
+
+```{r}
+column_cmd <- paste("cut -f", paste(tab_vars$colpos, collapse = ","), tab_name)
+cat(column_cmd)
+```
+
+```{r}
+meds_dat <- read_tsv(
+    file = pipe(column_cmd),
+    ## Replacing column header with variable names
+    skip = 1, col_names = tab_vars$colnam,
+    ## All columns are numeric
+    col_types = cols(.default = "n")
+)
+head(meds_dat)
+```
+
+Long format
+
+```{r}
+meds_dat2 <- meds_dat %>%
+    pivot_longer(
+        -eid,
+        names_sep = "_",
+        names_to = c(".value", NA),
+        values_drop_na = TRUE
+    )
+head(meds_dat2)
+```
+
+Joining with coding:
+
+```{r}
+meds_coding <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/coding4.tsv",
+                      show_col_types = FALSE)
+head(meds_coding)
+```
+
+```{r}
+meds_dat2 <- meds_dat2 %>%
+    inner_join(meds_coding,
+               by = c("medcode" = "coding"))
+head(meds_dat2)
+```
+
+Map of UK Biobank medication to ATC codes (available for some medications only):
+
+```{r}
+ATC_matches <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/ATC_matches.tsv", 
+                        col_types = "ncc")
+head(ATC_matches)
+```
+
+Selected codes to use:
+
+- A10A - Insulin
+- A10B - Diabetes drugs
+- C01 - Cardiac therapy
+- C02 - Antihypertensives
+- C03 - Diuretics
+- C07 - Beta blocking agents
+- C08 - Calcium channel blockers
+- C09 - RA system drugs
+- C10 - Lipid lowering
+
+```{r}
+ATC_matches_sel <- ATC_matches %>%
+    filter(grepl("^A10|^C01|^C02|^C03|^C07|^C08|^C09|^C10", ATC_codes)) %>%
+    unique
+head(ATC_matches_sel)
+```
+
+```{r}
+selmeds <- meds_dat2 %>%
+    inner_join(ATC_matches_sel, by = c("medcode" = "ukb_code", 
+                                       "meaning" = "med_name"))
+head(selmeds)
+```
+
+Ocurrence of selected medication:
+
+```{r}
+selected_meds <- selmeds %>%
+    group_by(eid) %>%
+    summarise(
+        Insulin = any(ATC_codes %in% c("A10A")),
+        AntiDM = any(ATC_codes %in% c("A10B")),
+        AntiHT = any(grepl("^C01|^C02|^C03|^C07|^C08|^C09", ATC_codes)),
+        LipidLower = any(grepl("^C10", ATC_codes))
+    ) %>%
+    mutate(across(-eid, ~.x * 1)) %>%
+    right_join(meds_dat[,"eid"], by = "eid") %>%
+    mutate(across(-eid, ~replace_na(.x, 0)))
+head(selected_meds)
+nrow(selected_meds)
+```
+
+---
+
+## Joining pre-existing conditions and medication data
+
+```{r}
+covar_dat <- inner_join(selected_dx, selected_meds, by = "eid")
 head(covar_dat)
+nrow(covar_dat)
 ```
+
+---
 
@@
     right_join(datedat, by = "eid") %>%
-    filter(eid %in% recoded_ethn$eid) %>%
+    filter(eid %in% recoded_dat$eid) %>%
     mutate(
@@
 
+We will extract information contained in primary care data to assess who developed diabetes after recruitment.
+The first thing needed is the set of individuals for which primary care data is available - not everyone has data available:
+
+```{r}
+system(
+    paste(
+        "cut", "-f", 1, 
+        "/ludc/Home/daniel_c/projects/DVA/Data/UKBiobank/ukb_57232/gp_data/gp_clinical.txt", "|",
+        "tail", "-n", "+2", "|",
+        "sort", "|",
+        "uniq", ">", "../data/primcareinds.tsv"
+    )
+)
+```
+
 ```{r}
@@
 primcareinds <- primcareinds %>%
-    filter(eid %in% recoded_ethn$eid)
+    filter(eid %in% recoded_dat$eid)
 nrow(primcareinds)
@@
 primcareinds <- primcareinds %>%
-    inner_join(covar_dat[,c("eid", "T1D", "T2D", "Insulin", "AntiDM")], by = "eid")
+    inner_join(selected_dx[,c("eid", "T1D", "T2D")], by = "eid") %>%
+    inner_join(selected_meds[,c("eid", "Insulin", "AntiDM")])
 head(primcareinds)
@@
 They are organized in the following table:
+
+```{r}
+primcarecodes_dm <- read_tsv("../data/PrimaryCareCodesDM.txt", show_col_types = FALSE)
+head(primcarecodes_dm)
+```
+
+Saving only the codes to look for them in the primary care table:
+
+```{r}
+primcarecodes_dm %>%
+    select(Code) %>%
+    write_tsv("../data/PrimCareCodesDM.tsv", col_names = FALSE)
+```
+
+Looking for codes in primary care data table:
+
+```{r}
+system(
+    paste(
+        "grep", "-Fwf", "../data/PrimCareCodesDM.tsv",
+        "/ludc/Home/daniel_c/projects/DVA/Data/UKBiobank/ukb_57232/gp_data/gp_clinical.txt", ">",
+        "../data/DM_events_PrimCare.tsv"
+    )
+)
+```
 
@@
 ```{r}
-write_tsv(recoded_ethn, "../data/recoded_ethn.tsv")
+write_tsv(recoded_dat, "../data/recoded_dat.tsv")
 ```
+
+~~~~~~~~~~~~~~~~~~~~~~~
+
+Pre-existing conditions and medication data:
+
+```{r}
+write_tsv(covar_dat,  "../data/covar_dat.tsv")
+```
+
+~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 ```{r}
-write_tsv(survmacedat, "../data/survmacedat_ethn.tsv")
+write_tsv(survmacedat, "../data/survmacedat.tsv")
 ```
@@
 ```{r}
-write_tsv(survdmdat, "../data/survdmdat_ethn.tsv")
+write_tsv(survdmdat, "../data/survdmdat.tsv")
 ```

### 2024-02-20 scripts/01_Clustering.Rmd
Organise in folders

--- 2023-07-29 01_Clustering.Rmd
@@
 ```{r}
-source("cross_sectional_FX.R")
+source("01_cross_sectional_FX.R")
 ```
@@
 
+The `umap_model` object contains the necessary elements to map new individuals to an existing embedding, we can extract this object with the following function:
+
+```{r}
+umap_models <- get_umap_mod(cluster_results)
+```
+
+This object may be useful for future analyses.
+However, this object may contain individual data.
+Additionally, for the current analyses the only object we require from `umap_model` from other cohorts is the embedding for visual assessment.
+Therefore, we have the following function so that in `umap_model` only the embedding is retained:
+
+```{r}
+cluster_results <- clean_cluster_results(cluster_results)
+```
+
 ---
@@
 
-## Saving discovery cohort used
+## Saving other objects generated
+
+Other objects generated that we would ask to save for future analyses, but we won't ask you to upload:
+
+- The inputa data, not residualised:
 
@@
 ```
+
+- The input data, residualised:
+
+```{r}
+save(residtab, file = "../data/ukb/residtab.RData")
+```
+
+- The UMAP models:
+
+```{r}
+iwalk(
+    umap_models,
+    ~save_uwot(
+        model = .x, 
+        file = paste0("/ludc/Home/daniel_c/projects/DVA/CrossSectionalBMI/data/ukb/umapmod_", .y)
+    )
+)
+```

### 2024-07-18 scripts/01_cross_sectional_FX.R
Update scripts
Updated scripts

@@ -1,0 +1,470 @@
+## Functions to run clustering analysis to detect subgroups with different relationships between BMI and risk
+
+# Declaring variables that are expected to be found in the data to run analysis
+IDCOL <- "eid"
+SEXVAR <- "sex"
+SEXGROUPS <- c("Female", "Male")
+BODYSIZEINDEX <- "bmi"
+COVARIATES <- c("age", "smoking")
+BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
+PERSEXVARS <- c(BODYSIZEINDEX, COVARIATES, BIOMARKERS)
+ALLVARS <- c(IDCOL, SEXVAR, PERSEXVARS)
+
+
+# Outlier removal
+remove_outliers <- function(x, sdunits = 5){
+    m <- mean(x, na.rm = TRUE)
+    s <- sd(x, na.rm = TRUE)
+    upperbound <- m + (sdunits*s)
+    lowerbound <- m - (sdunits*s)
+    ifelse((x > lowerbound) & (x < upperbound), x, NaN)
+}
+
+# Checking data input
+checkinput <- function(X){
+    stopifnot(is.data.frame(X))
+    stopifnot(ALLVARS %in% names(X))
+    stopifnot(names(X) %in% ALLVARS)
+}
+
+# Variable types
+vartype <- function(x){
+    if(is.numeric(x)){
+        if(length(unique(x)) > 10){
+            return("Numeric")
+        } else { 
+            return("Categorical")
+        }
+    } else {
+        if(is.character(x) | is.factor(x)){
+            return("Categorical")
+        } else {
+            return(NA)
+        }
+    }
+}
+
+# Summary of categorical variables
+categorical_sumstats <- function(x){
+    N <- length(x)
+    N_miss <- sum(is.na(x)) 
+    props <- table(x)
+    props <- sort(props, decreasing = TRUE)
+    props <- data.frame(N = N, N_miss = N_miss,
+                        Summary1 = names(props), 
+                        Summary2 = as.vector(props))
+    props$Summary2 <- paste0(props$Summary2, " (", round(100 * props$Summary2 / sum(props$Summary2), 2), "%)")
+    props
+}
+
+# Summary of continuous variables
+continuous_sumstats <- function(x){
+    N <- length(x)
+    N_miss <- sum(is.na(x))
+    Mn <- round(mean(x, na.rm = TRUE), 2)
[…]

### 2024-07-18 scripts/18_LassoFX.R

@@ -1,0 +1,243 @@
+## Functions to assess predictive ability of clusters
+
+## Columns needed
+BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
+BODYSIZEINDEX <- "bmi"
+COVARIATES <- c("age", "smoking")
+DISEASES <- c("HT", "CHD", "Stroke", "PAD", "CKD", "LiverFailure", "RA", "T2D", "T1D")
+MEDICATION <- c("Insulin", "AntiDM", "AntiHT", "LipidLower")
+
+# Calculating cluster probabilities
+clusterprobcalc <- function(ClusModDf, StratDat){
+    alldat <- dplyr::mutate(ClusModDf, data = purrr::map(sex, ~StratDat[[.x]]))
+    dplyr::transmute(
+        alldat,
+        sex,
+        # Calculate standardized residuals
+        resids = purrr::map2(
+            data, residmod,
+            function(DATA, RESIDMOD){
+                sapply(
+                    BIOMARKERS,
+                    function(biomarker){
+                        PREDICTDAT <- as.matrix(cbind(Intercept = 1, DATA[,c(BODYSIZEINDEX, COVARIATES)]))
+                        MODELDAT <- dplyr::filter(RESIDMOD, Biomarker == biomarker)
+                        RESIDSD <- MODELDAT$SDRes
+                        COEFDAT <- MODELDAT[,c("Intercept", BODYSIZEINDEX, COVARIATES)]
+                        COEFDAT <- t(COEFDAT)
+                        PREDVAL <- as.vector(PREDICTDAT %*% COEFDAT)
+                        RESIDVAL <- DATA[[biomarker]] - PREDVAL
+                        RESIDVAL / RESIDSD
+                    }
+                )
+            }
+        ),
+        # Calculate probabilities to valid clusters
+        probs = purrr::map2(
+            resids, clusmod,
+            function(RESIDDF, CLUSMOD){
+                mus <- CLUSMOD$validclus_centers
+                mus <- purrr::map(mus, function(mu){ mu[BIOMARKERS] })
+                covmats <- CLUSMOD$validclus_covmats
+                covmats <- purrr::map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
+                wts <- CLUSMOD$validclus_weights
[…]

### 2024-07-25 scripts/16_PrevDxAdj.Rmd
Demo and other updates

---
title: An R Markdown document converted from "20_PrevDxAdj.ipynb"
output: html_document
---

# Associations with prevalent conditions adjusted for other prevalent conditions

## Libraries

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(ggplot2)
```

```{r}
options(repr.plot.res = 300)
```

## Pre-defined functions

```{r}
source("06_cross_sectional_FX.R")
```

## Stratified data

```{r}
load("../data/ukb/strat_dat.RData")
```

```{r}
map(strat_dat, head)
```

## Validated clusters

```{r}
load("../data/validclusmod.RData")
print(validclusmod)
```

## Pre-existing conditions and medications

```{r}
covar_dat <- read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

## Calculation of cluster probabilities

```{r}
clusterdfs <- clusterprobcalc(ClusModDf = validclusmod, StratDat = strat_dat)
```

```{r}
print(clusterdfs)
```

## Adding covariates

```{r}
clusterdfs <- addcovardat(X = clusterdfs, CovarDat = covar_dat)
```

```{r}
print(clusterdfs)
```

## Counting cases for each combination

```{r}
countdf <- clusterdfs %>%
    expand_grid(
        Adj = c("Un", "Dx", "Tx")
    ) %>%
    mutate(
        data = map(
            data, pivot_longer, starts_with("prob"),
            names_to = "Cluster",
            values_to = "ClusProb"
        ),
        data = map2(
            Adj, data,
            ~switch(
                .x,
                Un = group_by(.y, CHD, Cluster),
                Dx = group_by(.y, CHD, Cluster, HT, T2D, LipidLower),
                Tx = group_by(.y, CHD, Cluster, AntiHT, AntiDM, Insulin, LipidLower)
            )
        ),
        data = map(
            data,
            summarise,
            Nind = sum(ClusProb),
            .groups = "drop"
        ),
        data = map(
            data,
            pivot_wider,
            names_from = CHD, 
            values_from = Nind, 
            names_prefix = "CHD_"
        )
    )
print(countdf)
```

## Binomial regressions

```{r}
binreg <- countdf %>%
    mutate(
        data = map(
            data,
            ~glm(cbind(round(CHD_1), round(CHD_0)) ~ ., data = .x, family = binomial)
        ),
        data = map(
            data,
            broom::tidy
        )
    ) %>%
    unnest(data) %>%
    filter(grepl("Cluster", term)) %>%
    transmute(
        sex, Adj,
        Cluster = gsub("Clusterprob", "", term),
        OR = exp(estimate),
        LOWERCI = exp(estimate - qnorm(1 - .05/2)*std.error),
        UPPERCI = exp(estimate + qnorm(1 - .05/2)*std.error)
    )
head(binreg)
```

## Cluster map

```{r}
cluscolmap <- validclusmod %>%
    transmute(
        clusmod = map(clusmod, select, Cluster = validclus_name, W = validclus_weights)
    ) %>%
    unnest(clusmod) %>%
    group_by(Cluster) %>%
    summarise(W2 = exp(sum(log((W))))) %>%
    arrange(-W2) %>%
    mutate(
        Cluster = factor(Cluster, levels = Cluster),
        ClusCol = scales::hue_pal()(n())
    )
cluscolmap
```

## Plotting results

```{r}
options(repr.plot.width = 5, repr.plot.height = 3)
binreg %>%
    mutate(
        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster)),
        Adj = factor(
            Adj,
            levels = c("Un", "Dx", "Tx"),
            labels = c("None", "With\ncomorbidities", "With\ntreatment")
        ),
    ) %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed", na.rm = TRUE) +
    geom_linerange(
        aes(xmin = LOWERCI, xmax = UPPERCI, group = Adj),
        position = position_dodge(width = .5), linewidth = .5
    ) +
    geom_point(
        aes(color = Adj, group = Adj), size = .75,
        position = position_dodge(width = .5)
    ) +
    scale_color_manual(
        values = c("black", "red", "green")
    ) +
    facet_grid(
        ~sex, scales = "free", space = "free_y"
    ) +
    labs(x = "CAD OR relative to BC", y = NULL, color = "Adjustment") +
    theme_bw() +
    theme(
        legend.position = "top"
    )
```

### 2024-07-26 scripts/01_Clustering.Rmd
Update demo

@@
 
-- Elena Santos
-
 - Ali Farzaneh
@@
 
-Not everyone has to follow all the steps, as we require different things from every cohort. We have divided the participating cohorts into 4 groups:
-
 |Cohort group|Cohorts|
@@
 ||GHS|
-|*Mental health*|Girona|
-||Maastricht|
-|*Intervention*|SCALE|
-||ABOS|
 
@@
 Our objective is to test the hypothesis that clustering-based approaches can be used to better capture these subgroups.
-An overview of the steps of our pipeline to test this hypothesis is shown in Figure 1.
-
-|![image.png](attachment:f8b0dea3-1987-4811-8e4a-0877b1297a75.png)|
-|-|
-|*Figure 1. Pipeline overview*|
 
@@
 
-Since our clustering method ignores individuals with missing values for any biomarker, the input data should only contain individuals who have all biomarker values.
-From the previous analyses, we know that:
+Since our clustering method ignores individuals with missing values for any biomarker, the input data should only contain individuals who have all biomarker values. We know that:
 
-- Only including complete cases without losing too much data is possible in UK Biobank, Maastricht, GHS and ABOS.
+- Only including complete cases without losing too much data is possible in UK Biobank, Maastricht and GHS.
 - In Rotterdam the initial input table contains some values have been imputed using a random forest algorithm.
-- In Girona only a small subset of individuals have CRP values.
-- In SCALE there are values for waist but not for hip circumference, so it is not possible to calculate WHR.
-
-Based on these observations we have made the following decisions on how to deal with missing values:
-
-|How to handle missing values|Cohorts|
-|-:|-:|
-|Only include complete cases.|UK Biobank|
-||Maastricht|
-||GHS|
-||ABOS|
-|||
-|Use data that has been imputed.|Rotterdam|
-|||
-|Retain all individuals.|Girona|
-||SCALE|
-
-***For the cohorts in the last group***, to be able to apply our method, we will assume that BMI explains the variability in the biomarkers that are missing.
-This assumption is based on what we have observed in the other cohorts.
-In practice, this means that the clustering method will focus on the biomarkers that are available to group individuals into clusters.
-The input table should still have the same columns so that the functions in `cross_sectional_FX.R` work properly.
 

### 2024-07-26 scripts/01_cross_sectional_FX.R

@@ -202,7 +202,7 @@ umapclus_fx <- function(RESIDMAT){
     umapres <- uwot::umap(
         Xmat, n_components = 2, 
         n_neighbors = nn, nn_method = "annoy", n_trees = 100, n_sgd_threads = "auto", 
-        init = "pca", n_epochs = 500, approx_pow = TRUE,
+        init = "pca", n_epochs = 500,
         binary_edge_weights = TRUE, dens_scale = 1, 
         ret_extra = c("model", "fgraph"), verbose = FALSE       
     )

### 2025-11-12 scripts/14_Rebuttal.Rmd
Initial clean commit

---
title: An R Markdown document converted from "17_Rebuttal.ipynb"
output: html_document
---

# Reviewer's comments

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(purrr)
library(ggplot2)
```

```{r}
options(repr.plot.res = 300)
```

## Reviewer #1

1. The authors identified the five BMI-discordant profiles (DHT, DAL, DLT, DIS, and DHG). However, as they could easily assign these profile names, all these profiles were just characterized by a single biomarker (or a few biomarkers which are well-known to be highly associated with each other, such as TG, LDL, and HDL for the DAL profile) (lines 144–152 of pages 10–11; Figure 1 right panel). Therefore, it is unclear about the advantage of using the identified profiles compared to simply using each biomarker-based classification (with BMI). Based on the authors' aim, I would have expected profiles that were characterized by multiple biomarkers (with various weights) whose associations are not so obvious (e.g., TG vs. CRP). Hence, I have concerns about the conceptual advances of the identified discordant profiles. Even if the authors keep claiming the usefulness of the current discordant profiles, they must present the comparisons between the identified discordant profiles and the single biomarker-based profiles in their downstream analyses (e.g., the additional net benefits by including the profile information), because the current analyses just evaluate the comparisons between with vs. without the discordant profile information.

R/

The reviewer accurately points out that the clusters we have identified are largely characterised by a single biomarker being at values that are higher than the expected for their BMI.

```{r}
load("../data/validclusmod.RData")
```

```{r}
print(validclusmod)
```

```{r}
validclusmod <- validclusmod %>%
    mutate(
        umapmod = map(
            sex,
            ~uwot::load_uwot(paste0("../data/ukb/umapmod_", .x))
        )
    )
print(validclusmod)
```

Let's see what happens when we take an individual that is 1) predominantly concordant, and then 2) increase triglycerides only, or 3) increase LDL and decrease HDL (what happens in DAL):

```{r}
NewInd <- tibble(
    eid = 1:3,
    whr = 1,
    sbp = 140, 
    dbp = 80, 
    alt = 28, 
    scr = 70, 
    crp = 2.5, 
    hdl = c(1.3, 1.3, 1), 
    tg = c(1.8, 3.1, 3.1),
    ldl = c(3.5, 3.5, 4.5),
    fg = 5
)
validclusmod %>%
    filter(sex == "Male") %>%
    transmute(
        resids = map(
            residmod,
            ~.x %>%
                transmute(
                    Biomarker,
                    PredValue = Intercept + (bmi * 30) + (age * 55),
                    ResidValue = pmap(
                        list(Biomarker, PredValue, SDRes),
                        function(BM, PV, SDV, NewD = NewInd){
                            (NewD[,BM,drop=TRUE] - PV) / SDV
                        }
                    )
                ) %>%
                select(Biomarker, ResidValue) %>%
                pivot_wider(names_from = Biomarker, values_from = ResidValue) %>%
                unnest(everything()) %>%
                as.matrix
        ),
        probs = map2(
            resids, clusmod,
            function(RESIDDF, CLUSMOD){
                BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
                RESIDDF <- RESIDDF[,BIOMARKERS]
                mus <- CLUSMOD$validclus_centers
                mus <- map(mus, function(mu){ mu[BIOMARKERS] })
                covmats <- CLUSMOD$validclus_covmats
                covmats <- map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
                wts <- CLUSMOD$validclus_weights
                pdfs <- mapply(
                    function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) },
                    mus, covmats
                )
                L <- pdfs %*% diag(wts)
                colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
                tibble(data.frame(L / rowSums(L)))
            }
        )
    ) %>%
    pull(probs) %>%
    pluck(1) %>%
    round(2)
```

Now let's take the DHG profile and show what happens when a 1) concordant person has 2) isolated hyperglycaemia or 2) has also lower LDL and HDL - the other caracteristic of DHG.

```{r}
NewInd <- tibble(
    eid = 1:3,
    whr = 0.99,
    sbp = 140, 
    dbp = 80, 
    alt = 28, 
    scr = 70, 
    crp = 2.1, 
    hdl = c(1.3, 1.3, 1),
    tg = 1.8,
    ldl = c(3.5, 3.5, 2.5),
    fg = c(4.9, 6.7, 6.7)
)
validclusmod %>%
    filter(sex == "Male") %>%
    transmute(
        resids = map(
            residmod,
            ~.x %>%
                transmute(
                    Biomarker,
                    PredValue = Intercept + (bmi * 30) + (age * 55),
                    ResidValue = pmap(
                        list(Biomarker, PredValue, SDRes),
                        function(BM, PV, SDV, NewD = NewInd){
                            (NewD[,BM,drop=TRUE] - PV) / SDV
                        }
                    )
                ) %>%
                select(Biomarker, ResidValue) %>%
                pivot_wider(names_from = Biomarker, values_from = ResidValue) %>%
                unnest(everything()) %>%
                as.matrix
        ),
        probs = map2(
            resids, clusmod,
            function(RESIDDF, CLUSMOD){
                BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
                RESIDDF <- RESIDDF[,BIOMARKERS]
                mus <- CLUSMOD$validclus_centers
                mus <- map(mus, function(mu){ mu[BIOMARKERS] })
                covmats <- CLUSMOD$validclus_covmats
                covmats <- map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
                wts <- CLUSMOD$validclus_weights
                pdfs <- mapply(
                    function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) },
                    mus, covmats
                )
                L <- pdfs %*% diag(wts)
                colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
                tibble(data.frame(L / rowSums(L)))
            }
        )
    ) %>%
    pull(probs) %>%
    pluck(1) %>%
    round(2)
```

Additional information of discordant profiles to models containing single biomarker discordances - represented by biomarker-BMI ratios.

```{r}
load("../data/ukb/strat_dat.RData")
```

```{r}
map(strat_dat, head)
```

Then the covariate data:

```{r}
covar_dat <- read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

Then the survival data:

```{r}
survmacedat <- read_tsv("../data/survmacedat.tsv", show_col_types = FALSE)
head(survmacedat)
```

```{r}
survdmdat <- read_tsv("../data/survdmdat.tsv", show_col_types = FALSE)
head(survdmdat)
```

Calculate cluster probabilities:

```{r}
BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
BODYSIZEINDEX <- "bmi"
COVARIATES <- c("age", "smoking")
DISEASES <- c("HT", "CHD", "Stroke", "PAD", "CKD", "LiverFailure", "RA", "T2D", "T1D")
MEDICATION <- c("Insulin", "AntiDM", "AntiHT", "LipidLower")
```

```{r}
clusterdfs <- validclusmod %>%
    mutate(
        Data = map(sex, ~strat_dat[[.x]]),
        resids = map2(
            Data, residmod,
            function(DATA, RESIDMOD){
                sapply(
                    BIOMARKERS,
                    function(biomarker){
                        PREDICTDAT <- as.matrix(cbind(Intercept = 1, DATA[,c(BODYSIZEINDEX, COVARIATES)]))
                        MODELDAT <- filter(RESIDMOD, Biomarker == biomarker)
                        RESIDSD <- MODELDAT$SDRes
                        COEFDAT <- MODELDAT[,c("Intercept", BODYSIZEINDEX, COVARIATES)]
                        COEFDAT <- t(COEFDAT)
                        PREDVAL <- as.vector(PREDICTDAT %*% COEFDAT)
                        RESIDVAL <- DATA[[biomarker]] - PREDVAL
                        RESIDVAL / RESIDSD
                    }
                )
            }
        ),
        probs = map2(
            resids, clusmod,
            function(RESIDDF, CLUSMOD){
                mus <- CLUSMOD$validclus_centers
                mus <- map(mus, function(mu){ mu[BIOMARKERS] })
                covmats <- CLUSMOD$validclus_covmats
                covmats <- map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
                wts <- CLUSMOD$validclus_weights
                pdfs <- mapply(function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) }, mus, covmats)
                L <- pdfs %*% diag(wts)
                colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
                data.frame(L / rowSums(L))
            }
        ),
        Data = map2(Data, probs, bind_cols),
        resids = NULL, probs = NULL,
        Data = map(Data, inner_join, covar_dat, by = "eid")
    )
print(clusterdfs)
```

Models for MACE:

```{r}
clusmacedfs <- clusterdfs %>%
    mutate(
        Data = map(Data, inner_join, survmacedat, by = "eid"),
        Data = map(Data, filter, CHD == 0, Stroke == 0, PAD == 0),
        Data = map2(
            Data, sex,
            function(DATA, SEX){
                MODELDAT <- transmute(
                    DATA,
                    outcome_value,
                    outcome_timeyrs,
                    age,
                    smoking,
                    bmi,
                    sbp,
                    tchol = hdl + ldl + (tg/2.2),
                    hdl,
                    age_smoking = age*smoking,
                    age_sbp = age*sbp,
                    age_tchol = age*tchol,
                    age_hdl = age*hdl,
                    fg,
                    scr = scr / 88.42,
                    alpha = case_when(SEX == "Female" ~ -0.241, SEX == "Male" ~ -0.302),
                    kappa = case_when(SEX == "Female" ~ 0.7, SEX == "Male" ~ 0.9),
                    creat_kappa = scr / kappa,
                    minkappa = pmin(creat_kappa, 1), 
                    maxkappa = pmax(creat_kappa, 1),
                    minkappa_alpha = minkappa^alpha,
                    maxkappa_exp = maxkappa^(-1.200),
                    age_term = 0.9938^age,
                    sex_term = case_when(SEX == "Female" ~ 1.012, SEX == "Male" ~ 1), 
                    egfr = 142 * minkappa_alpha * maxkappa_exp * age_term * sex_term,
                    lnegfr = log(egfr + 0.01),
                    lnegfrsq = lnegfr^2,
                    age_fg = age*fg,
                    age_lnegfr = age*lnegfr,
                    whr, alt, crp,
                    across(
                        any_of(
                            c("T2D", "T2Dage", "HT", "CKD", "LiverFailure", "RA", "T1D",
                              "Insulin", "AntiDM", "AntiHT", "LipidLower")
                        )
                    ),
                    across(all_of(BIOMARKERS), \(x){ x/bmi }, .names = "{.col}_ratio"),
                    across(starts_with("prob"), \(x) ifelse(x < 1e-15, 1e-15, x)),
                    across(starts_with("prob"), \(x) ifelse(x > (1 - 1e-15), (1 - 1e-15), x)),
                    across(starts_with("prob"), \(P) log(P/probBC), .names = "alr{.col}"),
                    alrprobBC = NULL,
                    across(starts_with("prob"), \(P) NULL)
                )
                if(any(names(MODELDAT) == "T2D")){
                    MODELDAT$age_t2d <- MODELDAT$age * MODELDAT$T2D
                }
                select(
                    MODELDAT, 
                    -c(scr, alpha, kappa, creat_kappa, minkappa, maxkappa, 
                       minkappa_alpha, maxkappa_exp, age_term, sex_term, egfr)
                )
            }
        )
    )
print(clusmacedfs)
```

```{r}
clusmacedfs$Data[[1]] %>% names
```

```{r}
clusmacedfs %>%
    mutate(
        NullMod = map(
            Data, 
            ~survival::coxph(
                survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ 1, 
                data = .x
            )
        ),
        mod_base = map(
            Data,
            ~{
                MODELDAT <- select(.x, -starts_with("alrprob"))
                survival::coxph(
                    survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ ., 
                    data = MODELDAT
                )
            }
        ),
        mod_clus = map(
            Data,
            ~survival::coxph(
                survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ ., 
                data = .x
            )
        )
    ) %>%
    transmute(
        sex,
        LL0 = purrr::map_dbl(NullMod, logLik),
        LLBase = purrr::map_dbl(mod_base, logLik),
        NVBase = purrr::map_dbl(mod_base, \(MOD) sum(!is.na(MOD$coefficients))),
        LLBaseCl = purrr::map_dbl(mod_clus, logLik),
        NVBaseCl = purrr::map_dbl(mod_clus, \(MOD) sum(!is.na(MOD$coefficients))),
        LRTstat = 2 * abs(LLBaseCl - LLBase),
        LRTdf = NVBaseCl - NVBase,
        LRTp = stats::pchisq(
            q = LRTstat,
            df = LRTdf,
            lower.tail = FALSE
        ),
        AdeqInd = (LLBase - LL0) / (LLBaseCl - LL0),
        RelExpVar = round(100 * (1 - AdeqInd), 2)
    )
```

For diabetes:

```{r}
clusdmdfs <- clusterdfs %>%
    mutate(
        Data = map(Data, inner_join, survdmdat, by = "eid"),
        Data = map(Data, filter, T2D == 0, T1D == 0, Insulin == 0, AntiDM == 0),
        Data = purrr::map(
            Data,
            function(DATA){
                dplyr::transmute(
                    DATA,
                    outcome_timeyrs, outcome_value,
                    across(
                        any_of(
                            c(
                                BIOMARKERS,
                                BODYSIZEINDEX,
                                COVARIATES,
                                DISEASES[!DISEASES %in% c("T1D", "T2D")],
                                MEDICATION[!MEDICATION %in% c("Insulin", "AntiDM")]
                            )
                        )
                    ),
                    across(all_of(BIOMARKERS), \(x){ x/bmi }, .names = "{.col}_ratio"),
                    across(starts_with("prob"), \(x) ifelse(x < 1e-15, 1e-15, x)),
                    across(starts_with("prob"), \(x) ifelse(x > (1 - 1e-15), (1 - 1e-15), x)),
                    across(starts_with("prob"), \(P) log(P/probBC), .names = "alr{.col}"),
                    alrprobBC = NULL,
                    across(starts_with("prob"), \(P) NULL)
                )
            }
        )
    )
print(clusdmdfs)
```

```{r}
clusdmdfs$Data[[1]] %>% names
```

```{r}
clusdmdfs %>%
    mutate(
        NullMod = map(
            Data, 
            ~survival::coxph(
                survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ 1, 
                data = .x
            )
        ),
        mod_base = map(
            Data,
            ~{
                MODELDAT <- select(.x, -starts_with("alrprob"))
                survival::coxph(
                    survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ ., 
                    data = MODELDAT
                )
            }
        ),
        mod_clus = map(
            Data,
            ~survival::coxph(
                survival::Surv(time = outcome_timeyrs, event = outcome_value) ~ ., 
                data = .x
            )
        )
    ) %>%
    transmute(
        sex,
        LL0 = purrr::map_dbl(NullMod, logLik),
        LLBase = purrr::map_dbl(mod_base, logLik),
        NVBase = purrr::map_dbl(mod_base, \(MOD) sum(!is.na(MOD$coefficients))),
        LLBaseCl = purrr::map_dbl(mod_clus, logLik),
        NVBaseCl = purrr::map_dbl(mod_clus, \(MOD) sum(!is.na(MOD$coefficients))),
        LRTstat = 2 * abs(LLBaseCl - LLBase),
        LRTdf = NVBaseCl - NVBase,
        LRTp = stats::pchisq(
            q = LRTstat,
            df = LRTdf,
            lower.tail = FALSE
        ),
        AdeqInd = (LLBase - LL0) / (LLBaseCl - LL0),
        RelExpVar = round(100 * (1 - AdeqInd), 2)
    )
```

# Projection using other methods

```{r}
load("../data/ukb/residtab.RData")
```

```{r}
map(residtab, head)
```

```{r}
probdfs <- residtab %>%
    imap(
        ~{
            RESIDDF <- residtab[[.y]][,-1]
            CLUSMOD <- validclusmod$clusmod[validclusmod$sex == .y][[1]]
            pdfs <- mapply(
                function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) },
                CLUSMOD$validclus_centers, 
                CLUSMOD$validclus_covmats
            )
            L <- pdfs %*% diag(CLUSMOD$validclus_weights)
            colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
            tibble(data.frame(L / rowSums(L)))
        }
    )
```

```{r}
map(probdfs, head)
```

```{r}
maxprobdat <- probdfs %>%
    map(
        ~{
            MAXCLUS <- apply(.x, 1, \(x) colnames(.x)[which.max(x)])
            MAXCLUSP <- apply(.x, 1, max)
            tibble(data.frame(Cluster = MAXCLUS, Prob = MAXCLUSP))
        }
    )
map(maxprobdat, head)
```

```{r}
tsne <- map(
    residtab,
    ~Rtsne::Rtsne(
        .x[,-1], 
        dims = 2,
        perplexity = 30,
        max_iter = 500,
        check_duplicates = FALSE
    )
)
```

```{r}
cluscolmap <- validclusmod %>%
    transmute(
        clusmod = map(clusmod, select, Cluster = validclus_name, W = validclus_weights)
    ) %>%
    unnest(clusmod) %>%
    group_by(Cluster) %>%
    summarise(W2 = exp(sum(log((W))))) %>%
    arrange(-W2) %>%
    mutate(
        Cluster = factor(Cluster, levels = Cluster),
        ClusCol = scales::hue_pal()(n())
    )
cluscolmap
```

```{r}
options(repr.plot.res = 300, repr.plot.width = 6, repr.plot.height = 3)
tsne %>%
    map("Y") %>%
    imap(~data.frame(.x, maxprobdat[[.y]])) %>%
    bind_rows(.id = "sex") %>%
    filter(Prob >= .95) %>%
    ggplot(aes(X1, X2)) +
    geom_point(aes(color = Cluster), shape = ".", alpha = .1) +
    scale_color_discrete(
        guide = guide_legend(
            override.aes = list(shape = 19, alpha = 1)
        )
    ) +
    facet_grid(~sex) +
    theme_bw() +
    labs(x = "t-SNE dim 1", y = "t-SNE dim 2")
```

```{r}
cv_kpc <- map(
    1:5,
    function(cvn){
        set.seed(cvn)
        SAMP <- sample(60000, 10000)
        map(
            residtab,
            function(x){
                mod <- kernlab::kpca(
                    x = ~.,
                    data = x[SAMP,-1],
                    kernel = "rbfdot",
                    features = 2
                )
                kernlab::predict(mod, x[,-1])
            }
        )
    }
)
```

```{r}
cvkpcres <- cv_kpc %>%
    map(
        function(CVS){
            imap(
                CVS,
                function(DAT, SEX){
                    tibble(data.frame(DAT, maxprobdat[[SEX]]))
                }
            ) %>%
            bind_rows(.id = "sex")
        }
    ) %>%
    bind_rows(.id = "cvn")
head(cvkpcres)
```

```{r}
options(repr.plot.res = 300, repr.plot.width = 8, repr.plot.height = 3)
cvkpcres %>%
    ggplot(aes(X1, X2)) +
    geom_point(aes(color = Cluster), shape = ".", alpha = .1) +
    scale_color_discrete(
        guide = guide_legend(
            override.aes = list(shape = 19, alpha = 1)
        )
    ) +
    facet_grid(sex~cvn) +
    theme_bw() +
    labs(x = "PC1", y = "PC2")
```

## Colinearity across variables

```{r}
strat_dat %>%
    bind_rows %>%
    count(age <= 50) %>%
    mutate(prop = round(100*n/sum(n), 2))
```

```{r}
corsig <- strat_dat %>%
    map(select, -c(eid, sex, age, bmi, smoking)) %>%
    map(cor) %>%
    map(~{.x[lower.tri(.x, diag = TRUE)] <- NaN;.x}) %>%
    map(reshape2::melt, na.rm = TRUE) %>%
    bind_rows(.id = "sex") %>%
    group_by(Var1, Var2) %>%
    filter(any(abs(value) > 0.2)) %>%
    ungroup %>%
    pivot_wider(names_from = sex, names_prefix = "bf_")
corsig
```

```{r}
residtab %>%
    map(select, sbp, dbp, whr, alt, crp, hdl, tg, ldl) %>%
    map(cor) %>%
    map(~{.x[lower.tri(.x, diag = TRUE)] <- NaN;.x}) %>%
    map(reshape2::melt, na.rm = TRUE) %>%
    bind_rows(.id = "sex") %>%
    pivot_wider(names_from = sex, names_prefix = "af_") %>%
    inner_join(corsig) %>%
    transmute(
        across(starts_with("Var")),
        across(c(bf_Female, af_Female), \(x) round(x, digits = 2)),
        across(c(bf_Male, af_Male), \(x) round(x, digits = 2))
    )
```

```{r}
hclustdat <- residtab %>%
    map(select, -eid, -dbp) %>%
    map(as.matrix) %>%
    map(ClustOfVar::hclustvar)
```

```{r}
hclustdat <- residtab %>%
    map(select, -eid) %>%
    map(cor) %>%
    map(~as.dist(1 - .x)) %>%
    map(hclust, method = "single")
```

```{r}
plot(hclustdat$Female)
```

```{r}
plot(hclustdat$Male)
```

```{r}
pcdat <- residtab %>%
    map(~prcomp(.x[,-1]))
```

```{r}
options(repr.plot.res = 300, repr.plot.width = 8, repr.plot.height = 3)
pcdat %>%
    map(~{
        eigs <- .x$sdev^2
        tibble(X = paste0("PC", 1:length(eigs)),
               VarExp = eigs / sum(eigs))
    }) %>%
    bind_rows(.id = "sex") %>%
    ggplot(aes(reorder(X, desc(VarExp)), 100*VarExp)) +
    geom_col() +
    #geom_hline(yintercept = 95, lty = "dashed", color = "red") +
    facet_grid(~sex) +
    theme_bw() +
    labs(x = NULL, y = "%VarExp")
```

```{r}
options(repr.plot.res = 300, repr.plot.width = 8, repr.plot.height = 3)
pcdat %>%
    map(~{
        eigs <- .x$sdev^2
        tibble(X = paste0("PC", 1:length(eigs)),
               VarExp = eigs / sum(eigs)) %>%
        mutate(CumVarExp = cumsum(VarExp))
    }) %>%
    bind_rows(.id = "sex") %>%
    ggplot(aes(reorder(X, desc(VarExp)), 100*CumVarExp)) +
    geom_col() +
    geom_hline(yintercept = 90, lty = "dashed", color = "red") +
    facet_grid(~sex) +
    theme_bw() +
    labs(x = NULL, y = "Cumulative\n%VarExp")
```

```{r}
pcumap <- map(
    pcdat,
    ~uwot::umap(
        X = .x$x[,1:8],
        n_components = 2, 
        n_neighbors = max(10, round(10 + 15 * (log10(nrow(.x$x) - 4)))), 
        nn_method = "annoy", n_trees = 100, n_sgd_threads = "auto", 
        init = "pca", n_epochs = 500, approx_pow = TRUE,
        binary_edge_weights = TRUE, dens_scale = 1, 
        ret_extra = c("model", "fgraph"), verbose = FALSE       
    )
)
```

```{r}
cluscolmap
```

```{r}
options(repr.plot.res = 300, repr.plot.width = 7, repr.plot.height = 4)
pcumap %>%
    map("embedding") %>%
    map(data.frame) %>%
    map2(maxprobdat, bind_cols) %>%
    bind_rows(.id = "sex") %>%
    filter(Prob > .8) %>%
    mutate(Cluster = gsub("prob", "", Cluster)) %>%
    ggplot(aes(X1, X2)) +
    geom_point(aes(color = Cluster), alpha = .1, shape = ".") +
    scale_alpha_identity() +
    scale_color_manual(
        values = setNames(
            cluscolmap$ClusCol,
            cluscolmap$Cluster
        ),
        guide = guide_legend(
            override.aes = list(shape = 19, alpha = 1)
        )
    ) +     
    facet_wrap(~sex) +
    theme_bw() +
    labs(x = "UMAP1", y = "UMAP2")
```

```{r}
validclusmod %>%
    transmute(
        sex,
        modularity = map2_dbl(
            umapmod, maxprobdat,
            ~{
                Gr <- igraph::graph_from_adjacency_matrix(
                    .x$fgraph, mode = "undirected"
                )
                memb <- as.numeric(factor(.y$Cluster))
                igraph::modularity(Gr, memb)
            }   
        )
    )
```

```{r}
pcumap %>%
    imap(
        ~{
            Gr <- igraph::graph_from_adjacency_matrix(
                .x$fgraph, mode = "undirected"
            )
            MAXPROB <- maxprobdat[[.y]]
            igraph::V(Gr)$name <- 1:nrow(MAXPROB)
            MAXPROB %>%
                mutate(eid = row_number()) %>%
                nest(DAT = -Cluster) %>%
                transmute(
                    Cluster,
                    tval = map_dbl(
                        DAT,
                        function(XCL){
                            clusGr <- igraph::induced_subgraph(Gr, as.character(XCL$eid))
                            igraph::transitivity(clusGr)
                        }
                    )
                )
        }
    )
```

```{r}
pcclusparams <- pcdat %>%
    map2(
        probdfs, 
        function(PCD, PRB){
            apply(
                PRB, 2, 
                function(clus){
                    cov.wt(
                        x = PCD$x[,1:8], 
                        wt = clus
                    )
                }
            )
        }
    )
```

```{r}
coord_straightpolar <- function(theta = 'x', start = 0, direction = 1, clip = "on") {
  theta <- match.arg(theta, c("x", "y"))
  r <- if (theta == "x") 
    "y"
  else "x"
  ggproto(NULL, CoordPolar, theta = theta, r = r, start = start,
          direction = sign(direction), clip = clip,
          # This is the different bit
          is_linear = function(){TRUE})
}
```

```{r}
pcclusparams %>%
    map(map, "center") %>%
    map(
        map, 
        ~tibble(
            VAR = names(.x), 
            value = .x
        )
    ) %>%
    map(bind_rows, .id = "Cluster") %>%
    bind_rows(.id = "sex") %>%
    ggplot(aes(VAR, value)) +
    geom_polygon(
        group = 1, 
        color = "blue",
        linewidth = .2,
        fill = alpha("grey", .25)
    ) +
    geom_point(size = .25) +
    coord_straightpolar(theta = "x") +
    facet_grid(sex ~ Cluster) +
    theme_bw()
```

```{r}
pcprobs <- imap(
    pcdat,
    function(PCD, SEX){
        DAT <- PCD$x[,1:8]
        PCL <- pcclusparams[[SEX]]
        MUs <- map(PCL, "center")
        COVMATs <- map(PCL, "cov")
        W <- validclusmod$clusmod[
            validclusmod$sex == SEX
        ][[1]]$validclus_weights
        pdfs <- mapply(
            function(mu, covmat){ mvtnorm::dmvnorm(DAT, mu, covmat) },
            MUs, COVMATs
        )
        L <- pdfs %*% diag(W)
        colnames(L) <- names(PCL)
        tibble(data.frame(L / rowSums(L)))
    }
)
```

```{r}
pcprobs$Female %>% head
```

```{r}
pcprobs %>%
    map(
        ~{
            pp <- pmax(as.matrix(.x), .Machine$double.xmin)
            ent <- -sum(pp * log(pp))
            1 - ent / (nrow(pp) * log(ncol(pp)))
        }
    )
```

```{r}
gausmixres <- pcdat %>% 
    map(
        ~mclust::mclustBIC(data = .x$x[,1:8], G = 1:20, modelNames = "VVV")
    )
```

```{r}
options(repr.plot.height = 3, repr.plot.width = 7)
gausmixres %>%
    map(
        ~data.frame(
            Nclus = 1:20,
            BIC = .x[,1]
        )
    ) %>%
    bind_rows(.id = "sex") %>%
    ggplot(aes(Nclus, BIC)) +
    geom_line() +
    geom_point() +
    facet_wrap(~sex, scales = "free") +
    theme_bw()
```

```{r}
gmbest <- pcdat %>% 
    map(
        ~{
            mclustBIC <- mclust::mclustBIC
            mclust::Mclust(.x$x[,1:8], G = 4, modelNames = "VVV")
        }
    )
```

```{r}
gmbest %>%
    map(
        ~{
            pp <- pmax(as.matrix(.x$z), .Machine$double.xmin)
            ent <- -sum(rowSums(pp * log(pp)))
            1 - ent / (nrow(pp) * log(ncol(pp)))
        }
    ) %>%
    map(round, 2)
```

```{r}
gpairs_lower <- function(g){
  g$plots <- g$plots[-(1:g$nrow)]
  g$yAxisLabels <- g$yAxisLabels[-1]
  g$nrow <- g$nrow -1

  g$plots <- g$plots[-(seq(g$ncol, length(g$plots), by = g$ncol))]
  g$xAxisLabels <- g$xAxisLabels[-g$ncol]
  g$ncol <- g$ncol - 1

  g
}
```

```{r}
pcdat %>%
    map("x") %>%
    map_dbl(~max(abs(.x))) %>%
    max %>%
    round
```

```{r}
options(repr.plot.height = 1, repr.plot.width = 10)
cluscolp <- cluscolmap %>%
    ggplot(aes(Cluster, W2, color = Cluster)) +
    geom_point() +
    scale_color_discrete(
        guide = guide_legend(
            position = "top",
            direction = "horizontal",
            nrow = 1
        )
    ) +
    theme_bw()
cluscolguide <- GGally::grab_legend(cluscolp)
cluscolguide
```

```{r}
GPointsUpper <- function(data, mapping, ...){
    GGally::ggally_points(data, mapping, ...) +
    scale_color_manual(
            values = setNames(
                cluscolmap$ClusCol,
                cluscolmap$Cluster
            )
    ) +
    scale_x_continuous(
        limits = c(-10,10),
        breaks = c(-5, 0, 5)
    ) +
    scale_y_continuous(
        limits = c(-10,10),
        breaks = c(-5, 0, 5)
    )
}
GPointsLower <- function(data, mapping, ...){
    GGally::ggally_points(data, mapping, ...) +
    scale_x_continuous(
        limits = c(-10,10),
        breaks = c(-5, 0, 5)
    ) +
    scale_y_continuous(
        limits = c(-10,10),
        breaks = c(-5, 0, 5)
    )
}
GDiagFx <- function(data, mapping, ...){
    GGally::ggally_densityDiag(data, mapping, ...) +
    scale_x_continuous(
        limits = c(-10,10),
        breaks = c(-5, 0, 5)
    ) +
    theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()
    )
}
```

```{r}
g <- pcdat %>%
    map("x") %>%
    map(data.frame) %>%
    map2(maxprobdat, bind_cols) %>%
    map(mutate, Cluster = gsub("prob", "", Cluster)) %>%
    map(
        GGally::ggpairs,
        columns = 1:10,
        mapping = aes(color = Cluster),
        lower = list(
            continuous = GGally::wrap(
                GPointsUpper,
                shape = ".", alpha = .1
            )
        ),
        diag = list(
            continuous = GGally::wrap(
                GDiagFx,
                color = "black", fill = "grey",
                linewidth = .1
            )
        ),
        upper = list(
            continuous = GGally::wrap(
                GPointsLower, 
                color = "black", 
                shape = ".", alpha = .1
            )
        )
    ) %>%
    map(
        ~.x +
        theme(
            axis.text = element_text(size = 4),
            axis.ticks = element_line(linewidth = .1),
            strip.text = element_text(size = 6),
            panel.spacing = unit(0, "mm"),
            panel.border = element_rect(colour = "black", fill = NA),
            strip.background = element_rect(colour = "black", fill = NA)
        )
    ) %>%
    map(GGally::ggmatrix_gtable) %>%
    map(patchwork::wrap_elements) %>%
    patchwork::wrap_plots(nrow = 1)
```

```{r}
options(repr.plot.height = 5, repr.plot.width = 10)
patchwork::wrap_plots(
    cluscolguide, g,
    ncol = 1,
    heights = c(.01,.99)
)
```

```{r}
exdat_dhg <- pmap(
    list(
        strat_dat,
        residtab,
        probdfs
    ),
    function(ORIG, RESID, PROB){
        tibble(
            BMI = ORIG$bmi,
            FGres = RESID$fg,
            DHG = PROB$probDHG,
            BC = PROB$probBC
        )
    }
) %>%
    bind_rows(.id = "sex")
head(exdat_dhg)
```

```{r}
cluscolmap
```

```{r}
options(repr.plot.height = 5, repr.plot.width = 10)
dhgres <- exdat_dhg %>%
    ggplot(aes(BMI, FGres)) +
    geom_point(aes(color = DHG), size = .25) +
    scale_color_gradient(
        low = alpha(
            cluscolmap$ClusCol[
                cluscolmap$Cluster == "DHG"
            ], 
            0
        ),
        high = cluscolmap$ClusCol[
            cluscolmap$Cluster == "DHG"
        ]
    ) +
    ggdensity::geom_hdr() +
    geom_hline(yintercept = 0, lty = "dashed") +
    facet_grid(~sex) +
    theme_bw() +
    labs(
        x = NULL, y = "FG\nstandardised\nresiduals",
        color = "DHG\nprobability",
        alpha = "Overall\ndensity"
    ) +
    theme(
        strip.text.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank()
    )
dhgres
```

```{r}
wtd_quantile <- function(x, p, q){
    n <- sum(p)
    indexes <- order(x)
    x <- x[indexes]
    p <- p[indexes]
    p <- p / n
    cdfprobs <- cumsum(c(0, p))
    sapply(
        setNames(q, q), 
        function(QUANTILE) {
            Q <- pbeta(cdfprobs, n * QUANTILE, n * (1 - QUANTILE))
            W <- tail(Q, -1) - head(Q, -1)
            sum(W * x)
        }
    )
}
```

```{r}
options(repr.plot.height = 1, repr.plot.width = 10)
boxpld <- exdat_dhg %>%
    pivot_longer(
        c(DHG, BC), 
        names_to = "Cluster",
        values_to = "Prob"
    ) %>%
    nest(D = -c(sex, Cluster)) %>%
    mutate(
        D = map(
            D,
            ~wtd_quantile(
                .x$BMI, .x$Prob, 
                c(.025, .25, .5, .75, .975)
            )
        ),
        D = map(D, ~tibble(name = names(.x), value = .x))
    ) %>%
    unnest(D) %>%
    pivot_wider(names_prefix = "p")
boxpld
```

```{r}
options(repr.plot.height = 1, repr.plot.width = 10)
bx1 <- boxpld %>%
    filter(Cluster == "DHG") %>%
    ggplot(aes(y = "DHG")) +
    geom_boxplot(
        aes(
            xmin = p0.025,
            xlower = p0.25,
            xmiddle = p0.5,
            xupper = p0.75,
            xmax = p0.975
        ),
        stat = "identity",
        fill = cluscolmap$ClusCol[
            cluscolmap$Cluster == "DHG"
        ]
    ) +
    facet_grid(~sex) +
    coord_cartesian(
        xlim = c(
            min(exdat_dhg$BMI),
            max(exdat_dhg$BMI)
        )
    ) +
    labs(x = NULL, y = NULL) +
    theme_bw() +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(face = "bold"),
        plot.margin = margin()
    )
bx1
```

```{r}
options(repr.plot.height = 1, repr.plot.width = 10)
bx2 <- boxpld %>%
    filter(Cluster == "BC") %>%
    ggplot(aes(y = "BC")) +
    geom_boxplot(
        aes(
            xmin = p0.025,
            xlower = p0.25,
            xmiddle = p0.5,
            xupper = p0.75,
            xmax = p0.975
        ),
        stat = "identity",
        fill = cluscolmap$ClusCol[
            cluscolmap$Cluster == "BC"
        ]
    ) +
    facet_grid(~sex) +
    coord_cartesian(
        xlim = c(
            min(exdat_dhg$BMI),
            max(exdat_dhg$BMI)
        )
    ) +
    labs(x = "BMI (kg/m2)", y = NULL) +
    theme_bw() +
    theme(
        strip.text.x = element_blank(),
        axis.text.y = element_text(face = "bold"),
        plot.margin = margin()
    )
bx2
```

```{r}
options(repr.plot.height = 5, repr.plot.width = 10)
patchwork::wrap_plots(
    bx1,
    patchwork::plot_spacer(),
    dhgres +
    geom_vline(
        data = filter(boxpld, Cluster == "BC"),
        aes(xintercept = p0.5),
        color = cluscolmap$ClusCol[
            cluscolmap$Cluster == "BC"
        ]
    ) +
    geom_vline(
        data = filter(boxpld, Cluster == "DHG"),
        aes(xintercept = p0.5),
        color = cluscolmap$ClusCol[
            cluscolmap$Cluster == "DHG"
        ]
    ),
    patchwork::plot_spacer(),
    bx2,
    ncol = 1,
    heights = c(.05, -0.04, .9, -0.04, .05)
)
```

```{r}
options(repr.plot.height = 4, repr.plot.width = 15)
strat_dat %>%
    map(select, eid, all_of(BIOMARKERS)) %>%
    map(pivot_longer, -eid) %>%
    bind_rows(.id = "sex") %>%
    ggplot(aes(value)) +
    geom_density() +
    ggh4x::facet_grid2(sex ~ name, scales = "free", independent = "all") +
    theme_bw()
```

```{r}
options(repr.plot.height = 4, repr.plot.width = 15)
logtrdat <- strat_dat %>%
    map(mutate, across(.cols = c(alt, crp, tg, fg, scr), .fns = log))
map(logtrdat, head)
```

```{r}
logtrdat %>%
    map(select, eid, all_of(BIOMARKERS)) %>%
    map(pivot_longer, -eid) %>%
    bind_rows(.id = "sex") %>%
    ggplot(aes(value)) +
    geom_density() +
    ggh4x::facet_grid2(sex ~ name, scales = "free", independent = "all") +
    theme_bw()
```

```{r}
residlogtrdat <- logtrdat %>% 
    map(
        ~sapply(
            select(.x, all_of(BIOMARKERS)),
            function(X){
                lm(X ~ .x$age + .x$bmi + .x$smoking) %>%
                    resid %>%
                    scale %>%
                    as.vector
            }
        )
    )
map(residlogtrdat, head)
```

```{r}
logumap <- map(
    residlogtrdat,
    ~uwot::umap(
        X = .x,
        n_components = 2, 
        n_neighbors = max(10, round(10 + 15 * (log10(nrow(.x) - 4)))), 
        nn_method = "annoy", n_trees = 100, n_sgd_threads = "auto", 
        init = "pca", n_epochs = 500, approx_pow = TRUE,
        binary_edge_weights = TRUE, dens_scale = 1, 
        ret_extra = c("model", "fgraph"), verbose = FALSE       
    )
)
```

```{r}
options(repr.plot.res = 300, repr.plot.width = 7, repr.plot.height = 4)
logumap %>%
    map("embedding") %>%
    map(data.frame) %>%
    map2(maxprobdat, bind_cols) %>%
    bind_rows(.id = "sex") %>%
    mutate(Cluster = gsub("prob", "", Cluster)) %>%
    ggplot(aes(X1, X2)) +
    geom_point(aes(color = Cluster), alpha = .1, shape = ".") +
    scale_alpha_identity() +
    scale_color_manual(
        values = setNames(
            cluscolmap$ClusCol,
            cluscolmap$Cluster
        ),
        guide = guide_legend(
            override.aes = list(shape = 19, alpha = 1)
        )
    ) +     
    facet_wrap(~sex) +
    theme_bw() +
    labs(x = "UMAP1", y = "UMAP2")
```

```{r}
residtab %>%
    map(
        mutate,
        fav_adip = (
            whr < 0 &
            sbp < 0 &
            dbp < 0 &
            fg < 0 &
            tg < 0 &
            ldl < 0 &
            hdl > 0 &
            crp < 0 &
            alt < 0 &
            scr < 0
        )
    ) %>%
    map(count, fav_adip) %>%
    map(mutate, prop = 100 * n/sum(n)) %>%
    bind_rows(.id = "sex")
```

```{r}
residtab %>%
    map(
        mutate,
        fav_adip = (
            whr < 0 &
            sbp < 0 &
            dbp < 0 &
            fg < 0 &
            tg < 0 &
            ldl < 0 &
            hdl > 0 &
            crp < 0 &
            alt < 0 &
            scr < 0
        )
    ) %>%
    imap(
        ~.x %>%
            select(fav_adip) %>%
            bind_cols(
                data.frame(
                    validclusmod$umapmod[validclusmod$sex == .y][[1]]$embedding
                )
            )
    ) %>%
    bind_rows(.id = "sex") %>%
    ggplot(aes(X1, X2)) +
    geom_point(alpha = .1, shape = ".") +
    geom_point(data = . %>% filter(fav_adip), size = .25, alpha = .25, color = "red") + 
    facet_wrap(~sex) +
    theme_bw() +
    labs(x = "UMAP1", y = "UMAP2")
```

# Generating a table of biomarkers for different levels of discordant probabilities

```{r}
newclus <- validclusmod %>%
    transmute(
        sex, 
        clusmod = map(
            clusmod, select, 
            Cluster = validclus_name
        )
    ) %>%
    unnest(clusmod)
newclus
```

```{r}
newclus <- newclus %>%
    nest(D = -sex) %>%
    mutate(
        D = map(
            D, 
            expand_grid, 
            ClusProf = Cluster,
            ProbLevel = c(.2, .4, .6, .8)
        )
    ) %>%
    unnest(D) %>%
    filter(ClusProf != "BC") %>%
    mutate(
        Weight = (
            ((Cluster == ClusProf) * ProbLevel) +
            ((Cluster == "BC") * (1 - ProbLevel))
        )
    ) %>%
    filter(Weight > 0) %>%
    arrange(sex, ClusProf, ProbLevel)
head(newclus)
```

```{r}
newclusres <- newclus %>%
    nest(D = -c(sex, ClusProf, ProbLevel)) %>%
    inner_join(validclusmod, by = "sex") %>%
    mutate(
        sex,
        D = map2(
            D, clusmod, 
            inner_join, 
            by = c("Cluster" = "validclus_name")
        ),
        D = map(
            D,
            ~MGMM::rGMM(
                n = 100000, 
                d = length(.x$validclus_centers[[1]]),
                k = nrow(.x),
                pi = .x$Weight,
                means = .x$validclus_centers, 
                covs = .x$validclus_covmats
            )
        ),
        D = map(D, colMeans),
        D = map2(D, clusmod, ~setNames(.x, names(.y$validclus_centers[[1]])))
    )
```

```{r}
newclusres <- newclusres %>%
    mutate(
        D = map2(
            D, residmod,
            ~tibble(
                Biomarker = names(.x), sdresidval = .x
            ) %>%
            inner_join(.y, by = "Biomarker") %>%
            mutate(
                sdresid = sdresidval * SDRes,
                bmi = 30 * bmi,
                age = 55 * age,
                bmvalue = rowSums(pick(Intercept, bmi, age, sdresid)),
                bmvalue = round(bmvalue, 2)
            ) %>%
            select(Biomarker, bmvalue) %>%
            pivot_wider(names_from = Biomarker, values_from = bmvalue)
        )
    ) %>%
    select(sex, ClusProf, ProbLevel, D) %>%
    unnest(D)
head(newclusres)
```

```{r}
newclusres %>%
    mutate(
        across(c(sbp, dbp), round)
    )
```

### 2025-11-12 scripts/02_ValClus.Rmd
Initial clean commit

--- 2024-02-19 02_ValClus.Rmd
@@
                 "ukb" = "UKB",
-                "maastricht" = "MS",
+                "maastricht" = "TMS",
                 "rotterdam" = "RS",
@@
             "rotterdam" = "RS",
-            "maastricht" = "MS",
+            "maastricht" = "TMS",
             "ghs" = "GHS"
@@
                     option = "A", begin = .25, end = .75, direction = -1,
-                    labels = c(U = "UKB", M = "MS", R = "RS", G = "GHS"),
+                    labels = c(U = "UKB", M = "TMS", R = "RS", G = "GHS"),
                     guide = guide_legend(override.aes = list(size = 2))

### 2025-11-12 scripts/03_CentBased.Rmd
Initial clean commit

---
title: An R Markdown document converted from "03_CentBased.ipynb"
output: html_document
---

# Centroid-based clustering

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(ggplot2)
```

```{r}
options(repr.plot.res = 300)
```

## Loading data

```{r}
load("../data/validclusmod.RData")
```

```{r}
print(validclusmod)
```

```{r}
validclusmod <- validclusmod %>%
    mutate(
        umapmod = map(
            sex,
            ~uwot::load_uwot(paste0("../data/ukb/umapmod_", .x))
        )
    )
print(validclusmod)
```

```{r}
validclusmod <- validclusmod %>%
    mutate(
        residtab = map(
            umapmod,
            ~map(
                1:.x$nn_index$ann$getNItems(),
                function(i){
                    .x$nn_index$ann$getItemsVector(i-1)
                }
            )
        ),
        residtab = map(residtab, do.call, what = rbind)
    )
print(validclusmod)
```

## Centroid based clustering

```{r}
gausmixres <- validclusmod %>%
    transmute(
        sex,
        choosemod = map(residtab, mclust::mclustBIC, G = 1:20, modelNames = "VVV")
    )
```

```{r}
print(gausmixres)
```

```{r}
options(repr.plot.height = 3, repr.plot.width = 7)
gausmixres %>%
    mutate(
        choosemod = map(
            choosemod,
            ~data.frame(
                Nclus = 1:20,
                BIC = .x[,1]
            )
        )
    ) %>%
    unnest(choosemod) %>%
    ggplot(aes(Nclus, BIC)) +
    geom_line() +
    geom_point() +
    facet_wrap(~sex, scales = "free") +
    theme_bw()
```

```{r}
gmbest <- validclusmod %>%
    transmute(
        sex,
        mod = map(
            residtab, 
            ~{
                mclustBIC <- mclust::mclustBIC
                mclust::Mclust(.x, G = 6, modelNames = "VVV")
            }
        )
    )
```

```{r}
options(repr.plot.height = 4, repr.plot.width = 8)
gmbestcent <- gmbest %>%
    transmute(
        sex,
        meandat = map(mod, ~.x$parameters$mean)
    )
print(gmbestcent)
```

```{r}
options(repr.plot.height = 4, repr.plot.width = 8)
gmbestcent %>%
    mutate(
        meandat = map(
            meandat,
            ~{
                res <- t(.x)
                colnames(res) <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
                data.frame(Cluster = paste0("Cluster", 1:nrow(res)), res)
            }
        )
    ) %>%
    unnest(meandat) %>%
    pivot_longer(-c(sex, Cluster), names_to = "trait", values_to = "residval") %>%
    mutate(trait = toupper(trait)) %>%
    ggplot(aes(residval, trait)) +
    geom_segment(aes(xend = 0, yend = trait)) +
    geom_vline(xintercept = 0, lty = "dashed") +
    geom_point() +
    facet_grid(sex ~ Cluster) +
    labs(x = "Residuals - SD units", y = NULL) +
    theme_bw()
```

```{r}
options(repr.plot.height = 2.5, repr.plot.width = 4.5)
gmbest %>%
    transmute(
        sex,
        weightdat = map(
            mod,
            ~data.frame(Cluster = paste0("Cluster", 1:.x$G), 
                        Weight = .x$parameters$pro)
        )
    ) %>%
    unnest(weightdat) %>%
    ggplot(aes(Weight, reorder(Cluster, desc(Cluster)))) +
    geom_col() +
    facet_wrap(~sex) +
    scale_x_continuous(labels = \(x) 100 * x) +
    theme_bw() +
    labs(x = "Weight (%)", y = NULL)
```

```{r}
gmbest %>%
    transmute(
        sex,
        RelEnt = map_dbl(
            mod,
            ~{
                pp <- pmax(as.matrix(.x$z), .Machine$double.xmin)
                ent <- -sum(rowSums(pp * log(pp)))
                1 - ent / (nrow(pp) * log(ncol(pp)))
            }
        )
    )
```

```{r}
embeddf <- validclusmod %>%
    transmute(sex, umapmod = map(umapmod, ~data.frame(.x$embedding))) %>%
    unnest(umapmod)
head(embeddf)
```

```{r}
centembed <- validclusmod %>%
    inner_join(gmbestcent, by = "sex") %>%
    transmute(
        sex,
        meandat = map2(
            meandat, umapmod,
            ~data.frame(uwot::umap_transform(t(.x), .y))
        )
    ) %>%
    unnest(meandat)
head(centembed)
```

```{r}
options(repr.plot.width = 5, repr.plot.height = 3)
embeddf %>%
    ggplot(aes(X1, X2)) +
    geom_point(shape = ".", alpha = .1) +
    geom_point(data = centembed, color = "red") +
    facet_wrap(~sex) +
    theme_bw() +
    labs(x = "UMAP1", y = "UMAP2")
```

### 2025-11-12 scripts/08_PrevDx.Rmd
Initial clean commit

--- 2024-02-19 08_PrevDx.Rmd
@@
 head(bm_meansd)
+```
+
+```{r}
+bmdiff <- bm_meansd %>%
+    group_by(cohort, sex, Variable) %>%
+    mutate(
+        across(
+            c(N, Mean, SD),
+            ~.x[Cluster == "BC"],
+            .names = "{.col}_BC"
+        )
+    ) %>%
+    ungroup %>%
+    filter(Cluster != "BC") %>%
+    mutate(
+        estimate = Mean - Mean_BC,
+        se = sqrt(
+            (SD^2/N) + 
+            (SD_BC^2/N_BC)
+        ),
+        pval = 2*pnorm(-abs(estimate/se))
+    )
+head(bmdiff)
+```
+
+```{r}
+pooledbmdiff <- bmdiff %>%
+    nest(D = -c(sex, Variable, Cluster)) %>%
+    mutate(
+        D = map(
+            D,
+            ~meta::metagen(TE = .x$estimate, seTE = .x$se)
+        ),
+        D = map(
+            D,
+            ~tibble(
+                cohort = c("Pooled_FE", "Pooled_RE"),
+                estimate = c(.x$TE.fixed, .x$TE.random), 
+                se = c(.x$seTE.fixed, .x$seTE.random)
+            )
+        )
+    ) %>%
+    unnest(D) %>%
+    mutate(
+        pval = 2*pnorm(-abs(estimate/se))
+    )
+head(pooledbmdiff)
+```
+
+```{r}
+smkdiff <- result_crosssect %>%
+    filter(element_name == "MarkerDistrib") %>%
+    unnest(element_value) %>%
+    filter(Variable == "smoking", Summary1 == 1) %>%
+    transmute(
+        sex, Variable, 
+        cohort = stringr::str_replace_all(
+            cohort,
+            c(
+                "ukb" = "UKB",
+                "maastricht" = "TMS",
+                "rotterdam" = "RS",
+                "ghs" = "GHS"
+            )
+        ),
+        Cluster = factor(Cluster, levels = cluscolmap$Cluster),
+        Smk = ceiling(as.numeric(Summary2)), 
+        NSmk = ceiling(N - as.numeric(Summary2))
+    ) %>%
+    nest(D = -c(sex, cohort, Variable)) %>%
+    mutate(
+        D = map(
+            D,
+            ~glm(cbind(Smk, NSmk) ~ Cluster, data = .x, family = binomial)
+        ),
+        D = map(D, broom::tidy),
+        D = map(
+            D, transmute,
+            Cluster = gsub("Cluster", "", term),
+            estimate, se = std.error, pval = p.value
+        )
+    ) %>%
+    unnest(D) %>%
+    filter(Cluster != "(Intercept)")
+head(smkdiff)
+```
+
+```{r}
+pooledsmkdiff <- smkdiff %>%
+    nest(D = -c(sex, Variable, Cluster)) %>%
+    mutate(
+        D = map(
+            D,
+            ~meta::metagen(TE = .x$estimate, seTE = .x$se)
+        ),
+        D = map(
+            D,
+            ~tibble(
+                cohort = c("Pooled_FE", "Pooled_RE"),
+                estimate = c(.x$TE.fixed, .x$TE.random), 
+                se = c(.x$seTE.fixed, .x$seTE.random)
+            )
+        )
+    ) %>%
+    unnest(D) %>%
+    mutate(
+        pval = 2*pnorm(-abs(estimate/se))
+    )
+head(pooledsmkdiff)
+```
+
+```{r}
+alldiffs <- list(
+    bmdiff, pooledbmdiff,
+    smkdiff, pooledsmkdiff
+) %>%
+    map(select, c(cohort, sex, Variable, Cluster, estimate, se, pval)) %>%
+    bind_rows %>%
+    nest(D = -c(sex, Variable, Cluster)) %>%
+    mutate(
+        pvalfdr = map_dbl(D, ~.x$pval[.x$cohort == "Pooled_RE"]),
+        pvalfdr = p.adjust(pvalfdr, "fdr") < 0.05
+    ) %>%
+    unnest(D)
+head(alldiffs)
+```
+
+```{r}
+options(repr.plot.width = 8, repr.plot.height = 6)
+alldiffs %>%
+    mutate(
+        ci = qnorm(1 - .05/2) * se,
+        lwr = estimate - ci,
+        upr = estimate + ci,
+        across(
+            c(estimate, upr, lwr),
+            ~ifelse(Variable == "smoking", exp(.x), .x)
+        ),
+        cohort = factor(
+            cohort,
+            levels = rev(c(
+                "UKB", "RS", "TMS", "GHS",
+                "Pooled_FE", "Pooled_RE"
+            ))
+        ),
+        Variable = toupper(Variable),
+        VarClass = ifelse(
+            Variable %in% c("AGE", "BMI", "SMOKING"),
+            "Covariates",
+            "Biomarkers"
+        ),
+        Units = case_match(
+            Variable,
+            "AGE" ~ "years",
+            "BMI" ~ "kg/m2",
+            "SMOKING" ~ "OR",
+            c("SBP", "DBP") ~ "mmHg",
+            c("HDL", "LDL", "TG", "FG") ~ "mmol/L",
+            "ALT" ~ "U/L",
+            "WHR" ~ "cm/cm",
+            "SCR" ~ "umol/L",
+            "CRP" ~ "mg/L"
+        ),
+        Variable = paste(Variable, Units, sep = "\n"),
+        Cluster = factor(Cluster, levels = cluscolmap$Cluster)
+    ) %>%
+    ggplot(aes(estimate, y = cohort)) +
+    geom_vline(
+        aes(
+            xintercept = ifelse(Variable == "SMOKING\nOR", 1, 0)
+        ),
+        lty = "dashed", color = "red"
+    ) +
+    geom_linerange(aes(xmin = lwr, xmax = upr)) +
+    geom_point(
+        aes(shape = grepl("Pooled", cohort),
+            fill = cohort == "Pooled_RE" & pvalfdr),
+        stroke = 0
+    ) +
+    scale_fill_manual(
+        name = "Significant after 5% FDR correction",
+        values = c("black", "red"),
+        guide = guide_legend(
+            position = "top",
+            override.aes = list(shape = 21)
+        )
+    ) +
+    scale_shape_manual(
+        values = c(19, 23),
+        guide = guide_none()
+    ) +
+    ggh4x::facet_nested(
+        sex + Cluster ~ VarClass + Variable, 
+        scales = "free",
+        space = "free_y",
+        strip = ggh4x::strip_nested(
+            text_y = ggh4x::elem_list_text(angle = 0), 
+            size = "variable",
+            background_y = ggh4x::elem_list_rect(
+                fill = alpha(
+                    c(
+                        "grey", "grey",
+                        cluscolmap$ClusCol[
+                            cluscolmap$Cluster != "BC"
+                        ],
+                        cluscolmap$ClusCol[
+                            !cluscolmap$Cluster %in% c("BC", "DHT")
+                        ]
+                    ),
+                    .8
+                )
+            )
+        )
+    ) +
+    ggh4x::scale_x_facet(
+        Variable == "SMOKING\nOR", 
+        limits = c(0, 2)
+    ) +
+    labs(x = NULL, y = NULL) + 
+    theme_bw() +
+    theme(
+        axis.text.x = element_text(size = 4),
+        axis.text.y = element_text(size = 5),
+        axis.title.x = element_text(size = 6),
+        strip.text = element_text(size = 6)
+    )
+```
+
+```{r}
+alldiffs %>%
+    mutate(
+        ci = qnorm(1 - .05/2) * se,
+        lwr = estimate - ci,
+        upr = estimate + ci
+    ) %>%
+    filter(
+        cohort == "Pooled_RE", Variable == "smoking", Cluster == "DIS"
+    ) %>%
+    transmute(
+        sex,
+        across(c(estimate, lwr, upr), exp),
+        estimate = paste0(
+            round(estimate, 2),
+            ", 95% CI: ",
+            round(lwr, 2),
+            ", ",
+            round(upr, 2)
+        )
+    )
 ```
@@
     )
+```
+
+## Prevalence of diseases by cohort
+
+```{r}
+result_crosssect %>%
+    filter(element_name == "CountCovars") %>%
+    unnest(element_value) %>%
+    filter(
+        Covariate %in% c(
+            "CHD", "Stroke", "T2D", "HT", "T1D", "LiverFailure", "RA"
+        )
+    ) %>%
+    group_by(cohort, sex, Covariate) %>%
+    summarise(
+        across(c(Ncases, Nnoncases), sum),
+        .groups = "drop"
+    ) %>%
+    mutate(Prop = round(100 * Ncases/(Ncases+Nnoncases), 2)) %>%
+    select(-c(Ncases, Nnoncases)) %>%
+    pivot_wider(names_from = cohort, values_from = Prop)
 ```
@@
 ```{r}
+allORs <- read_tsv("../data/allORs.tsv", show_col_types = FALSE)
+```
+
+```{r}
 options(repr.plot.width = 10, repr.plot.height = 3)
@@
 ```{r}
+allORmeds <- read_tsv("../data/allORmeds.tsv", show_col_types = FALSE)
+```
+
+```{r}
 options(repr.plot.width = 10, repr.plot.height = 3)
@@
 options(repr.plot.width = 5, repr.plot.height = 3)
-allORmeds %>%
+orplot2 <- allORmeds %>%
     drop_na %>%
@@
         bigeff = OR > 10, xint = ifelse(bigeff, NA, 1),
-        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster))
+        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster)),
+        Covariate = factor(Covariate, levels = c("Insulin", "AntiHT", "LipidLower"))
     ) %>%
@@
         axis.text.x = element_text(size = 5)
+    )
+orplot2
+```
+
+## OR of prevalent diseases adjusted by medication
+
+```{r}
+MedAdjDat <- result_crosssect %>%
+    filter(element_name %in% c("CountCovars", "CountDXMeds")) %>%
+    pivot_wider(names_from = element_name, values_from = element_value) %>%
+    mutate(
+        CountDXMeds = map(CountDXMeds, filter, Med == "NoMed"),
+        CountDXMeds = map(CountDXMeds, rename, Covariate = Dx),
+        CountDXMeds = map(CountDXMeds, select, -Med),
+        across(
+            everything(),
+            ~map(
+                .x, 
+                pivot_longer, 
+                cols = c(Ncases, Nnoncases), 
+                names_to = "DxStatus", values_to = "N"
+            )
+        ),
+        CountDXMeds = map(CountDXMeds, rename, NoMed = N),
+        CountD = map2(CountCovars, CountDXMeds, inner_join)
+    ) %>%
+    select(CountD) %>%
+    unnest(CountD) %>%
+    mutate(Med = N - NoMed) %>%
+    select(-c(Nclus, N)) %>%
+    pivot_longer(
+        c(Med, NoMed), 
+        names_to = "MedStatus", 
+        values_to = "NMed"
+    ) %>%
+    pivot_wider(names_from = DxStatus, values_from = NMed)
+head(MedAdjDat)
+```
+
+```{r}
+adjORs <- MedAdjDat %>%
+    filter(
+        Covariate %in% c(
+            "CHD", "Stroke", "T2D", "HT", "T1D", "LiverFailure", "RA"
+        )
+    ) %>%
+    mutate(
+        MedStatus = relevel(factor(MedStatus), ref = "NoMed"),
+        across(c(Ncases, Nnoncases), round)
+    ) %>%
+    nest(D = -c(sex, Covariate)) %>%
+    mutate(
+        D = map(D, drop_na),
+        D = map(
+            D,
+            ~if(length(unique(.x$cohort)) > 1){
+                glmer(
+                    cbind(Ncases, Nnoncases) ~ Cluster + MedStatus + (1|cohort),
+                    data = .x, family = binomial
+                )
+            } else {
+                glm(
+                    cbind(Ncases, Nnoncases) ~ Cluster + MedStatus,
+                    data = .x, family = binomial
+                )
+            }
+        ),
+        D = map(
+            D, 
+            ~if("glmerMod" %in% class(.x)) {
+                broom.mixed::tidy(.x, effects = "fixed")
+            } else {
+                broom::tidy(.x)
+            }
+        ),
+        D = map(D, select, term, estimate, std.error)
+    ) %>%
+    unnest(D)
+```
+
+```{r}
+head(adjORs)
+```
+
+```{r}
+adjORs %>%
+    filter(Covariate == "RA")
+```
+
+```{r}
+allORs %>%
+    filter(Covariate == "RA")
+```
+
+```{r}
+allORwithadj <- bind_rows(
+    Unadjusted = allORs %>%
+        filter(
+            Cluster != "BC",
+            (cohort == "Pooled_RE" | 
+             (Covariate == "RA" & cohort == "UKB"))
+        ) %>%
+        transmute(
+            sex, Covariate, Cluster,
+            OR, LOWERCI, UPPERCI
+        ),
+    Adjusted = adjORs %>%
+        filter(grepl("Clusterprob", term)) %>%
+        transmute(
+            sex, Covariate,
+            Cluster = gsub("Clusterprob", "", term),
+            OR = exp(estimate),
+            LOWERCI = exp(estimate - qnorm(1 - 0.05/2) * std.error),
+            UPPERCI = exp(estimate + qnorm(1 - 0.05/2) * std.error)
+    ),
+    .id = "Model"
+)
+head(allORwithadj)
+```
+
+```{r}
+write_tsv(allORwithadj, "../data/allORwithadj.tsv")
+```
+
+```{r}
+options(repr.plot.width = 10, repr.plot.height = 3)
+orplot1 <- allORwithadj %>%
+    mutate(
+        Covariate = factor(
+            Covariate,
+            levels = c(
+                "CHD", "Stroke", "T2D", "HT", "T1D", "LiverFailure", "RA"
+            )
+        ),
+        Cluster = factor(
+            Cluster, 
+            levels = rev(cluscolmap$Cluster)
+        ),
+        bigeff = OR > 10, xint = ifelse(bigeff, NA, 1)
+    ) %>%
+    ggplot(aes(OR, Cluster)) +
+    geom_vline(aes(xintercept = xint), lty = "dashed", na.rm = TRUE) +
+    geom_linerange(
+        aes(xmin = LOWERCI, xmax = UPPERCI, group = Model),
+        position = position_dodge(width = .5), linewidth = .5
+    ) +
+    geom_point(
+        aes(color = Model, group = Model), size = .75,
+        position = position_dodge(width = .5)
+    ) +
+    scale_color_manual(
+        values = c("red", "black")
+    ) +
+    ggh4x::facet_nested(
+        sex ~ Covariate + bigeff, scales = "free", space = "free_y",
+        strip = ggh4x::strip_nested(
+            size = "variable",
+            background_x = list(element_rect(), element_rect(fill = NA, colour = NA)),
+            text_x = list(element_text(), element_text(size = 0, margin = margin())),
+            by_layer_x = TRUE
+        )
+    ) +
+    ggh4x::force_panelsizes(cols = c(1, 1, 1, .5, 1, 1, .5, 1, 1)) +
+    ggh4x::scale_x_facet(!bigeff, n.breaks = 4, limits = c(0, NA)) +
+    ggh4x::scale_x_facet(bigeff, n.breaks = 4) +
+    labs(x = "OR relative to BC", y = NULL) +
+    theme_bw() +
+    theme(
+        axis.text.x = element_text(size = 5),
+        legend.position = "top"
+    )
+orplot1
+```
+
+```{r}
+dxfreeprop <- readRDS("../data/dxfreeprop.rds")
+dxfreeprop
+```
+
+```{r}
+options(repr.plot.width = 8, repr.plot.height = 6)
+patchwork::wrap_plots(
+    patchwork::wrap_elements(full = orplot1),
+    orplot2, 
+    dxfreeprop + 
+        theme(
+            legend.text = element_text(size = 6),
+            legend.margin = margin()
+        ), 
+    design = "AA\nBC",
+    heights = c(.65, .35)
+) +
+    patchwork::plot_annotation(
+        tag_level = "A"
     )

### 2025-11-12 scripts/12_EthnicAssoc.Rmd
Initial clean commit

---
title: An R Markdown document converted from "12_EthnicAssoc.ipynb"
output: html_document
---

# Analysis in other ethnicities

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(ggplot2)
library(Matrix, warn.conflicts = FALSE)
library(uwot)
library(survival)
options(repr.plot.res = 300)
```

## Input data

```{r}
recoded_ethn <- read_tsv("../data/recoded_ethn.tsv", show_col_types = FALSE)
```

```{r}
head(recoded_ethn)
```

```{r}
strat_dat <- recoded_ethn %>%
    nest(Data = -c(Ethnicity, sex))
print(strat_dat)
```

```{r}
BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
```

```{r}
strat_dat <- strat_dat %>%
    mutate(
        Data = map(
            Data,
            function(DAT){
                DAT %>%
                    mutate(
                        across(
                            all_of(c("bmi", BIOMARKERS)),
                            function(COL){
                                Mean <- mean(COL)
                                SD <- sd(COL)
                                upr <- Mean + (5*SD)
                                lwr <- Mean - (5*SD)
                                COL[COL < lwr] <- NA
                                COL[COL > upr] <- NA
                                return(COL)
                            }
                        )
                    ) %>%
                    drop_na
            }
        )
    )
```

```{r}
print(strat_dat)
```

```{r}
strat_dat %>%
    group_by(Ethnicity) %>%
    summarise(Total = sum(map_dbl(Data, nrow)))
```

## Cluster allocation probabilities and UMAP models

```{r}
load("../data/validclusmod.RData")
```

```{r}
print(validclusmod)
```

```{r}
validclusmod <- validclusmod %>%
    mutate(
        umapmod = map(
            sex, 
            ~load_uwot(paste0("../data/ukb/umapmod_", .x))
        )
    )
print(validclusmod)
```

```{r}
df <- inner_join(validclusmod, strat_dat, by = "sex")
print(df)
```

## Calculation of probabilities

```{r}
df <- df %>%
    mutate(
        resids = map2(
            Data, residmod,
            function(DATA, RESIDMOD){
                sapply(
                    BIOMARKERS,
                    function(biomarker){
                        PREDICTDAT <- as.matrix(cbind(Intercept = 1, DATA[,c("bmi", "age", "smoking")]))
                        MODELDAT <- dplyr::filter(RESIDMOD, Biomarker == biomarker)
                        RESIDSD <- MODELDAT$SDRes
                        COEFDAT <- MODELDAT[,c("Intercept", "bmi", "age", "smoking")]
                        COEFDAT <- t(COEFDAT)
                        PREDVAL <- as.vector(PREDICTDAT %*% COEFDAT)
                        RESIDVAL <- DATA[[biomarker]] - PREDVAL
                        RESIDVAL / RESIDSD
                    }
                )
            }
        )
    )
print(df)
```

```{r}
df <- df %>%
    mutate(
        probs = map2(
            resids, clusmod,
            function(RESIDDF, CLUSMOD){
                mus <- CLUSMOD$validclus_centers
                mus <- map(mus, function(mu){ mu[BIOMARKERS] })
                covmats <- CLUSMOD$validclus_covmats
                covmats <- map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
                wts <- CLUSMOD$validclus_weights
                pdfs <- mapply(
                    function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) },
                    mus, covmats
                )
                L <- pdfs %*% diag(wts)
                colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
                tibble(data.frame(L / rowSums(L)))
            }
        )
    )
print(df)
```

```{r}
head(df$probs[[1]])
```

```{r}
cluscolmap <- validclusmod %>%
    transmute(
        clusmod = map(clusmod, select, Cluster = validclus_name)
    ) %>%
    unnest(clusmod) %>%
    unique %>%
    mutate(ClusCol = scales::hue_pal()(n()))
cluscolmap
```

```{r}
PercDat <- df %>%
    transmute(
        sex, Ethnicity,
        probs = map(probs, colSums),
        probs = map(
            probs, 
            ~tibble(
                Cluster = gsub("prob", "", names(.x)),
                N = .x,
                Perc = .x/sum(.x)
            )
        )
    ) %>%
    unnest(probs)
head(PercDat)
```

```{r}
PercDatRef <- validclusmod %>%
    transmute(
        sex, 
        Ethnicity = "European",
        clusmod = map(
            clusmod, select, 
            Cluster = validclus_name, 
            Perc = validclus_weights
        )
    ) %>%
    unnest(clusmod)
head(PercDatRef)
```

```{r}
ClusOrd <- PercDatRef %>%
    group_by(sex) %>%
    arrange(desc(Perc), .by_group = TRUE) %>%
    mutate(ordn = row_number()) %>%
    ungroup %>%
    select(sex, Cluster, ordn)
ClusOrd
```

```{r}
options(repr.plot.height = 4, repr.plot.width = 12)
PercDat %>%
    select(-N) %>%
    bind_rows(PercDatRef) %>%
    inner_join(cluscolmap, by = "Cluster") %>%
    mutate(
        Perc2 = ifelse(Perc > 0.125, Perc, NA)
    ) %>%
    pivot_longer(c(Perc, Perc2), names_to = "PercLevel", values_to = "PercValue") %>%
    mutate(
        Base = case_when(
            PercLevel == "Perc" ~ 0,
            PercLevel == "Perc2" & !is.na(PercValue) ~ .6
        ),
        PercValue = ifelse(PercLevel == "Perc" & PercValue > .125, .125, PercValue),
        PercLevel = factor(PercLevel, levels = c("Perc2", "Perc")),
        across(c(Base, PercValue), \(x) 100 * x),
        Ethnicity = factor(Ethnicity, levels = c("European", "African", "SouthAsian"))
    ) %>%
    inner_join(ClusOrd, by = c("sex", "Cluster")) %>%
    ggplot(aes(reorder(Cluster, ordn), PercValue)) +
    geom_rect(
        aes(
            xmin = Cluster, xmax = Cluster,
            ymin = Base, ymax = PercValue,
            color = ClusCol
        ),
        linewidth = 8, na.rm = TRUE
    ) +
    scale_color_identity() +
    ggh4x::facet_nested(
        PercLevel ~ sex + Ethnicity,
        scales = "free", space = "free_x"
    ) +
    ggh4x::force_panelsizes(
        rows = c(1, 3)
    ) +
    ggh4x::scale_y_facet(
        PercLevel == "Perc",
        breaks = c(0, 5, 10),
        expand = c(.05, 0, 0, 0) 
    ) +
    ggh4x::scale_y_facet(
        PercLevel == "Perc2",
        breaks = c(60,80),
        limits = c(60,90),
        expand = expansion(mult = c(0, NA))
    ) +
    theme_bw() +
    theme(strip.text.y = element_blank()) +
    labs(x = NULL, y = "%")
```

## Comparing proportions

```{r}
RefDat <- validclusmod %>%
    transmute(
        sex, 
        residtab = map(
            umapmod,
            ~map(
                1:.x$nn_index$ann$getNItems(),
                function(i){
                    .x$nn_index$ann$getItemsVector(i-1)
                }
            )
        ),
        resids = map(residtab, do.call, what = rbind),
        probs = map2(
            resids, clusmod,
            function(RESIDDF, CLUSMOD){
                pdfs <- mapply(
                    function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) },
                    CLUSMOD$validclus_centers, 
                    CLUSMOD$validclus_covmats
                )
                L <- pdfs %*% diag(CLUSMOD$validclus_weights)
                colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
                tibble(data.frame(L / rowSums(L)))
            }
        )
    )
print(RefDat)
```

```{r}
RefN <- RefDat %>%
    transmute(
        sex, Ethnicity = "European",
        probs = map(probs, colSums),
        probs = map(
            probs, 
            ~tibble(
                Cluster = gsub("prob", "", names(.x)),
                N = .x, Total = sum(.x)
            )
        )
    ) %>%
    unnest(probs)
RefN
```

```{r}
RefNList <- nest(RefN, RefD = -c(sex, Cluster))
print(RefNList)
```

```{r}
BinomDat <- PercDat %>%
    select(-Perc) %>%
    group_by(sex, Ethnicity) %>%
    mutate(Total = sum(N)) %>%
    ungroup %>%
    nest(Data = -c(sex, Cluster, Ethnicity)) %>%
    inner_join(RefNList, by = join_by(sex, Cluster)) %>%
    mutate(
        Data = map2(Data, Ethnicity, ~mutate(.x, Ethnicity = .y)),
        Data = map2(RefD, Data, bind_rows),
        RefD = NULL,
        Data = map(
            Data, mutate, 
            Nnot = Total - N,
            Ethnicity = relevel(factor(Ethnicity), ref = "European")
        )
    )
print(BinomDat)
```

```{r}
BinomDat$Data[[1]]
```

```{r}
BinomDat <- BinomDat %>%
    mutate(
        Mod = map(Data, ~glm(cbind(round(N), round(Nnot)) ~ Ethnicity, data = .x, family = binomial))
    )
print(BinomDat)
```

```{r}
BinomRes <- BinomDat %>%
    transmute(
        sex, Cluster, Ethnicity,
        Data = map(Data, filter, Ethnicity != "European"),
        Data = map(Data, select, -Ethnicity),
        Mod = map(Mod, broom::tidy, exponentiate = TRUE, conf.int = TRUE),
        Mod = map(Mod, filter, term != "(Intercept)"),
        Mod = map(Mod, select, -c(term, std.error, statistic))
    ) %>%
    unnest(c(Data, Mod)) %>%
    mutate(
        padj = p.adjust(p.value, "fdr")
    )
BinomRes
```

```{r}
write_tsv(BinomRes, "../data/BinomRes_ethn.tsv")
```

```{r}
SigBinomRes <- BinomRes %>% 
    filter(padj < 0.05)
SigBinomRes
```

```{r}
SigUpBinomRes <- SigBinomRes %>% 
    filter(estimate > 1) %>%
    mutate(across(c(estimate, conf.low, conf.high), ~round(.x, digits = 2))) %>%
    arrange(Ethnicity, sex, desc(estimate))
SigUpBinomRes
```

## UMAP projection

Now we can plot the deviations from BMI of the new data points on top of the UK biobank projection:

```{r}
newcoords <- df %>%
    transmute(
        sex, Ethnicity,
        umapg = map2(
            resids, umapmod, 
            function(newx, mod){
                data.frame(umap_transform(newx, mod))
            }
        )
    ) %>%
    unnest(umapg)
head(newcoords)
```

```{r}
basecoords <- df %>%
    transmute(
        sex, Ethnicity,
        umapg = map(umapmod, ~data.frame(.x$embedding))
    ) %>%
    unnest(umapg)
head(basecoords)
```

```{r}
options(repr.plot.height = 5, repr.plot.width = 5)
embed <- basecoords %>%
    ggplot(aes(X1, X2)) +
    geom_point(shape = ".", alpha = .1) +
    geom_point(data = data.frame(newcoords),
               color = "red", shape = ".") +
    facet_grid(Ethnicity~sex) +
    theme_bw() +
    labs(x = "UMAP1", y = "UMAP2")
embed
```

## Covariate data

```{r}
covar_dat <- readr::read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

## Prevalence

```{r}
df <- df %>%
    mutate(
        # Joining complete biomarker data with allocation probabilities
        joindat = map2(Data, probs, bind_cols),
        # Adding covariate data
        joindat = map(joindat, inner_join, covar_dat, by = "eid")
    )
print(df)
```

```{r}
prevdat <- df %>%
    transmute(
        sex, Ethnicity,
        propdat = map(
            joindat,
            function(DAT){
                tibble(
                    Cluster = names(DAT)[grepl("^prob", names(DAT))]
                ) %>%
                    mutate(
                        clusprobdat = map(
                            Cluster,
                            function(CL){
                                map(
                                    c("HT", "CHD", "Stroke", "LiverFailure", "RA", "T2D", "T1D", 
                                      "Insulin", "AntiDM", "AntiHT", "LipidLower"),
                                    function(DX){
                                        tibble(
                                            Disease = DX,
                                            N = sum(DAT[[CL]]),
                                            Ncases = sum(DAT[[DX]] * DAT[[CL]])
                                        )
                                    }
                                ) %>%
                                    bind_rows
                            }
                        )
                    ) %>%
                    unnest(clusprobdat)
            }
        )
    ) %>%
    unnest(propdat) %>%
    mutate(Cluster = gsub("prob", "", Cluster))
```

```{r}
head(prevdat)
```

### Odds ratios relative to the concordant cluster

```{r}
ORdat <- prevdat %>%
    nest(Data = -c(sex, Ethnicity, Disease)) %>%
    mutate(
        Data = map(
            Data,
            mutate,
            Cluster = factor(Cluster),
            Cluster = relevel(Cluster, ref = "BC")
        ),
        mod = map(
            Data,
            ~glm(cbind(round(Ncases), round(N - Ncases)) ~ Cluster, data = .x, family = binomial)
        ),
        modres = map(mod, broom::tidy),
        modres = map(modres, filter, term != "(Intercept)"),
        modres = map(modres, mutate, Cluster = gsub("Cluster", "", term), term = NULL),
        modres = map2(Data, modres, left_join, join_by(Cluster))
    ) %>%
    select(sex, Ethnicity, Disease, modres) %>%
    unnest(modres) %>%
    mutate(
        OR = exp(estimate),
        lwr = exp(estimate - qnorm(1 - .05/2) * std.error),
        upr = exp(estimate + qnorm(1 - .05/2) * std.error)
    ) %>%
    select(-c(estimate, std.error, statistic))
head(ORdat)
```

```{r}
DiseaseDat <- ORdat %>%
    filter(Disease %in% c("HT", "CHD", "Stroke", "LiverFailure", "RA", "T2D", "T1D")) %>%
    mutate(padj = p.adjust(p.value, "fdr"))
head(DiseaseDat)
```

```{r}
write_tsv(DiseaseDat, "../data/DiseaseDat_ethn.tsv")
```

```{r}
DiseaseDat %>%
    filter(padj < 0.01)
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 5)
DiseaseDat %>%
    drop_na %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(aes(xmin = lwr, xmax = upr)) +
    geom_point() +
    ggh4x::facet_nested(sex + Ethnicity ~ Disease, scales = "free", space = "free_y") +
    labs(x = "OR relative to BC", y = NULL, color = "Model") +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5)
    )
```

```{r}
MedDat <- ORdat %>%
    filter(Disease %in% c("Insulin", "AntiHT", "LipidLower")) %>%
    mutate(padj = p.adjust(p.value, "fdr"))
head(MedDat)
```

```{r}
write_tsv(MedDat, "../data/MedDat_ethn.tsv")
```

```{r}
MedDat %>%
    filter(padj < 0.01)
```

```{r}
options(repr.plot.width = 6, repr.plot.height = 5)
ORdat %>%
    drop_na %>%
    filter(Disease %in% c("Insulin", "AntiHT", "LipidLower")) %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(aes(xmin = lwr, xmax = upr)) +
    geom_point() +
    ggh4x::facet_nested(sex + Ethnicity ~ Disease, scales = "free", space = "free_y") +
    labs(x = "OR relative to BC", y = NULL, color = "Model") +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5)
    )
```

## Incidence

```{r}
df <- df %>%
    expand_grid(
        Outcome = c("mace", "dm")
    )
print(df)
```

```{r}
df <- df %>%
    mutate(
        SurvDat = map(
            Outcome,
            ~readr::read_tsv(paste0("../data/surv", .x, "dat_ethn.tsv"), show_col_types = FALSE)
        ),
        SurvDat = map2(
            SurvDat, joindat, inner_join, by = "eid"
        ),
        SurvDat = map2(
            Outcome, SurvDat,
            ~switch(
                .x,
                mace = filter(.y, CHD == 0, Stroke == 0, PAD == 0),
                dm = filter(.y, T2D == 0, T1D == 0, Insulin == 0, AntiDM == 0)
            )
        ),
        SurvDat = map(SurvDat, select, -where(\(x) all(x == 0)))
    )
print(df)
```

## Data for modelling

```{r}
egfrcalc <- function(scr, sex, age){
    scr <- scr / 88.42
    alpha <- ((sex == "Female") * -0.241) + ((sex == "Male") * -0.302)
    kappa <- ((sex == "Female") * 0.7) + ((sex == "Male") * 0.9)
    creat_kappa <- scr / kappa
    minkappa <- pmin(creat_kappa, 1) 
    maxkappa <- pmax(creat_kappa, 1)
    minkappa_alpha <- minkappa^alpha
    maxkappa_exp <- maxkappa^(-1.200)
    age_term <- 0.9938^age
    sex_term <- ((sex == "Female") * 1.012) + ((sex == "Male") * 1) 
    142 * minkappa_alpha * maxkappa_exp * age_term * sex_term
}
```

```{r}
df <- df %>%
    mutate(
        SurvRegDat = pmap(
            list(SurvDat, Outcome, sex),
            function(DAT, OUT, SEX){
                alrprobdat <- DAT %>%
                    transmute(
                        across(starts_with("prob"), \(x) ifelse(x < 1e-15, 1e-15, x)),
                        across(starts_with("prob"), \(x) ifelse(x > (1 - 1e-15), (1 - 1e-15), x)),
                        across(starts_with("prob"), \(x) log(x/probBC)),
                        probBC = NULL
                    )
                names(alrprobdat) <- paste0("alr", names(alrprobdat))
                switch(
                    OUT,
                    mace = transmute(
                        DAT,
                        outcome_value, outcome_timeyrs,
                        age, smoking, bmi,
                        sbp,
                        tchol = hdl + ldl + (tg/2.2),
                        hdl,
                        fg,
                        whr, alt, crp,
                        T2D, T2Dage,
                        lnegfr = log(egfrcalc(scr, SEX, age)),
                        lnegfrsq = lnegfr^2,
                        age_smoking = age*smoking,
                        age_sbp = age*sbp,
                        age_tchol = age*tchol,
                        age_hdl = age*hdl,
                        age_fg = age*fg,
                        age_lnegfr = age*lnegfr,
                        age_t2d = age*T2D,
                        Insulin, AntiHT, LipidLower
                    ),
                    dm = transmute(
                        DAT,
                        outcome_value, outcome_timeyrs,
                        age, smoking, bmi,
                        whr, sbp, dbp, alt, scr, crp, hdl, tg, ldl, fg,
                        CHD, Stroke, AntiHT, LipidLower
                    )
                ) %>%
                     bind_cols(alrprobdat)   
            }
        )
    )
print(df)
```

## Survival models

```{r}
df <- df %>%
    mutate(
        NullMod = map(
            SurvRegDat, 
            ~coxph(Surv(time = outcome_timeyrs, event = outcome_value) ~ 1, data = .x)
        ),
        ModBase = map(
            SurvRegDat,
            ~{
                .x <- select(.x, -starts_with("alr"))
                coxph(Surv(time = outcome_timeyrs, event = outcome_value) ~ ., data = .x)
            }
        ),
        ModClus = map(
            SurvRegDat, 
            ~coxph(Surv(time = outcome_timeyrs, event = outcome_value) ~ ., data = .x)
        )
    )
print(df)
```

## Added value

```{r}
addedvaltab <- df %>%
    transmute(
        Outcome, sex, Ethnicity,
        LL0 = map_dbl(NullMod, logLik),
        LLBase = map_dbl(ModBase, logLik),
        NVBase = map_dbl(ModBase, \(MOD) sum(!is.na(MOD$coefficients))),
        LLClus = map_dbl(ModClus, logLik),
        NVClus = map_dbl(ModClus, \(MOD) sum(!is.na(MOD$coefficients))),
        LRTstat = 2 * abs(LLClus - LLBase),
        LRTdf = NVClus - NVBase,
        LRTp = stats::pchisq(
            q = LRTstat,
            df = LRTdf,
            lower.tail = FALSE
        ),
        LRTpadj = p.adjust(LRTp, "fdr"),
        AdeqInd = (LLBase - LL0) / (LLClus - LL0),
        CBase = purrr::map_dbl(ModBase, \(MOD) MOD[["concordance"]][["concordance"]]),
        CseBase = purrr::map_dbl(ModBase, \(MOD) MOD[["concordance"]][["std"]]),
        CClus = purrr::map_dbl(ModClus, \(MOD) MOD[["concordance"]][["concordance"]]),
        CseClus = purrr::map_dbl(ModClus, \(MOD) MOD[["concordance"]][["std"]]),
        Cdat = purrr::map2(ModClus, ModBase, survival::concordance),
        Cdiff = purrr::map_dbl(Cdat, \(MOD) c(1, -1) %*% coef(MOD)),
        Cdiffse = purrr::map_dbl(Cdat, \(MOD) sqrt(c(1, -1) %*% vcov(MOD) %*% c(1, -1))),
        Cdiffp = 2*pnorm(-abs(Cdiff / Cdiffse)),
        Cdiffpadj = p.adjust(Cdiffp, "fdr"),
        Cdat = NULL
    )
addedvaltab
```

```{r}
write_tsv(addedvaltab, "../data/addedvaltab_ethn.tsv")
```

```{r}
addedvaltab %>%
    select(Outcome, sex, Ethnicity, LRTp, Cdiffp) %>%
    filter(LRTp < 0.05)
```

#### Effect of increasing 10% in a discordant cluster at the expense of the concordant cluster on outcomes

```{r}
shift10pc <- validclusmod %>%
    expand_grid(Outcome = c("mace", "dm")) %>%
    transmute(
        sex, Outcome,
        NewD = map2(
            clusmod, residmod,
            function(CLUSMOD, RESIDMOD){
                averesidshift <- CLUSMOD %>%
                    expand_grid(
                        Cluster = CLUSMOD$validclus_name
                    ) %>%
                    mutate(
                        # Producing the shifts
                        toadd = case_when(
                            Cluster == "BC" ~ 0,
                            Cluster == validclus_name ~ .1,
                            validclus_name == "BC" ~ -.1,
                            TRUE ~ 0
                        ),
                        validclus_weights = validclus_weights + toadd
                    ) %>%
                    group_by(Cluster) %>%
                    summarise(
                        # Generating data with the shift
                        RGD = list(
                            MGMM::rGMM(
                                n = 5e5, 
                                d = length(validclus_centers[[1]]),
                                k = n(),
                                pi = validclus_weights,
                                means = validclus_centers, 
                                covs = validclus_covmats
                            ) %>%
                                colMeans %>%
                                setNames(
                                    names(validclus_centers[[1]])
                                )
                        )
                    ) %>%
                    mutate(
                        # Values corresponding to the shift
                        predbmvals = map(
                            RGD,
                            ~tibble(
                                Biomarker = names(.x), sdresidval = .x
                            ) %>%
                                inner_join(RESIDMOD, by = "Biomarker") %>%
                                mutate(
                                    sdresid = sdresidval * SDRes,
                                    bmi = 30 * bmi,
                                    age = 60 * age,
                                    bmvalue = rowSums(pick(Intercept, bmi, age, sdresid))
                                ) %>%
                                {setNames(.$bmvalue, .$Biomarker)} %>%
                                c(bmi = 30, age = 60) %>% t
                        ),
                        # Calculating probabilities and log-ratios
                        predprobs = map(
                            RGD, 
                            ~{
                                pdfs <- mapply(
                                    function(mu, covmat){ 
                                        mvtnorm::dmvnorm(
                                            .x, mu, covmat
                                        ) 
                                    },
                                    CLUSMOD$validclus_centers, 
                                    CLUSMOD$validclus_covmats
                                )
                                L <- pdfs %*% diag(CLUSMOD$validclus_weights)
                                colnames(L) <- CLUSMOD$validclus_name
                                L / rowSums(L)
                            }
                        ),
                        alrprobs = map(
                            predprobs,
                            ~{
                                cclus <- colnames(.x) == "BC"
                                alrp <- .x[,!cclus]/.x[,cclus]
                                names(alrp) <- paste0("alrprob", names(alrp))
                                t(alrp)
                            }
                        ),
                        joindat = map2(predbmvals, alrprobs, cbind)
                    ) %>%
                    transmute(Cluster, joindat = map(joindat, data.frame)) %>%
                    unnest(joindat)
            }
        )
    )
print(shift10pc)
```

```{r}
shift10pc$NewD[[1]]
```

```{r}
shift10pc <- shift10pc %>%
    mutate(
        NewD = pmap(
            list(Outcome, NewD, sex),
            function(OUT, DAT, SEX){
                switch(
                    OUT,
                    mace = transmute(
                        DAT,
                        Cluster,
                        sbp,
                        tchol = hdl + ldl + (tg/2.2),
                        hdl,
                        fg,
                        whr, alt, crp,
                        lnegfr = log(egfrcalc(scr, SEX, age)),
                        lnegfrsq = lnegfr^2,
                        age_sbp = age*sbp,
                        age_tchol = age*tchol,
                        age_hdl = age*hdl,
                        age_fg = age*fg,
                        age_lnegfr = age*lnegfr,
                        across(starts_with("alrprob"))
                    ),
                    dm = transmute(
                        DAT,
                        Cluster,
                        whr, sbp, dbp, alt, scr, crp, hdl, tg, ldl, fg,
                        across(starts_with("alrprob"))
                    )
                )
            }
        )
    )
print(shift10pc)
```

```{r}
shift10pc <- shift10pc %>%
    mutate(
        NewD = map(
            NewD,
            ~{
                mat <- as.matrix(select(.x, -Cluster))
                rownames(mat) <- .x$Cluster
                cclus <- rownames(mat) == "BC"
                t(t(mat[!cclus,]) - mat[cclus,])
            }
        )
    )
print(shift10pc)
```

```{r}
shift10pc$NewD[[1]]
```

```{r}
df <- df %>%
    inner_join(
        shift10pc, by = join_by(sex, Outcome)
    )
print(df)
```

And now we can calculate the effect of each shift:

```{r}
shift10est <- df %>%
    transmute(
        sex, Ethnicity, Outcome,
        ModClus = map2(
            ModClus, NewD,
            function(MOD, D){
                BETAS <- coef(MOD)
                nm <- intersect(colnames(D), names(BETAS))
                covm <- vcov(MOD)[nm,nm]
                data.frame(
                    Cluster = gsub("alrprob", "", rownames(D)),
                    estimate = drop(D[,nm] %*% BETAS[nm]),
                    se = apply(
                        D[,nm], 1,
                        \(clus){ 
                            sqrt(drop(clus %*% covm %*% clus))
                        }
                    )
                )
            }
        )
    ) %>%
    unnest(ModClus) %>%
    mutate(
        pval = 2 * pnorm(-abs(estimate/se)),
        HR = exp(estimate),
        lwr = exp(estimate - qnorm(.975)*se),
        upr = exp(estimate + qnorm(.975)*se),
        padj = p.adjust(pval, "fdr")
    )
head(shift10est)
```

```{r}
write_tsv(shift10est, "../data/shift10est_ethn.tsv")
```

```{r}
shift10est %>%
    filter(padj < 0.05)
```

```{r}
options(repr.plot.width = 4, repr.plot.height = 5)
shift10est %>%
    ggplot(aes(HR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(aes(xmin = lwr, xmax = upr)) +
    geom_point() +
    ggh4x::facet_nested(sex + Ethnicity ~ Outcome, scales = "free", space = "free_y") +
    labs(y = NULL) +
    theme_bw()
```

### 2025-11-12 scripts/01_cross_sectional_FX.R

@@ -1,0 +1,470 @@
+## Functions to run clustering analysis to detect subgroups with different relationships between BMI and risk
+
+# Declaring variables that are expected to be found in the data to run analysis
+IDCOL <- "eid"
+SEXVAR <- "sex"
+SEXGROUPS <- c("Female", "Male")
+BODYSIZEINDEX <- "bmi"
+COVARIATES <- c("age", "smoking")
+BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
+PERSEXVARS <- c(BODYSIZEINDEX, COVARIATES, BIOMARKERS)
+ALLVARS <- c(IDCOL, SEXVAR, PERSEXVARS)
+
+
+# Outlier removal
+remove_outliers <- function(x, sdunits = 5){
+    m <- mean(x, na.rm = TRUE)
+    s <- sd(x, na.rm = TRUE)
+    upperbound <- m + (sdunits*s)
+    lowerbound <- m - (sdunits*s)
+    ifelse((x > lowerbound) & (x < upperbound), x, NaN)
+}
+
+# Checking data input
+checkinput <- function(X){
+    stopifnot(is.data.frame(X))
+    stopifnot(ALLVARS %in% names(X))
+    stopifnot(names(X) %in% ALLVARS)
+}
+
+# Variable types
+vartype <- function(x){
+    if(is.numeric(x)){
+        if(length(unique(x)) > 10){
+            return("Numeric")
+        } else { 
+            return("Categorical")
+        }
+    } else {
+        if(is.character(x) | is.factor(x)){
+            return("Categorical")
+        } else {
+            return(NA)
+        }
+    }
+}
+
+# Summary of categorical variables
+categorical_sumstats <- function(x){
+    N <- length(x)
+    N_miss <- sum(is.na(x)) 
+    props <- table(x)
+    props <- sort(props, decreasing = TRUE)
+    props <- data.frame(N = N, N_miss = N_miss,
+                        Summary1 = names(props), 
+                        Summary2 = as.vector(props))
+    props$Summary2 <- paste0(props$Summary2, " (", round(100 * props$Summary2 / sum(props$Summary2), 2), "%)")
+    props
+}
+
+# Summary of continuous variables
+continuous_sumstats <- function(x){
+    N <- length(x)
+    N_miss <- sum(is.na(x))
+    Mn <- round(mean(x, na.rm = TRUE), 2)
[…]

### 2025-11-12 scripts/06_cross_sectional_FX.R

@@ -1,0 +1,1035 @@
+## Functions to assess predictive ability of clusters
+
+## Columns needed
+BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
+BODYSIZEINDEX <- "bmi"
+COVARIATES <- c("age", "smoking")
+DISEASES <- c("HT", "CHD", "Stroke", "PAD", "CKD", "LiverFailure", "RA", "T2D", "T1D")
+MEDICATION <- c("Insulin", "AntiDM", "AntiHT", "LipidLower")
+
+# Calculating cluster probabilities
+clusterprobcalc <- function(ClusModDf, StratDat){
+    alldat <- dplyr::mutate(ClusModDf, data = purrr::map(sex, ~StratDat[[.x]]))
+    dplyr::transmute(
+        alldat,
+        sex,
+        # Calculate standardized residuals
+        resids = purrr::map2(
+            data, residmod,
+            function(DATA, RESIDMOD){
+                sapply(
+                    BIOMARKERS,
+                    function(biomarker){
+                        PREDICTDAT <- as.matrix(cbind(Intercept = 1, DATA[,c(BODYSIZEINDEX, COVARIATES)]))
+                        MODELDAT <- dplyr::filter(RESIDMOD, Biomarker == biomarker)
+                        RESIDSD <- MODELDAT$SDRes
+                        COEFDAT <- MODELDAT[,c("Intercept", BODYSIZEINDEX, COVARIATES)]
+                        COEFDAT <- t(COEFDAT)
+                        PREDVAL <- as.vector(PREDICTDAT %*% COEFDAT)
+                        RESIDVAL <- DATA[[biomarker]] - PREDVAL
+                        RESIDVAL / RESIDSD
+                    }
+                )
+            }
+        ),
+        # Calculate probabilities to valid clusters
+        probs = purrr::map2(
+            resids, clusmod,
+            function(RESIDDF, CLUSMOD){
+                mus <- CLUSMOD$validclus_centers
+                mus <- purrr::map(mus, function(mu){ mu[BIOMARKERS] })
+                covmats <- CLUSMOD$validclus_covmats
+                covmats <- purrr::map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
+                wts <- CLUSMOD$validclus_weights
[…]
